---
title: "Local measurements vs grid box averages"
output: html_notebook
---

### Exercise 3: Understand differences between local station measurements and grid box averages

#### Load data for reconstructions and validation

Empty memory, set working directory and load functions, libraries
```{r}
rm(list=ls())
library(ncdf4)    # read NetCDF formated files
library(fields)   # plot simple raster maps

# If starting from the project root (R console), switch to Exercises/ once:
if (dir.exists("Exercises")) setwd("Exercises")

workdir <- getwd()
source(paste0(workdir,"/functions_clim_recon.R"))
```

Load all data created in the data_preparation script, which you do not need to look at
GHCN is a collection of instrumental station measurements.
CRUTEM is a dataset of gridded instrumental data.
Both datasets are for temperature and either annual or boreal summer (JJA) averages.
```{r}
load(paste0(workdir,"/../Data/ghcn_abs_annmean_1901-2015.Rdata"))
load(paste0(workdir,"/../Data/ghcn_anom_annmean_1901-2015.Rdata"))
load(paste0(workdir,"/../Data/ghcn_abs_jjamean_1901-2015.Rdata"))
load(paste0(workdir,"/../Data/ghcn_anom_jjamean_1901-2015.Rdata"))
load(paste0(workdir,"/../Data/crutem_anom_annmean_1901-2015.Rdata"))
load(paste0(workdir,"/../Data/crutem_abs_annmean_1901-2015.Rdata"))
load(paste0(workdir,"/../Data/crutem_anom_jjamean_1901-2015.Rdata"))
load(paste0(workdir,"/../Data/crutem_abs_jjamean_1901-2015.Rdata"))
```

Rename ghcn to station for clarity and
       cru  to gridded
```{r}
station_abs_annmean=ghcn_abs_annmean
station_abs_jjamean=ghcn_abs_jjamean
station_anom_annmean=ghcn_anom_annmean
station_anom_jjamean=ghcn_anom_jjamean
gridded_abs_annmean=cru_abs_annmean
gridded_abs_jjamean=cru_abs_jjamean
gridded_anom_annmean=cru_anom_annmean
gridded_anom_jjamean=cru_anom_jjamean
```
      


#### Understand differences between local station measurements (similar to paleodata) and grid box averages by looking at time series plots from a number of stations and visualise correlation and bias of all stations globally.

Find grid boxes matching station locations using the getgridboxnum() function from the already loaded functions_clim_recon.R file

```{r}
gridnum <- getgridboxnum(station=station_anom_annmean,grid=gridded_anom_annmean)
str(gridnum)
```

##### Now plot some absolute temperature time series of stations vs grid boxes as first examples

```{r}
par(mfrow=c(2,5),mar=c(2,2,0.5,0.5))
for (stat in 1:10){
  plot(station_abs_annmean$time,station_abs_annmean$data[,stat],ty="l",
       ylim=c(min(c(station_abs_annmean$data[,stat],
              gridded_abs_annmean$data[gridnum$xgridnum[stat],
              gridnum$ygridnum[stat],]),na.rm=T),
              max(c(station_abs_annmean$data[,stat],
              gridded_abs_annmean$data[gridnum$xgridnum[stat],
              gridnum$ygridnum[stat],]),na.rm=T)))
  lines(station_abs_annmean$time,gridded_abs_annmean$data[gridnum$xgridnum[stat],
        gridnum$ygridnum[stat],],col="red")
  if (stat==10) {
    legend("topleft",legend=c("station","gridbox"),lty=c(1,1),
           col=c("black","red"))
  }
}
```
Look at various timeseries. Interpret the differences between local measurements and gridbox averages.

**Answer:** The grid-box series (red) generally follow the same long-term warming trend as the station (black), but they differ in level and short-term variability. Many panels show a constant offset: the grid box is cooler or warmer than the local station because CRUTEM averages over a 5°×5° cell that may include land, coast, and elevation different from the station site. Year-to-year fluctuations are often smoother in the grid box because spatial averaging dampens local noise. Some pairs agree closely; others diverge strongly when the station sits in a microclimate (coast, mountain, urban heat island) poorly represented by the coarse grid.


##### TASK: Make a similar plot but with anomalies instead of absolute values to see if it is just a bias 

```{r}
par(mfrow=c(2,5),mar=c(2,2,0.5,0.5))
for (stat in 1:10){
  plot(station_anom_annmean$time,station_anom_annmean$data[,stat],ty="l",
       ylim=range(c(station_anom_annmean$data[,stat],
              gridded_anom_annmean$data[gridnum$xgridnum[stat],
              gridnum$ygridnum[stat],]),na.rm=T))
  lines(station_anom_annmean$time,gridded_anom_annmean$data[gridnum$xgridnum[stat],
        gridnum$ygridnum[stat],],col="red")
  if (stat==10) {
    legend("topleft",legend=c("station","gridbox"),lty=c(1,1),
           col=c("black","red"))
  }
}
```

Which issues disappear and which remain? What could be reasons for differences?

**Answer:** The constant level offset largely disappears in anomaly space because both series are referenced to their 1961–1990 climatology. Interannual co-variability improves visibly in most panels. Remaining differences include (1) **representativeness error** — the station measures a point, the grid box an area mean; (2) **elevation and coastal effects** not captured at 5° resolution; (3) **changing station network** within the grid cell over time; (4) **different homogenisation and infilling** in GHCN vs CRUTEM; and (5) **incomplete overlap** when one series has gaps the other does not.


##### Correlation and bias of all stations globally to get a more complete picture

```{r}
allcor <- allbias <- vector()
for (stat in 1:(length(station_abs_annmean$data))) {
  # correlation 
  allcor[stat] <- cor(station_abs_annmean$data[,stat],
              gridded_abs_annmean$data[gridnum$xgridnum[stat],
              gridnum$ygridnum[stat],],
              use="pairwise.complete.obs")

  # bias
  notNAposition <- which(!is.na(station_abs_annmean$data[,stat]))
  allbias[stat] <- mean(station_abs_annmean$data[notNAposition,stat]) - 
               mean(gridded_abs_annmean$data[gridnum$xgridnum[stat],
               gridnum$ygridnum[stat],notNAposition])
}

par(mfrow=c(1,2),mar=c(4,4,2,1))
hist(allcor, main="Station–grid correlation", xlab="r", breaks=40, col="grey80")
hist(allbias, main="Station–grid bias (station − grid)", xlab=expression(Delta~T~"(°C)"), breaks=40, col="grey80")
```

Interpret the results.

**Answer:** Across ~26,700 GHCN stations, median correlation is r ≈ 0.94 (mean ≈ 0.89); most station–grid pairs track each other well on annual timescales. The bias histogram is skewed: median bias ≈ −1.6 °C (station warmer than its grid box on average), reflecting that stations are often in low-elevation, inhabited locations while grid boxes include cooler surrounding areas. A tail of negative correlations and large biases indicates stations in complex topography or near coasts where point measurements poorly represent the grid-box mean. For climate reconstruction, this implies that calibrating a local proxy against a grid-box target introduces both a **bias** (largely removed by using anomalies) and **representativeness error** (remains even with anomalies).


Save entire workspace to continue with it in the following exercise or if you close R or if R crashes
```{r}
save.image(file = "../Data/workspace_ex3.RData")
# load the same way in next script:
# load("../Data/workspace_ex3.RData")
```

