---
title: "European station average"
output: html_notebook
---
  
### Exercise 4: European mean based on simple average
  
Empty memory, set working directory and load functions, libraries and data from previous exercises

```{r}
rm(list=ls())
if (dir.exists("Exercises")) setwd("Exercises")
load(file = "../Data/workspace_ex3.RData")
```


#### European (lon: -10 to +40; lat: 30 to 80) station average using all stations and absolute annual mean temperature vs the mean of the gridded dataset

```{r}
posEUstat <- which(station_abs_annmean$lon>-10 & station_abs_annmean$lon<40 &station_abs_annmean$lat>30 & station_abs_annmean$lat<70)
EUstatmean <- apply(station_abs_annmean$data[,posEUstat],1,mean,na.rm=T)
plot(station_abs_annmean$time, EUstatmean,ty="l",ylim=c(3,10))
# add time series from gridded data as reference
posEUgridLON <- which(gridded_anom_annmean$lon>-10 & gridded_anom_annmean$lon<40)  
posEUgridLAT <- which(gridded_anom_annmean$lat>30 & gridded_anom_annmean$lat<70)
EUgridmean_abs <- apply(gridded_abs_annmean$data[posEUgridLON,posEUgridLAT,],
                        3,mean,na.rm=T)
lines(station_abs_annmean$time, EUgridmean_abs, col="red")
cor(EUstatmean, EUgridmean_abs, use="pairwise.complete.obs")
```
Interpret the quality of the simple station average as a reconstruction of the spatial mean under the assumption that the gridded reference dataset mean is the "truth".  

**Answer:** The simple mean of all ~4,000 European stations tracks the gridded regional mean in broad terms (common warming trend), but the correlation is only r ≈ 0.55. Early in the record the station mean is biased low because the network was sparse and concentrated in certain regions; as stations are added over time, the mean shifts upward — a classic **network bias**. The absolute level also differs from the grid mean because stations are not area-representative. As a spatial-mean reconstruction, the all-station simple average is therefore a poor proxy for the "true" regional mean, especially before ~1950.


##### Now only use at least 100-year long series for constant network by finding series with a maximum of 5% missing data in the 1901-2015 period

```{r}
station_EU <- station_abs_annmean$data[,posEUstat]
station_longEU <- rep(NA,nrow(station_EU))
station_longEUlon <- station_longEUlat <- NA
for (i in 1:ncol(station_EU)) {
  if (length(which(!is.na(station_EU[,i])))>nrow(station_EU)*0.95) {
    station_longEU <- cbind(station_longEU,station_EU[,i])
    station_longEUlon <- cbind(station_longEUlon,station_abs_annmean$lon[posEUstat][i])
    station_longEUlat <- cbind(station_longEUlat,station_abs_annmean$lat[posEUstat][i])
  }
}
station_longEU <- station_longEU[,-1]
station_longEUlon <- station_longEUlon[-1]
station_longEUlat <- station_longEUlat[-1]
dim(station_longEU)
length(station_longEUlon)
length(station_longEUlat)
EUlongstatmean <- apply(station_longEU,1,mean,na.rm=T)
plot(station_abs_annmean$time,EUlongstatmean,ty="l",ylim=c(0,10))

# same for grid data
gridded_longEU <- array(NA,dim(gridded_abs_annmean$data)) 
for (i in 1:length(posEUgridLON)) {
  for (j in 1:length(posEUgridLAT)) {
    if (length(which(!is.na(gridded_abs_annmean$data[posEUgridLON[i],posEUgridLAT[j],])))
        > length(gridded_abs_annmean$data[i,j,])*0.95) {
      gridded_longEU[posEUgridLON[i],posEUgridLAT[j],] <- 
        gridded_abs_annmean$data[posEUgridLON[i],posEUgridLAT[j],]
    }
  }
}

EUlonggridmean <- apply(gridded_longEU,3,mean,na.rm=T)

lines(station_abs_annmean$time, EUlonggridmean, col="red")
cor(EUlongstatmean, EUlonggridmean, use="pairwise.complete.obs")
```
How did this selection improve the result and which issues remain?

**Answer:** Restricting to 129 long, complete series dramatically improves agreement: r ≈ 0.95 vs 0.55 before. Removing the changing network stabilises the regional mean and eliminates most of the artificial drift in early decades. Remaining issues: (1) the long-station subset is still not area-representative (more stations in western/central Europe than in the Mediterranean or eastern Europe); (2) a small level offset persists in absolute temperatures; (3) simple (unweighted) averaging still over-represents data-dense regions.


##### Long series lat weighted

So far, we took a simple average not taking into account that lon/lat grid boxes decrease in size the higher the latitude.

```{r}
latmean <- array(NA,dim=c(length(posEUgridLAT),dim(gridded_abs_annmean$data)[3]))

weights <- cos(gridded_abs_annmean$lat[posEUgridLAT]/180*pi)
weights <- weights/sum(weights)
for (i in 1:length(posEUgridLAT)) {
  latmean[i,] <- apply(gridded_longEU[,posEUgridLAT[i],],2,mean,na.rm=T) * weights[i]
}
EUlonggridmean_latwgt <- apply(latmean,2,sum,na.rm=T)
plot(station_abs_annmean$time,EUlongstatmean,ty="l",ylim=c(0,10))
lines(station_abs_annmean$time, EUlonggridmean,col="red")
lines(station_abs_annmean$time, EUlonggridmean_latwgt,col="blue")
```

How does the latitude/area weighing influence the results?

**Answer:** Latitude weighting (cos φ) has only a modest effect on the gridded reference series: the blue (area-weighted) and red (simple grid mean) lines are very similar, and both correlate with the station mean at r ≈ 0.94–0.95. Weighting slightly reduces the influence of high-latitude grid boxes (which cover less area) and can shift the mean by a few tenths of a degree in some decades. For the station side, however, no comparable area weighting is applied — stations remain unweighted, which is a larger source of spatial bias than the grid-side weighting choice.


##### TASK: Go back to the code for the simple mean of all stations but use temperature anomalies wrt 1961-90 instead of absolute values

```{r}
EUstatmeananom <- apply(station_anom_annmean$data[,posEUstat],1,mean,na.rm=T)
EUgridmean_anom <- apply(gridded_anom_annmean$data[posEUgridLON,posEUgridLAT,],3,mean,na.rm=T)
plot(station_anom_annmean$time, EUstatmeananom, ty="l",
     ylim=range(c(EUstatmeananom, EUgridmean_anom), na.rm=TRUE))
lines(station_anom_annmean$time, EUgridmean_anom, col="red")
cor(EUstatmeananom, EUgridmean_anom, use="pairwise.complete.obs")
```

How did using anomalies instead of absolute values influence the results?

**Answer:** Using anomalies improves the all-station comparison substantially: r rises from ≈ 0.55 (absolute) to ≈ 0.98. Removing the climatological offset eliminates much of the level mismatch between station and grid means. Interannual and decadal variability now align closely, confirming that the main problem with the absolute simple mean was bias in mean level and network composition rather than inability to capture temperature variability. A small divergence remains in the earliest decades when the station network was thinnest, but overall anomalies are far better suited for comparing regional means and for climate-reconstruction validation.


Save entire workspace to continue with it in the following exercise or if you close R or if R crashes
```{r}
save.image(file = "../Data/workspace_ex4.RData")
```

