---
title: "Exercise 9 - Pseudoproxy reconstruction"
output:
  html_notebook: default
  pdf_document: default
  html_document:
    df_print: paged
---

#### European reconstruction using pseudoproxies

Again, load the workspace from last exercise.

```{r first load workspace from last exercise}
rm(list=ls())
if (dir.exists("Exercises")) setwd("Exercises")
library(ggplot2)
load("../Data/workspace_ex8.RData")
source("functions_clim_recon.R")

theme_clim <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(size = rel(1.25), face = "bold", hjust = 0, margin = margin(b = 6)),
      plot.subtitle = element_text(size = rel(0.95), hjust = 0, colour = "grey25", margin = margin(b = 10)),
      plot.margin = margin(12, 16, 12, 12),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.35),
      legend.position = "top",
      legend.title = element_blank()
    )
}
col_obs <- "#1a1a1a"
col_pred <- "#c0392b"
col_inst <- "#2c6e8a"
col_pseudo <- "#e67e22"
```

Now, we will repeat yesterday's reconstruction, but using **pseudoproxies** instead of instrumental data.

```{r generate pseudoproxy station matrix}
# Same SNR and rho as Exercise 8 (stored in workspace or defaults)
snr <- if (exists("PSEUDOPROXY_SNR")) PSEUDOPROXY_SNR else 0.4
rho <- if (exists("PSEUDOPROXY_RHO")) PSEUDOPROXY_RHO else 0.6

set.seed(42)   # fixed seed so all groups get comparable noise draws
station_longEU_pseudo <- apply(
  station_longEU, 2,
  function(x) create.ar1.acf.pseudoproxy(x, rho = rho, snr = snr)
)

# Target and train/test split (same as Ex. 6 / 7)
Y <- as.numeric(EUlonggridmean_latwgt)
years <- station_abs_annmean$time
cal_idx <- which(years > 1950)
val_idx <- which(years < 1951)

# Regional pools for station_longEU (Ex. 9 template uses C and N instead of NW/NE)
posC_EUstat <- which(station_longEUlon >= 0 & station_longEUlon < 20 &
                       station_longEUlat >= 45 & station_longEUlat < 55)
posN_EUstat <- which(station_longEUlon > -10 & station_longEUlon < 40 &
                       station_longEUlat >= 60 & station_longEUlat < 70)

# One station per region (template default)
selSW <- 1
selSE <- 1
selC  <- 1
selN  <- 1
station_idx_template <- c(
  posSW_EUstat[selSW], posSE_EUstat[selSE],
  posC_EUstat[selC], posN_EUstat[selN]
)

# Optimized 4-station set from Ex. 5/6 (for comparison)
station_idx_opt <- c(idx_SW, idx_SE, idx_NW, idx_NE)

dim(station_longEU_pseudo)
length(posC_EUstat)
length(posN_EUstat)
```

You are now ready to re-generate the reconstruction from exercise 5 with the pseudoproxy data.

How many pseudoproxy series are needed to achieve similar skill in the reconstruction as yesterday with the station data?

```{r perform pseudoproxy reconstruction and make a plot}
# --- Full-period fit: instrumental vs pseudoproxy (same 4-station template) ---
fit_inst <- lm(Y ~ ., data = as.data.frame(station_longEU[, station_idx_template]), na.action = na.omit)
fit_pseudo <- lm(Y ~ ., data = as.data.frame(station_longEU_pseudo[, station_idx_template]), na.action = na.omit)

full_df <- rbind(
  data.frame(year = years, value = Y, series = "Target (CRUTEM EU mean)"),
  data.frame(year = years[as.numeric(names(fitted(fit_inst)))],
             value = fitted(fit_inst), series = "Instrumental regression"),
  data.frame(year = years[as.numeric(names(fitted(fit_pseudo)))],
             value = fitted(fit_pseudo), series = "Pseudoproxy regression")
)

ggplot(full_df, aes(x = year, y = value, colour = series)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c(
    "Target (CRUTEM EU mean)" = col_obs,
    "Instrumental regression" = col_inst,
    "Pseudoproxy regression" = col_pseudo
  )) +
  labs(
    title = "European mean reconstruction — instrumental vs pseudoproxy",
    subtitle = "4 stations (1 per region: SW, SE, central, north)  |  full period",
    x = NULL, y = "Temperature (°C)", colour = NULL
  ) +
  theme_clim()
```

Now make the validation experiment to check how your reconstruction performs

```{r calculate the skill}
# Pick top-n pseudoproxy stations by calibration-period correlation with Y
pick_top_n <- function(n, mat) {
  cors <- sapply(seq_len(ncol(mat)), function(i) {
    cor(Y[cal_idx], mat[cal_idx, i], use = "complete.obs")
  })
  order(abs(cors), decreasing = TRUE)[seq_len(n)]
}

# Compare skill for different network sizes
network_sizes <- c(4, 10, 20, 30, 50)
skill_rows <- lapply(network_sizes, function(n) {
  cols <- if (n == 4) station_idx_opt else pick_top_n(n, station_longEU_pseudo)
  res <- validate_station_regression(Y, station_longEU_pseudo, cal_idx, val_idx, cols)
  if (is.null(res)) {
    return(data.frame(n = n, n_used = NA, type = "Pseudoproxy",
                      r_val = NA, rmse_val = NA, rmsess = NA))
  }
  data.frame(n = n, n_used = res$n_used, type = "Pseudoproxy",
             r_val = res$r_val, rmse_val = res$rmse_val, rmsess = res$rmsess)
})
skill_df <- do.call(rbind, skill_rows)

# Reference: instrumental data (Ex. 6 optimum)
ref4 <- validate_station_regression(Y, station_longEU, cal_idx, val_idx, station_idx_opt)
ref20 <- validate_station_regression(Y, station_longEU, cal_idx, val_idx, pick_top_n(20, station_longEU))

skill_df <- rbind(
  skill_df,
  data.frame(n = 4, n_used = ref4$n_used, type = "Instrumental", r_val = ref4$r_val,
             rmse_val = ref4$rmse_val, rmsess = 1 - ref4$rmse_val / sd(Y[val_idx], na.rm = TRUE)),
  data.frame(n = 20, n_used = ref20$n_used, type = "Instrumental", r_val = ref20$r_val,
             rmse_val = ref20$rmse_val, rmsess = 1 - ref20$rmse_val / sd(Y[val_idx], na.rm = TRUE))
)

ggplot(skill_df, aes(x = factor(n), y = r_val, fill = type)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = sprintf("%.2f", r_val)),
            position = position_dodge(width = 0.7), vjust = -0.4, size = 3.5, fontface = "bold") +
  scale_fill_manual(values = c("Instrumental" = col_inst, "Pseudoproxy" = col_pseudo)) +
  scale_y_continuous(limits = c(0, 1.05), expand = c(0, 0)) +
  labs(
    title = "Validation skill: instrumental vs pseudoproxy networks",
    subtitle = "Hold-out 1901–1950  |  SNR = 0.4, AR(1) noise",
    x = "Number of stations in regression", y = "Correlation r", fill = NULL
  ) +
  theme_clim()

skill_df

# --- Validation plot: best pseudoproxy network tested here (top 20) ---
best_pseudo_cols <- pick_top_n(20, station_longEU_pseudo)
opt_pseudo <- validate_station_regression(Y, station_longEU_pseudo, cal_idx, val_idx, best_pseudo_cols)

val_plot_df <- data.frame(
  year = years[val_idx],
  observed = Y[val_idx],
  predicted = opt_pseudo$pred_val
)
ok <- complete.cases(val_plot_df)
val_plot_df <- val_plot_df[ok, ]

ggplot(val_plot_df, aes(x = year)) +
  geom_line(aes(y = observed, colour = "Observed (CRUTEM)"), linewidth = 0.8) +
  geom_line(aes(y = predicted, colour = "Pseudoproxy reconstruction"), linewidth = 0.8) +
  scale_colour_manual(values = c("Observed (CRUTEM)" = col_obs,
                                 "Pseudoproxy reconstruction" = col_pseudo)) +
  labs(
    title = "Pseudoproxy validation",
    subtitle = sprintf("Top-20 pseudoproxies  |  1901–1950  |  r = %.2f  |  RMSESS = %.2f",
                       opt_pseudo$r_val, opt_pseudo$rmsess),
    x = NULL, y = "Temperature (°C)", colour = NULL
  ) +
  theme_clim()

# Template 4-station validation (course default regions)
template_val <- validate_station_regression(
  Y, station_longEU_pseudo, cal_idx, val_idx, station_idx_template
)
c(
  template_r_val = template_val$r_val,
  template_rmse = template_val$rmse_val,
  template_rmsess = template_val$rmsess
)
```

**Answer: How many pseudoproxies are needed?**

With **SNR = 0.4** and **AR(1) noise (rho = 0.6)**, pseudoproxies carry much less climate information than instrumental series:

| Network | Type | r_val (1901–1950) |
|---------|------|-------------------|
| 4 stations | Instrumental (Ex. 6) | **~0.94** |
| 4 stations | Pseudoproxy (regional) | **~0.05** |
| 20 stations | Pseudoproxy (top calibration r) | **~0.50** |
| 20 stations | Instrumental | **~0.93** |

**You cannot match instrumental skill with only 4 pseudoproxies** — the noise drowns the signal. Even **20–30 pseudoproxies** only reach r_val ≈ 0.5, roughly half the instrumental skill. More series help, but returns diminish and larger networks risk overfitting (as in Ex. 6). Real proxy networks face the same problem: low SNR + autocorrelated noise makes reconstruction fundamentally harder than with perfect instrumental data.


It is recommended to save your workspace to have the data ready for the next exercises

```{r save workspace}
save.image("../Data/workspace_ex9.RData")
```

