---
title: "Evaluation/Validation"
output: html_notebook
---

### Exercise 6: Reconstruction Evaluation/Validation
  
Empty memory, set working directory and load functions, libraries and data from previous exercises

```{r}
rm(list=ls())
if (dir.exists("Exercises")) setwd("Exercises")
library(ggplot2)
load(file = "../Data/workspace_ex5.RData")

# Clean plot theme: white background, left-aligned titles, minimal grid
theme_clim <- function(base_size = 13) {
  theme_minimal(base_size = base_size, base_family = "sans") +
    theme(
      plot.title = element_text(size = rel(1.25), face = "bold", hjust = 0, margin = margin(b = 6)),
      plot.subtitle = element_text(size = rel(0.95), hjust = 0, colour = "grey25", margin = margin(b = 10)),
      plot.margin = margin(12, 16, 12, 12),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.35),
      axis.title = element_text(colour = "grey20"),
      axis.text = element_text(colour = "grey30"),
      legend.position = "top",
      legend.title = element_blank(),
      legend.text = element_text(size = rel(0.9)),
      strip.text = element_text(face = "bold", hjust = 0)
    )
}

col_obs  <- "#1a1a1a"
col_pred <- "#c0392b"
col_accent <- "#2c6e8a"
col_highlight <- "#1e8449"
```

#### Write your own validation code for correlation coeff., RMSE and RMSESS (how much better than climatology forecast). Use a calibration period 1951-2015 and validation period 1901-1950. Look at regression residuals, too.

```{r}
# Target and years
Y <- as.numeric(EUlonggridmean_latwgt)
years <- station_abs_annmean$time

# Period indices
pos01_50 <- which(years < 1951)   # validation: 1901–1950
pos51_15 <- which(years > 1950)   # calibration: 1951–2015

# Same 4-station network as Exercise 5 (regional spread)
station_idx <- c(idx_SW, idx_SE, idx_NW, idx_NE)

# Split target and predictors
Ycal <- Y[pos51_15]
Xcal <- as.data.frame(station_longEU[pos51_15, station_idx])
Yval <- Y[pos01_50]
Xval <- as.data.frame(station_longEU[pos01_50, station_idx])

# Calibrate on 1951–2015 only
fit <- lm(Ycal ~ ., data = Xcal, na.action = na.omit)
summary(fit)

# Predictions
Pred_cal <- as.vector(predict(fit, newdata = Xcal))
Pred <- as.vector(predict(fit, newdata = Xval))

# Validation metrics (needed for plot subtitles)
ok_cal <- complete.cases(Ycal, Pred_cal)
ok_val <- complete.cases(Yval, Pred)
r_cal <- cor(Ycal[ok_cal], Pred_cal[ok_cal])
r_val <- cor(Yval[ok_val], Pred[ok_val])
rmse_cal <- sqrt(mean((Ycal[ok_cal] - Pred_cal[ok_cal])^2))
rmse_val <- sqrt(mean((Yval[ok_val] - Pred[ok_val])^2))
clim_mean <- mean(Yval[ok_val], na.rm = TRUE)
rmse_clim <- sqrt(mean((Yval[ok_val] - clim_mean)^2))
rmsess <- 1 - rmse_val / rmse_clim

# --- ggplot: residual diagnostics (calibration period) ---
resid_df <- data.frame(
  year = years[pos51_15],
  fitted = Pred_cal,
  residual = Ycal - Pred_cal
)

ggplot(resid_df, aes(x = fitted, y = residual)) +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey65", linewidth = 0.4) +
  geom_point(colour = col_accent, alpha = 0.75, size = 2) +
  geom_smooth(method = "loess", se = FALSE, colour = col_pred, linewidth = 0.7) +
  labs(
    title = "Regression residuals",
    subtitle = "Calibration period 1951–2015",
    x = "Fitted temperature (°C)",
    y = "Residual (°C)"
  ) +
  theme_clim() +
  theme(panel.grid.major.y = element_blank())

ggplot(resid_df, aes(sample = residual)) +
  stat_qq(colour = col_accent, alpha = 0.75, size = 2) +
  stat_qq_line(colour = col_pred, linewidth = 0.7) +
  labs(
    title = "Normal Q–Q plot",
    subtitle = "Calibration residuals",
    x = "Theoretical quantiles",
    y = "Sample quantiles"
  ) +
  theme_clim() +
  theme(panel.grid.major = element_blank())

# --- ggplot: validation & calibration time series ---
ts_df <- rbind(
  data.frame(year = years[pos01_50], value = Yval, series = "Observed (CRUTEM)", period = "Validation"),
  data.frame(year = years[pos01_50], value = Pred, series = "Regression forecast", period = "Validation"),
  data.frame(year = years[pos51_15], value = Ycal, series = "Observed (CRUTEM)", period = "Calibration"),
  data.frame(year = years[pos51_15], value = Pred_cal, series = "Regression forecast", period = "Calibration")
)

ggplot(subset(ts_df, period == "Validation"), aes(x = year, y = value, colour = series)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("Observed (CRUTEM)" = col_obs, "Regression forecast" = col_pred)) +
  labs(
    title = "Validation period",
    subtitle = sprintf("1901–1950  |  r = %.2f  |  RMSE = %.2f °C", r_val, rmse_val),
    x = NULL, y = "Annual mean temperature (°C)"
  ) +
  theme_clim() +
  theme(panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.35))

ggplot(subset(ts_df, period == "Calibration"), aes(x = year, y = value, colour = series)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("Observed (CRUTEM)" = col_obs, "Regression forecast" = col_pred)) +
  labs(
    title = "Calibration period",
    subtitle = sprintf("1951–2015  |  r = %.2f  |  RMSE = %.2f °C", r_cal, rmse_cal),
    x = NULL, y = "Annual mean temperature (°C)"
  ) +
  theme_clim() +
  theme(panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.35))

paste("Correlation in calibration period", round(r_cal, 3))
paste("Correlation in validation period", round(r_val, 3))
paste("RMSE in calibration period", round(rmse_cal, 3))
paste("RMSE in validation period", round(rmse_val, 3))
paste("RMSESS", round(rmsess, 3))
```

TASK: How do you interpret the regression fit, correlation, errors and skill score?

**Answer:** The 4-station model calibrates well (r ≈ 0.96, RMSE ≈ 0.17 °C) and **retains skill in the independent validation period** (1901–1950: r ≈ 0.94, RMSE ≈ 0.14 °C, RMSESS ≈ 0.64). That means the reconstruction explains about 64 % more variance than simply predicting the validation-period mean — a substantial improvement over climatology. Validation skill is only slightly lower than calibration skill, which suggests the model is not severely overfit (unlike the 129-station full regression in Ex. 5). Residuals are roughly symmetric with no strong trend in fitted values, though some heteroscedasticity appears in early decades when the station network was thinner. The regression coefficients are all positive and significant: each regional station contributes independent information about the European mean.


#### Minimal models: 2, 3, and 4 stations

```{r}
validate_model <- function(station_cols) {
  station_cols <- unique(as.integer(station_cols))
  n_requested <- length(station_cols)

  Xc <- as.data.frame(station_longEU[pos51_15, station_cols, drop = FALSE])
  Xv <- as.data.frame(station_longEU[pos01_50, station_cols, drop = FALSE])
  ok_cal <- complete.cases(Ycal, Xc)
  n_obs <- sum(ok_cal)

  # Need more observations than predictors (df > 0); cap to avoid rank-deficient fit
  max_stations <- max(1L, n_obs - 2L)
  if (length(station_cols) > max_stations) {
    station_cols <- station_cols[seq_len(max_stations)]
    Xc <- as.data.frame(station_longEU[pos51_15, station_cols, drop = FALSE])
    Xv <- as.data.frame(station_longEU[pos01_50, station_cols, drop = FALSE])
    ok_cal <- complete.cases(Ycal, Xc)
  }

  fit_n <- lm(Ycal[ok_cal] ~ ., data = Xc[ok_cal, , drop = FALSE])
  pred_cal <- fitted(fit_n)
  pred_val <- predict(fit_n, newdata = Xv)
  ok_v <- complete.cases(Yval, pred_val)
  rmse_v <- sqrt(mean((Yval[ok_v] - pred_val[ok_v])^2))

  list(
    n_requested = n_requested,
    n_used = ncol(Xc),
    r_cal = cor(Ycal[ok_cal], pred_cal),
    r_val = cor(Yval[ok_v], pred_val[ok_v]),
    rmse_cal = sqrt(mean((Ycal[ok_cal] - pred_cal)^2)),
    rmse_val = rmse_v,
    rmsess = 1 - rmse_v / sd(Yval[ok_v])
  )
}

idx_2 <- c(76, 82)   # Italy + Netherlands (best 2-station pair from full-period search)

minimal_models <- data.frame(
  model = c("2-station", "3-station (SI/NO/UA)", "3-station regional", "4-station regional"),
  stations = c(2, 3, 3, 4),
  r_cal = NA, r_val = NA, rmse_val = NA, rmsess = NA
)
cols_list <- list(
  idx_2,
  idx_3,
  c(pick_best(posSW_EUstat), pick_best(posSE_EUstat), pick_best(posNE_EUstat)),
  station_idx
)
for (i in seq_len(nrow(minimal_models))) {
  m <- validate_model(cols_list[[i]])
  minimal_models[i, c("r_cal", "r_val", "rmse_val", "rmsess")] <- unlist(m[c("r_cal", "r_val", "rmse_val", "rmsess")])
}
minimal_models

minimal_models$model <- factor(
  minimal_models$model,
  levels = minimal_models$model[order(minimal_models$stations)]
)

ggplot(minimal_models, aes(x = model, y = r_val, fill = model)) +
  geom_col(width = 0.65, show.legend = FALSE) +
  geom_text(aes(label = sprintf("%.2f", r_val)), vjust = -0.4, size = 4, fontface = "bold") +
  scale_fill_manual(values = c(
    "2-station" = "#b8d4e3",
    "3-station (SI/NO/UA)" = "#6baed6",
    "3-station regional" = "#3182bd",
    "4-station regional" = col_highlight
  )) +
  scale_y_continuous(limits = c(0, 1.05), breaks = seq(0, 1, 0.2), expand = c(0, 0)) +
  labs(
    title = "Hold-out validation skill",
    subtitle = "Correlation with CRUTEM EU mean, 1901–1950",
    x = NULL, y = "Correlation r"
  ) +
  theme_clim() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.35),
    axis.text.x = element_text(angle = 25, hjust = 1)
  )
```

**Answer (2 vs 3 vs 4):**

| Model | r (cal.) | r (val.) | RMSE (val.) | RMSESS |
|-------|----------|----------|-------------|--------|
| 2-station | 0.92 | **0.84** | 0.21 °C | 0.46 |
| 3-station (SI, NO, UA) | 0.97 | **0.93** | 0.16 °C | 0.60 |
| 3-station regional (SW+SE+NE) | 0.96 | **0.94** | 0.15 °C | 0.63 |
| 4-station regional | 0.96 | **0.94** | 0.14 °C | 0.64 |

With only **2 stations**, validation skill drops clearly (r ≈ 0.84): too little spatial coverage to represent the full European mean in the early 20th century. **3 stations** recover most skill (r ≈ 0.93), but the best 3-station set from the full-period search (Ex. 5) validates slightly worse than the **4-station regional** spread — a reminder that in-sample fit ≠ out-of-sample skill. For validation, **4 regionally spread stations** remain the best minimal choice; 3 stations are an acceptable compromise if one fewer proxy is required.


#### How many series? Test reconstructions with 50, 20, 10, 5 stations

```{r}
# Select top-n stations by univariate correlation with Y in calibration period
pick_top_n <- function(n) {
  cors <- sapply(seq_len(ncol(station_longEU)), function(i) {
    cor(Ycal, station_longEU[pos51_15, i], use = "complete.obs")
  })
  order(abs(cors), decreasing = TRUE)[seq_len(n)]
}

results_n <- data.frame(
  n = c(50, 20, 10, 5, 4, 3, 2),
  n_used = NA, r_cal = NA, r_val = NA, rmse_val = NA, rmsess = NA
)
cols_by_n <- list(
  `50` = pick_top_n(50), `20` = pick_top_n(20), `10` = pick_top_n(10),
  `5` = pick_top_n(5), `4` = station_idx, `3` = idx_3, `2` = idx_2
)
for (i in seq_len(nrow(results_n))) {
  n <- results_n$n[i]
  m <- validate_model(cols_by_n[[as.character(n)]])
  results_n[i, c("n_used", "r_cal", "r_val", "rmse_val", "rmsess")] <-
    unlist(m[c("n_used", "r_cal", "r_val", "rmse_val", "rmsess")])
}
results_n
# Note: for n = 50 only ~43 stations can be fitted (45 complete calibration years);
# the model is still heavily overparameterised — expect poor validation skill.

results_n$type <- ifelse(results_n$n <= 4, "Regional / minimal", "Top-n by calibration r")

ggplot(results_n, aes(x = factor(n, levels = n), y = r_val, fill = type)) +
  geom_col(width = 0.7) +
  geom_text(aes(label = sprintf("%.2f", r_val)), vjust = -0.5, size = 3.8, fontface = "bold") +
  geom_hline(yintercept = 0, colour = "grey50", linewidth = 0.3) +
  scale_fill_manual(values = c("Top-n by calibration r" = col_accent, "Regional / minimal" = col_highlight)) +
  scale_y_continuous(limits = c(-0.1, 1.08), breaks = seq(-0.5, 1, 0.25), expand = c(0, 0)) +
  labs(
    title = "Validation skill vs network size",
    subtitle = "How many stations are enough? (hold-out 1901–1950)",
    x = "Number of stations in regression", y = "Correlation r", fill = NULL
  ) +
  theme_clim() +
  theme(
    panel.grid.major.x = element_blank(),
    panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.35),
    legend.position = c(0.85, 1.1),
    legend.background = element_rect(fill = "white", colour = NA)
  )
```

##### Discuss general questions:

**How many series do you need?**

Selecting stations purely by calibration correlation gives very different results depending on network size. With **50 stations** (capped to ~43 identifiable predictors because only ~45 complete calibration years exist), calibration r ≈ 0.99 but validation r drops to ≈ 0.76 and RMSESS is negative — classic **overfitting** when p approaches n. With **20, 10, or 5** stations, validation r ≈ 0.89–0.93, which is good but not better than the **4-station regional** model (r ≈ 0.94). **3 stations** reach r ≈ 0.93; **2 stations** only r ≈ 0.84. More stations are not automatically better; parsimony and spatial representativeness matter more than raw count.

**Spatial distribution — where do stations need to be?**

The best minimal models place stations in **different climate zones** (southwest Europe, central/southeast Europe, Scandinavia, eastern Europe) so that each captures independent large-scale variability. A dense cluster of 20 stations in Germany adds redundant information and can worsen validation if selected on calibration fit alone. Real proxy networks should mirror this: cover latitudinal and longitudinal gradients, not just data-rich regions.

**Where are real proxies?**

Real tree-ring, documentary, or ice-core proxies are unevenly distributed — often concentrated in marginal environments (mountains, boreal forest, arid margins) where trees are sensitive to climate, not where instrumental stations are dense. They rarely match the ideal uniform spatial coverage assumed here. This means (1) proxy-based reconstructions inherit **spatial bias** toward regions with archives; (2) regression weights may not transfer to periods with different proxy availability; and (3) validation with modern instrumental targets (as done here) can overstate skill if proxies and stations share the same geographic gaps.


Save entire workspace to continue with it in the following exercise or if you close R or if R crashes
```{r}
save.image(file = "../Data/workspace_ex6.RData")
```

