---
title: "Proxy screening"
output: html_notebook
---

#### Exercise 12: Proxy data screening

We load the workspace from Exercise 11. It already contains correlation coefficients between each proxy record and the **local CRUTEM grid box** (nearest grid cell to the proxy location) for the overlap period **1901–2012**.

```{r setup}
rm(list = ls())
if (dir.exists("Exercises")) setwd("Exercises")
library(ggplot2)
library(maps)
source("functions_clim_recon.R")
load(file = "../Data/workspace_ex11.RData")

theme_clim <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(face = "bold", hjust = 0),
      plot.subtitle = element_text(colour = "grey30", hjust = 0),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      panel.grid.minor = element_blank()
    )
}
```

## 1. Significance screening

**Idea:** Before using a proxy in a reconstruction, we check whether it covaries with local instrumental temperature strongly enough that the relationship is unlikely to be random (p-value from `cor.test()`).

```{r screening counts}
# NTREND: tree-ring network pre-selected for temperature sensitivity
n_ntr_05 <- sum(ntrallcor_jja_pval < 0.05, na.rm = TRUE)
n_ntr_01 <- sum(ntrallcor_jja_pval < 0.01, na.rm = TRUE)

# PAGES2k: broader multi-archive collection (also temperature-oriented, but less strict)
n_p2k_jja_05 <- sum(p2kallcor_jja_pval < 0.05, na.rm = TRUE)
n_p2k_ann_05 <- sum(p2kallcor_ann_pval < 0.05, na.rm = TRUE)

cat("NTREND (JJA), p < 0.05:", n_ntr_05, "of", length(ntrallcor_jja_pval), "\n")
cat("NTREND (JJA), p < 0.01:", n_ntr_01, "\n")
cat("PAGES2k (JJA), p < 0.05:", n_p2k_jja_05, "of", length(p2kallcor_jja_pval), "\n")
cat("PAGES2k (annual), p < 0.05:", n_p2k_ann_05, "\n")
```

**Discussion (NTREND vs PAGES2k, p < 0.05):**

- Only **`r n_ntr_05`** of **`r length(ntrallcor_jja_pval)`** NTREND series pass the screen — surprisingly few for a collection marketed as temperature-sensitive. Reasons: (1) we compare **absolute** local grid-box temperature, while many tree rings respond to **anomalies** or moisture; (2) the local grid box may not represent micro-climate at the tree site; (3) short or noisy overlap reduces statistical power.
- **`r n_p2k_jja_05`** PAGES2k records pass — more, but still far from 100%, because the archive mix includes non-temperature signals, dating uncertainty, and spatial mismatch.
- Simple correlation screening is a **first filter only** (see PAGES2k screening papers). It ignores autocorrelation, multiple testing, and non-linear responses.

## 2. Influence of seasonality (annual vs JJA)

Tree rings in the Northern Hemisphere often integrate **summer (JJA)** conditions. Comparing proxies to the wrong season weakens correlations artificially.

```{r season comparison}
screen_df <- data.frame(
  dataset = rep(c("PAGES2k", "NTREND"), each = 2),
  season  = rep(c("Annual", "JJA"), times = 2),
  n_sig   = c(
    sum(p2kallcor_ann_pval < 0.05, na.rm = TRUE),
    sum(p2kallcor_jja_pval < 0.05, na.rm = TRUE),
    sum(ntrallcor_ann_pval < 0.05, na.rm = TRUE),
    sum(ntrallcor_jja_pval < 0.05, na.rm = TRUE)
  ),
  median_r = c(
    median(p2kallcor_ann, na.rm = TRUE),
    median(p2kallcor_jja, na.rm = TRUE),
    median(ntrallcor_ann, na.rm = TRUE),
    median(ntrallcor_jja, na.rm = TRUE)
  )
)

ggplot(screen_df, aes(x = season, y = n_sig, fill = dataset)) +
  geom_col(position = position_dodge(width = 0.7), width = 0.6) +
  geom_text(aes(label = n_sig), position = position_dodge(width = 0.7), vjust = -0.3, size = 3.5) +
  labs(
    title = "Effect of season choice on screening",
    subtitle = "Number of proxies with p < 0.05 vs local CRUTEM grid box (1901–2012)",
    x = NULL, y = "Count (p < 0.05)", fill = NULL
  ) +
  theme_clim()
```

For PAGES2k, JJA correlations are slightly higher (`r round(median(p2kallcor_jja, na.rm=TRUE), 2)` vs `r round(median(p2kallcor_ann, na.rm=TRUE), 2)` median r). For NTREND the pattern is similar. **Lesson:** always match proxy seasonality to the target variable.

## 3. Record length (degrees of freedom)

The p-value of a correlation depends on **how many independent years** overlap. Short records → few degrees of freedom → hard to reach significance even with decent r.

```{r df histogram}
df_df <- rbind(
  data.frame(dataset = "PAGES2k JJA", df = p2kallcor_jja_df),
  data.frame(dataset = "NTREND JJA", df = ntrallcor_jja_df)
)

ggplot(df_df, aes(x = df, fill = dataset)) +
  geom_histogram(bins = 25, alpha = 0.75, position = "identity") +
  facet_wrap(~dataset, ncol = 1, scales = "free_y") +
  labs(
    title = "Effective sample size behind each correlation test",
    subtitle = "Degrees of freedom from cor.test() — longer overlap = more reliable screening",
    x = "Degrees of freedom", y = "Count"
  ) +
  theme_clim() +
  theme(legend.position = "none")
```

## 4. Outliers (> 5 standard deviations)

Extreme values can dominate correlations and regressions. We flag proxy years more than **5 SD** from the series mean.

```{r outlier count}
number_extr <- apply(p2kdata$data, 2, function(x) {
  pmean <- mean(x, na.rm = TRUE)
  psd   <- stats::sd(x, na.rm = TRUE)
  if (!is.finite(psd) || psd == 0) return(0L)
  sum(abs(x - pmean) > 5 * psd, na.rm = TRUE)
})

cat("PAGES2k records with at least one >5SD value:", sum(number_extr > 0), "\n")
cat("Total flagged years:", sum(number_extr), "\n")
```

## 5. Inspect flagged series

```{r outlier plots}
pos <- which(number_extr > 0)
# Show metadata for the first few flagged records (name, archive type, coordinates)
if (length(pos) > 0) {
  show_idx <- pos[seq_len(min(4, length(pos)))] + 1L  # +1: first meta column is row label
  print(t(p2kmeta[c(1, 2, 3, 4), show_idx, drop = FALSE]))
}

par(mfrow = c(2, 2), mar = c(3, 3, 2, 1))
n_show <- min(4, length(pos))
for (i in seq_len(n_show)) {
  p_idx <- pos[i]
  plot(
    p2kdata$time, p2kdata$data[, p_idx], type = "l",
    main = paste0(p2kmeta[1, p_idx + 1L], " (", number_extr[p_idx], " extremes)"),
    xlab = "Year", ylab = "Proxy value"
  )
  pmean <- mean(p2kdata$data[, p_idx], na.rm = TRUE)
  psd   <- stats::sd(p2kdata$data[, p_idx], na.rm = TRUE)
  abline(h = pmean + c(-5, 5) * psd, col = "red", lty = 2)
}
par(mfrow = c(1, 1))
```

**Discussion:** Outliers may be real (volcanic cooling, drought) or artefacts (measurement error, wrong dating, unit mix-up, compositing). They should be checked manually — automatic removal is risky.

## 6. Effect of trends on correlation

Pearson r measures **linear co-variation around the mean**. If both series share a strong trend, r can be **inflated** even when short-term variability is unrelated (see "replication crisis" papers linked in the exercise).

```{r trend demo}
set.seed(42)
testdata1 <- stats::rnorm(100)
testdata2 <- stats::rnorm(100)
cat("Without shared trend: r =", round(stats::cor(testdata1, testdata2), 3), "\n")

trenddata <- seq(0, 2, length.out = 100)
testdata1tr <- testdata1 + trenddata
testdata2tr <- testdata2 + trenddata
cat("With shared trend:    r =", round(stats::cor(testdata1tr, testdata2tr), 3), "\n")

ggplot(data.frame(x = testdata1, y = testdata2), aes(x, y)) +
  geom_point(alpha = 0.6) +
  labs(title = "No trend — correlation near zero", x = "Series 1", y = "Series 2") +
  theme_clim()
```

**Take-home:** screening on **detrended** or **high-pass filtered** data is often more appropriate for climate reconstruction than raw correlation over the 20th century warming period.

---

Further reading: Neukom et al. (screening), Santer et al. (trends vs correlation).
