---
title: "Validation of spatial reconstruction"
output: html_notebook
---

#### Exercise 15: Validation of spatial reconstruction

## What is the task?

In **Exercise 6** you fitted a regression on the **calibration period** and evaluated it on an **independent validation period** — without re-fitting. Same logic here:

| Exercise 6 (stations) | Exercise 15 (CFR) |
|-----------------------|-------------------|
| Target: EU mean temperature | Target: CRUTEM JJA anomaly at each grid cell |
| Predictor: station matrix | Predictor: proxy network (already used inside `Master_CFR.R`) |
| **Observed** = CRUTEM in validation years | **Observed** = `reconstruction.target` |
| **Predicted** = `predict(fit, Xval)` | **Predicted** = `recon` from Exercise 14 (PCR output for 750–2011) |

The CFR was **calibrated on 1941–2000** in Exercise 14. We must **not** re-run the reconstruction. We only ask:

> How well does the existing reconstruction match CRUTEM in **1901–1940** — a period that was **not** used for calibration?

Per grid cell we compute **r**, **RMSE**, and **RMSESS** exactly as in Exercise 6:

- `RMSE_clim` = RMSE of a forecast that always predicts the **local mean of the validation-period observations**
- `RMSESS = 1 − RMSE / RMSE_clim`

```{r setup}
rm(list = ls())
if (dir.exists("Exercises")) setwd("Exercises")

library(ggplot2)
source("functions_clim_recon.R")
load("reconstruction_input_and_output.RData")

theme_clim <- function(base_size = 13) {
  theme_minimal(base_size = base_size, base_family = "sans") +
    theme(
      plot.title = element_text(face = "bold", hjust = 0),
      plot.subtitle = element_text(colour = "grey30", hjust = 0, size = rel(0.9)),
      strip.text = element_text(face = "bold", hjust = 0),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      panel.grid = element_blank(),
      legend.title = element_blank()
    )
}

val_years <- 1901:1940   # independent validation (before calibration)
cal_years <- 1941:2000   # calibration period used in Master_CFR.R
min_years <- 30L         # need enough paired years for stable correlations
```

##### 1. Grid-cell validation — explicit code (following Exercise 6)

```{r validation loop}
nx <- length(recon$lon)
ny <- length(recon$lat)

# Result arrays: one value per grid cell
r_cal <- r_val <- rmse_val <- rmsess <- array(NA_real_, dim = c(nx, ny))
rmse_cal <- rmsess_cal <- array(NA_real_, dim = c(nx, ny))

# --- helper: same formulas as Exercise 6, but for one grid cell ---
score_cell <- function(obs, pred) {
  ok <- complete.cases(obs, pred)
  if (sum(ok) < min_years) {
    return(list(r = NA_real_, rmse = NA_real_, rmsess = NA_real_, n = sum(ok)))
  }
  obs <- obs[ok]
  pred <- pred[ok]
  rmse <- sqrt(mean((obs - pred)^2))
  clim <- mean(obs)
  rmse_clim <- sqrt(mean((obs - clim)^2))
  list(
    r = stats::cor(obs, pred),
    rmse = rmse,
    rmsess = if (rmse_clim > 0) 1 - rmse / rmse_clim else NA_real_,
    n = sum(ok)
  )
}

val_idx_recon <- match(val_years, recon$time)
val_idx_tgt   <- match(val_years, reconstruction.target$time)
cal_idx_recon <- match(cal_years, recon$time)
cal_idx_tgt   <- match(cal_years, reconstruction.target$time)

for (i in seq_len(nx)) {
  for (j in seq_len(ny)) {
    # Validation period (main result of this exercise)
    obs_val <- reconstruction.target$data[i, j, val_idx_tgt]
    pred_val <- recon$data[i, j, val_idx_recon]
    s_val <- score_cell(obs_val, pred_val)
    r_val[i, j]    <- s_val$r
    rmse_val[i, j] <- s_val$rmse
    rmsess[i, j]   <- s_val$rmsess

    # Calibration period (for comparison maps — how well do we fit the tuning period?)
    obs_cal <- reconstruction.target$data[i, j, cal_idx_tgt]
    pred_cal <- recon$data[i, j, cal_idx_recon]
    s_cal <- score_cell(obs_cal, pred_cal)
    r_cal[i, j]      <- s_cal$r
    rmse_cal[i, j]   <- s_cal$rmse
    rmsess_cal[i, j] <- s_cal$rmsess
  }
}

cat("Grid cells with validation skill (≥", min_years, "years):", sum(!is.na(r_val)), "of", nx * ny, "\n")
```

```{r summary table}
skill_summary <- data.frame(
  period = rep(c("Validation 1901–1940", "Calibration 1941–2000"), each = 3),
  metric = rep(c("median r", "median RMSE (°C)", "median RMSESS"), 2),
  value = c(
    median(r_val, na.rm = TRUE),
    median(rmse_val, na.rm = TRUE),
    median(rmsess, na.rm = TRUE),
    median(r_cal, na.rm = TRUE),
    median(rmse_cal, na.rm = TRUE),
    median(rmsess_cal, na.rm = TRUE)
  )
)
skill_summary$value <- round(skill_summary$value, 3)
print(skill_summary)
```

**Read the numbers carefully:** median grid-cell **r ≈ 0.2–0.3** is modest — many cells are poorly constrained by tree rings (especially oceans, where recon is often `NA`). That is very different from Exercise 6 (station regression), where **r_val ≈ 0.9** was achievable.

##### 2. Domain mean — use the index reconstruction from Master_CFR

For the **NH extratropics spatial mean**, do not simply average all grid cells. Exercise 14 already produced `spatmean_recon` (latitude-weighted index from the workflow). Compare that to the **latitude-weighted CRUTEM mean** — same definition as in `Master_CFR.R`.

```{r lat-weighted domain mean}
lat_w <- cos(reconstruction.target$lat * pi / 180)^0.5
W <- outer(rep(1, nx), lat_w)

weighted_field_mean <- function(field, time_idx) {
  vapply(time_idx, function(t) {
    slice <- field[, , t]
    ok <- !is.na(slice)
    if (!any(ok)) return(NA_real_)
    sum(slice[ok] * W[ok]) / sum(W[ok])
  }, numeric(1))
}

obs_index  <- weighted_field_mean(reconstruction.target$data, val_idx_tgt)
pred_index <- as.numeric(spatmean_recon[val_idx_recon])
ok_idx     <- complete.cases(obs_index, pred_index)

r_index      <- stats::cor(obs_index[ok_idx], pred_index[ok_idx])
rmse_index   <- sqrt(mean((obs_index[ok_idx] - pred_index[ok_idx])^2))
clim_index   <- mean(obs_index[ok_idx])
rmse_clim_ix <- sqrt(mean((obs_index[ok_idx] - clim_index)^2))
rmsess_index <- 1 - rmse_index / rmse_clim_ix

cat("Latitude-weighted index validation 1901–1940:\n")
cat("  r =", round(r_index, 3), "| RMSE =", round(rmse_index, 3), "°C | RMSESS =", round(rmsess_index, 3), "\n")
```

This index skill is usually **much higher** than the median grid-cell skill — the mean integrates information across proxy-rich land areas.

```{r index ts plot}
ts_ix <- data.frame(
  year = val_years[ok_idx],
  value = c(obs_index[ok_idx], pred_index[ok_idx]),
  series = rep(c("CRUTEM (lat-weighted)", "CFR index (spatmean_recon)"), each = sum(ok_idx))
)

ggplot(ts_ix, aes(x = year, y = value, colour = series)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_line(linewidth = 0.75) +
  scale_colour_manual(values = c("CRUTEM (lat-weighted)" = "#2c6e8a", "CFR index (spatmean_recon)" = "#c0392b")) +
  labs(
    title = "Independent validation — latitude-weighted spatial mean",
    subtitle = sprintf("1901–1940 | r = %.2f | RMSESS = %.2f", r_index, rmsess_index),
    x = "Year", y = "JJA anomaly (°C)", colour = NULL
  ) +
  theme_clim() +
  theme(legend.position = "top")
```

##### 3. Skill maps (as in the original exercise template)

```{r map helpers}
world <- load_world_coastline("../Data/world_coastline.dat")

skill_to_df <- function(arr) {
  df <- expand.grid(lon = recon$lon, lat = recon$lat)
  df$value <- as.vector(arr)
  df
}

plot_skill_map <- function(arr, title, limits, mid = 0, palette = "diverging") {
  df <- skill_to_df(arr)
  p <- ggplot() +
    geom_tile(data = df, aes(x = lon, y = lat, fill = value)) +
    geom_path(data = world, aes(x = lon, y = lat, group = seg), inherit.aes = FALSE,
              colour = "grey35", linewidth = 0.15) +
    coord_fixed(ratio = 1.35, xlim = c(lon_recon_W, lon_recon_E), ylim = c(lat_recon_S, lat_recon_N)) +
    labs(title = title, x = "Longitude", y = "Latitude", fill = NULL) +
    theme_clim()
  if (palette == "diverging") {
    p + scale_fill_gradient2(
      low = "#2166ac", mid = "white", high = "#b2182b",
      midpoint = mid, limits = limits, oob = scales::squish, na.value = "grey92"
    )
  } else {
    p + scale_fill_distiller(
      palette = "YlOrRd", direction = 1,
      limits = limits, oob = scales::squish, na.value = "grey92"
    )
  }
}
```

```{r skill maps}
rmse_cap <- stats::quantile(rmse_val, 0.95, na.rm = TRUE)

plot_skill_map(r_cal, "r — calibration (1941–2000)", limits = c(-1, 1))
plot_skill_map(r_val, "r — validation (1901–1940)", limits = c(-1, 1))
plot_skill_map(rmse_val, "RMSE — validation (1901–1940)", limits = c(0, rmse_cap), palette = "sequential")
plot_skill_map(rmsess, "RMSESS — validation (1901–1940)", limits = c(-0.5, 1))
```

Grey cells = no skill computed (too few overlapping years or missing recon).

##### 4. Calibration vs validation

```{r cal vs val facet}
compare_df <- rbind(
  data.frame(skill_to_df(r_cal), period = "Calibration 1941–2000"),
  data.frame(skill_to_df(r_val), period = "Validation 1901–1940")
)

ggplot(compare_df, aes(x = lon, y = lat, fill = value)) +
  geom_tile() +
  geom_path(data = world, aes(x = lon, y = lat, group = seg), inherit.aes = FALSE,
            colour = "grey35", linewidth = 0.12) +
  scale_fill_gradient2(
    low = "#2166ac", mid = "white", high = "#b2182b",
    midpoint = 0, limits = c(-1, 1), oob = scales::squish, na.value = "grey92",
    name = "Correlation r"
  ) +
  facet_wrap(~period, ncol = 2) +
  coord_fixed(ratio = 1.2) +
  labs(
    title = "Where does the reconstruction work?",
    subtitle = "Compare proxy-rich land (Europe, N America) to unconstrained regions",
    x = "Longitude", y = "Latitude"
  ) +
  theme_clim()
```

## Interpretation

1. **Validation is the point** — calibration maps show how the method fits the tuning period; validation maps show whether it generalises to **1901–1940**.
2. **Use the right target for the domain mean** — `spatmean_recon` vs latitude-weighted CRUTEM, not an unweighted grid average.
3. **Grid-cell skill is often weak** — PCR spreads proxy information smoothly; many cells have no local proxy constraint (`NA` in recon).
4. **RMSESS < 0** at a cell means: predicting the local validation mean would have been better than using the reconstruction.
