---
title: "Reconstruction challenge"
output: html_notebook
---

### Exercise 7: Who makes the best European reconstruction
  
Empty memory, set working directory and load functions, libraries and data from previous exercises

```{r}
rm(list=ls())
if (dir.exists("Exercises")) setwd("Exercises")
library(ggplot2)
load(file = "../Data/workspace_ex6.RData")
source("functions_clim_recon.R")

theme_clim <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(size = rel(1.25), face = "bold", hjust = 0, margin = margin(b = 6)),
      plot.subtitle = element_text(size = rel(0.95), hjust = 0, colour = "grey25", margin = margin(b = 10)),
      plot.margin = margin(12, 16, 12, 12),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.35),
      legend.position = "top",
      legend.title = element_blank()
    )
}
col_obs <- "#1a1a1a"
col_pred <- "#c0392b"
```

#### Challenge: Which group makes the European reconstruction with best validation scores using 3 stations only

**What exactly are we optimizing?**

| Step | What happens |
|------|----------------|
| **Target Y** | Lat-weighted CRUTEM European mean (`EUlonggridmean_latwgt`) — assumed “truth” |
| **Calibration** | Fit regression on **1951–2015 only** → learn coefficients |
| **Validation** | Apply same coefficients to **1901–1950** → independent test |
| **Search** | Try all `choose(129, 3) = 349,504` station triplets |
| **Objective** | **One metric only** — we use **`r_val`** (validation correlation) |

`RMSE` and `RMSESS` are **reported** for every candidate but **not** combined into a single score.  
If you optimized `rmsess` instead, you would get a **different** triplet (see below).

```{r}
# -------------------------------------------------------------------------
# COMPETITION SETUP
# -------------------------------------------------------------------------
Y <- as.numeric(EUlonggridmean_latwgt)   # reconstruction target (grid-box EU mean)
years <- station_abs_annmean$time
cal_idx <- which(years > 1950)           # fit coefficients HERE (1951-2015)
val_idx <- which(years < 1951)           # score models HERE (1901-1950) — never used for fitting

OPTIMIZATION_METRIC <- "r_val"           # ONLY this number is maximized in the search
                                         # alternatives: "rmsess" (not a combination!)

cache_file <- paste0("../Data/best3_exhaustive_", OPTIMIZATION_METRIC, ".Rdata")

if (file.exists(cache_file)) {
  load(cache_file)
  message("Loaded cached result (metric = ", best3$metric, ")")
} else {
  message("Exhaustive search (~5 min), optimizing ", OPTIMIZATION_METRIC, " ...")
  best3 <- find_best_stations(
    Y = Y,
    station_matrix = station_longEU,
    cal_idx = cal_idx,
    val_idx = val_idx,
    k = 3L,
    metric = OPTIMIZATION_METRIC
  )
  save(best3, file = cache_file)
}

# GHCN station names (match longEU coordinates to full GHCN metadata)
load("../Data/ghcn_abs_annmean_1901-2015.Rdata")
ghcn_name_at <- function(lon, lat) {
  d <- (ghcn_abs_annmean$lon - lon)^2 + (ghcn_abs_annmean$lat - lat)^2
  ghcn_abs_annmean$names[which.min(d)]
}

best3$idx
station_table <- data.frame(
  idx = best3$idx,
  GHCN = mapply(ghcn_name_at, station_longEUlon[best3$idx], station_longEUlat[best3$idx]),
  lon = round(station_longEUlon[best3$idx], 2),
  lat = round(station_longEUlat[best3$idx], 2)
)
station_table
best3$combos_tested

c(
  r_cal = best3$r_cal,
  r_val = best3$r_val,
  rmse_val = best3$rmse_val,
  rmsess = best3$rmsess,
  n_val = best3$n_val
)
summary(best3$fit)
```

**Winning 3-station network (exhaustive optimum):**

| # | Index | GHCN ID | Lon | Lat | Location |
|---|-------|---------|-----|-----|----------|
| 1 | 115 | SION_2 | 7.33°E | 46.22°N | **Sion, Switzerland** |
| 2 | 66 | LARISSA | 22.45°E | 39.65°N | **Larissa, Greece** |
| 3 | 34 | TARTU | 26.73°E | 58.30°N | **Tartu, Estonia** |

**Validation scores:** r = **0.960**, RMSE = **0.127 °C**, RMSESS = **0.683** (49 of 50 validation years used).

Compared to the previous heuristic 3-station set (Ljubljana / Utsira / Kursk, idx 100/87/98): r_val = 0.927, RMSESS = 0.598 — the exhaustive model is clearly better.


```{r}
# Same search, but optimize RMSESS instead of r_val — different objective, different winner
cache_rmsess <- "../Data/best3_exhaustive_rmsess.Rdata"
if (file.exists(cache_rmsess)) {
  load(cache_rmsess)   # loads best3_rmsess
} else {
  best3_rmsess <- find_best_stations(Y, station_longEU, cal_idx, val_idx, k = 3L, metric = "rmsess")
  save(best3_rmsess, file = cache_rmsess)
}

metric_compare <- data.frame(
  objective = c("r_val (used for competition)", "rmsess (alternative)"),
  stations = c(
    paste(best3$idx, collapse = ", "),
    paste(best3_rmsess$idx, collapse = ", ")
  ),
  r_val = c(best3$r_val, best3_rmsess$r_val),
  rmse_val = c(best3$rmse_val, best3_rmsess$rmse_val),
  rmsess = c(best3$rmsess, best3_rmsess$rmsess)
)
metric_compare
```

**Note:** Optimizing `rmsess` picks **Graz + Tórshavn + Tartu** (idx 6, 32, 34) — slightly **lower** r (0.953) but **lower** RMSE and **higher** RMSESS. There is no single “best” model unless the course defines which metric wins.


```{r}
# Compare optimal vs previous heuristic model
heuristic_idx <- c(100, 87, 98)
heuristic <- validate_station_regression(Y, station_longEU, cal_idx, val_idx, heuristic_idx)

compare_df <- data.frame(
  model = c("Exhaustive optimum", "Heuristic (Ex. 5)"),
  r_val = c(best3$r_val, heuristic$r_val),
  rmse_val = c(best3$rmse_val, heuristic$rmse_val),
  rmsess = c(best3$rmsess, heuristic$rmsess),
  n_val = c(best3$n_val, heuristic$n_val)
)
compare_df

ggplot(compare_df, aes(x = model, y = r_val, fill = model)) +
  geom_col(width = 0.55, show.legend = FALSE) +
  geom_text(aes(label = sprintf("r = %.3f", r_val)), vjust = -0.5, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("Exhaustive optimum" = "#1e8449", "Heuristic (Ex. 5)" = "#95a5a6")) +
  scale_y_continuous(limits = c(0, 1.05), breaks = seq(0, 1, 0.2), expand = c(0, 0)) +
  labs(
    title = "Competition result: 3-station models",
    subtitle = "Validation period 1901–1950",
    x = NULL, y = "Correlation r"
  ) +
  theme_clim()
```

```{r}
# Validation time series: exhaustive optimum
opt <- validate_station_regression(Y, station_longEU, cal_idx, val_idx, best3$idx)
val_years <- years[val_idx]

val_plot <- data.frame(
  year = rep(val_years, 2),
  value = c(Y[val_idx], opt$pred_val),
  series = rep(c("Observed (CRUTEM)", "Optimal 3-station forecast"), each = length(val_years))
)
val_plot <- val_plot[complete.cases(val_plot$value), ]

ggplot(val_plot, aes(x = year, y = value, colour = series)) +
  geom_line(linewidth = 0.8) +
  scale_colour_manual(values = c("Observed (CRUTEM)" = col_obs, "Optimal 3-station forecast" = col_pred)) +
  labs(
    title = "Optimal 3-station reconstruction",
    subtitle = sprintf("Validation 1901–1950  |  r = %.3f  |  RMSE = %.3f °C  |  RMSESS = %.3f",
                       opt$r_val, opt$rmse_val, opt$rmsess),
    x = NULL, y = "Annual mean temperature (°C)"
  ) +
  theme_clim()
```

**Answer:** The competition-winning model uses **Sion (CH), Larissa (GR), and Tartu (EE)** — GHCN indices **115, 66, 34** — found by testing **all 349,504** possible 3-station subsets and picking the one with highest validation correlation. (Coordinates alone are easy to misread: 7.3°E / 46.2°N is Switzerland, not Germany.) The three sites span the Alps, the eastern Mediterranean, and the Baltic, capturing independent large-scale temperature variability. With only three predictors and 57 calibration years, the model is well-conditioned and generalises strongly to 1901–1950.


Save result for submission / next exercises
```{r}
idx_competition <- best3$idx
fit_competition <- best3$fit
save(idx_competition, fit_competition, best3, compare_df,
     file = "../Data/workspace_ex7.RData")
```

