---
title: "Exercise 8: Generating instrumental pseudoproxies"
output:
  bookdown::html_document2: default
    df_print: paged
---

**Aims:**

-   Understanding the concept and behaviour of noise in paleoclimate proxy data

-   Generate "pseudoproxy" time series by adding noise to instrumental climate time series

**Background**

Climate proxy data contain a climate signal that is 'hidden' behind a substantial amount of noise. This noise is all factors, that contribute to the formation of the climate archive (e.g. a tree-ring), but are not related to climate.

As a statistical concept, the climate proxy record can be understood as consisting as a climate signal plus noise. For example:

`Equation 1`

$$P = T + N$$ Where $P$ is the proxy record, $T$ ist temperature and $N$ is all the non-climatic noise.

We can thus generate a Pseudoproxy by adding a certain amount of Noise to an instrumental temperature time series.

Typically, pseudoproxies have a signal-to-noise-ratio $SNR$ of around 0.4, which means that the standard deviation of the noise is about 2.5 times larger than the standard deviation of the signal.

`Equation 2`

$$SNR = \frac{\sigma_{\text{climate signal}}}{\sigma_{\text{noise}}}$$

We will now perform all steps to generate realistic pseudoproxy time series based on instrumental data and noise

#### 8.1 load instrumental data

Use the following section of code to load the station data used in day 1 (long European stations).

```{r first load workspace from last exercise}
rm(list=ls())
if (dir.exists("Exercises")) setwd("Exercises")
library(ggplot2)
load("../Data/workspace_ex6.RData")
source("functions_clim_recon.R")

theme_clim <- function(base_size = 13) {
  theme_minimal(base_size = base_size) +
    theme(
      plot.title = element_text(size = rel(1.25), face = "bold", hjust = 0, margin = margin(b = 6)),
      plot.subtitle = element_text(size = rel(0.95), hjust = 0, colour = "grey25", margin = margin(b = 10)),
      plot.margin = margin(12, 16, 12, 12),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      panel.grid.major.y = element_line(colour = "grey92", linewidth = 0.35),
      legend.position = "top",
      legend.title = element_blank()
    )
}
col_signal <- "#1a1a1a"
col_noise  <- "#7f8c8d"
col_proxy  <- "#2c6e8a"
```

Select a station with no missing data to be used for the pseudoproxy exercise.

```{r load and select instrumental data}
# Count missing values per European station (station_EU from Ex. 4)
noNA <- apply(station_EU, 2, function(x) sum(is.na(x)))
noNAstats <- which(noNA == 0)   # 24 stations with complete 1901-2015 records

# Pick station 8 (Fichtelberg, DE) — not the first, to make 8.6 harder for other groups
selstat_idx <- noNAstats[8]
selstat <- station_EU[, selstat_idx]
years <- station_abs_annmean$time

load("../Data/ghcn_abs_annmean_1901-2015.Rdata")
ghcn_name <- ghcn_abs_annmean$names[which.min(
  (ghcn_abs_annmean$lon - station_abs_annmean$lon[posEUstat[selstat_idx]])^2 +
  (ghcn_abs_annmean$lat - station_abs_annmean$lat[posEUstat[selstat_idx]])^2
)]
ghcn_name

ggplot(data.frame(year = years, temp = selstat), aes(x = year, y = temp)) +
  geom_line(colour = col_signal, linewidth = 0.8) +
  labs(
    title = "Selected instrumental station",
    subtitle = paste0(ghcn_name, "  (complete record 1901–2015)"),
    x = NULL, y = "Annual mean temperature (°C)"
  ) +
  theme_clim()
```

#### 8.2 Generate white noise time series

Normally distributed random (white noise) data can be generated in R using the "rnorm" function.

Use the following section of code to learn working with this function and generate a 100-year time series with mean 0 and standard deviation 1.

```{r gnenerate white noise}
# rnorm(n, mean, sd) draws n values from N(mean, sd^2)
whitenoise <- rnorm(100, mean = 0, sd = 1)

ggplot(data.frame(t = 1:100, x = whitenoise), aes(x = t, y = x)) +
  geom_line(colour = col_noise, linewidth = 0.6) +
  labs(title = "Example white noise", subtitle = "100 years, mean 0, sd 1",
       x = "Time", y = "Value") +
  theme_clim() +
  theme(panel.grid.major.y = element_blank())
```

#### 8.3 identify and generate correct amount of noise to be added to instrumental data

Now we need to find the correct magnitude of the noise, which is characterized by its standard deviation. For this we work with equation 1 above.

As target Signal-to-noise-ratio we use 0.4.

Use the following section of code to get the std.dev. of the instrumental data, use the target SNR to define the std.dev. of the noise to be generated and make a corresponding noise time series.

```{r get target std.dev. of noise and generate noise}
TARGET_SNR <- 0.4

sd_signal <- sd(selstat, na.rm = TRUE)
sd_noise  <- sd_signal / TARGET_SNR          # from SNR = sd_signal / sd_noise
var_noise <- sd_noise^2

set.seed(8)   # reproducible noise for this station
noise_white <- rnorm(length(selstat), mean = 0, sd = sd_noise)

c(sd_signal = sd_signal, sd_noise = sd_noise, SNR_check = sd_signal / sd_noise)
```

#### 8.4 Generate white noise pseudoproxy

Now the noise simply needs to be added to the instrumental data to generate the pseudoproxy.

Use the following section of Code to do this and make a plot comparing the original instrumental data, the "non-climatic-noise" and the pseudoproxy times series.

What are your observations and interpretations when comparing the different time series? What is the correlation between the station data and the pseudoproxy?

```{r make and plot white noise pseudoproxy}
white.pseudoproxy <- selstat + noise_white

plot_df <- rbind(
  data.frame(year = years, value = selstat, series = "Instrumental (signal)"),
  data.frame(year = years, value = noise_white, series = "White noise"),
  data.frame(year = years, value = white.pseudoproxy, series = "White-noise pseudoproxy")
)

ggplot(plot_df, aes(x = year, y = value, colour = series)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(
    "Instrumental (signal)" = col_signal,
    "White noise" = col_noise,
    "White-noise pseudoproxy" = col_proxy
  )) +
  labs(
    title = "White-noise pseudoproxy",
    subtitle = sprintf("SNR = 0.4  |  cor(signal, proxy) = %.2f", cor(selstat, white.pseudoproxy)),
    x = NULL, y = "Temperature / noise (°C)", colour = NULL
  ) +
  theme_clim()
```

```{r get correlation between the station data and the pseudoproxy}
cor(selstat, white.pseudoproxy)
```

**Answer (8.4):** With SNR = 0.4 the noise is ~2.5× stronger than the climate signal (sd). The pseudoproxy looks much noisier than the instrumental series; interannual and decadal patterns are barely visible. Correlation is low (r ≈ 0.3–0.4), because most variance in P comes from N, not T — as in real tree-ring or documentary proxies.


#### 8.5 improve spectral properties of the noise by adding autoregressive model

In reality, the non-climatic noise inherent in proxy data is barely white: The variance at different frequencies is not constant. Mostly, the noise is reddish, which means that it has higher variance at lower frequencies, i.e. high autocorrelation.

Think, for example, of a tree-ring, where the site-characteristics, a non-climatic factor, do not completely change every year (e.g. shading, soil conditions, competition, ...), but remain relatively stable and change slowly.

Therefore, we modify our noise to have more realistic autocorrelation. We do so by adding a simple 1st order autocorrelation (AR1) process to the data:

`Equation 3`

$$N(t) = \rho \cdot N(t-1) + r$$

$N(t)$ is the noise at time $t$, $\rho$ is the target 1st order autoregression (AR1) coefficient and $r$ is random white noise.

For the final variance of the noise to be correct, we have to adjust the variance of $r$ by:

`Equation 4`

$$\operatorname{var}(r) = \operatorname{var}(N) \cdot (1 - \rho^2)$$ Where $var(N)$ ist the target variance of the Noise, calculated above in step 8.3.

For the moment we work with a target AR1 coefficient of the noise ($\rho$) of 0.6

Use the following section of code to generate an AR1 Noise time series based on Equations 2-4 and add it to the instrumental time series to get your final pseudoproxy. Then compare the white-noise and AR1 pseudoproxies and the instrumental time series. What is the correlation between the station data and the AR1 pseudoproxy?

```{r generate red noise pseudoproxy}
rho <- 0.6
var_r <- var_noise * (1 - rho^2)           # Equation 4

set.seed(8)
whitenoise.for.ar1 <- rnorm(length(selstat), mean = 0, sd = sqrt(var_r))

# Build AR(1) noise (Equation 3)
arnoise <- numeric(length(selstat))
arnoise[1] <- whitenoise.for.ar1[1]
for (t in 2:length(selstat)) {
  arnoise[t] <- rho * arnoise[t - 1] + whitenoise.for.ar1[t]
}

ar1.pseudoproxy <- selstat + arnoise

compare_df <- rbind(
  data.frame(year = years, value = selstat, series = "Instrumental (signal)"),
  data.frame(year = years, value = white.pseudoproxy, series = "White-noise pseudoproxy"),
  data.frame(year = years, value = ar1.pseudoproxy, series = "AR(1) pseudoproxy")
)

ggplot(compare_df, aes(x = year, y = value, colour = series)) +
  geom_line(linewidth = 0.7) +
  scale_colour_manual(values = c(
    "Instrumental (signal)" = col_signal,
    "White-noise pseudoproxy" = col_noise,
    "AR(1) pseudoproxy" = col_proxy
  )) +
  labs(
    title = "White vs AR(1) pseudoproxy",
    subtitle = sprintf("rho = 0.6, SNR = 0.4  |  r(white) = %.2f, r(AR1) = %.2f",
                       cor(selstat, white.pseudoproxy), cor(selstat, ar1.pseudoproxy)),
    x = NULL, y = "Temperature (°C)", colour = NULL
  ) +
  theme_clim()
```

```{r correlate time series}
c(
  cor_white = cor(selstat, white.pseudoproxy),
  cor_ar1   = cor(selstat, ar1.pseudoproxy),
  sd_noise_ar1 = sd(arnoise)
)
```

**Answer (8.5):** AR(1) noise is smoother (more low-frequency variance) than white noise — similar to persistent site effects in trees. With the same target SNR and variance, correlation with the signal is still low and similar to the white-noise case. The main difference is spectral: AR(1) pseudoproxies can show spurious multi-year trends from noise alone, which makes reconstruction harder.

It is recommended to save your workspace to have the data ready for the next exercises.

```{r save workspace}
PSEUDOPROXY_SNR  <- TARGET_SNR
PSEUDOPROXY_RHO  <- rho
selstat_idx
selstat
ghcn_name

save.image("../Data/workspace_ex8.RData")
```

# Additional optional exercices

Convert the above code into an R function that allows you to generate noise based on an input time series, SNR and AR 1 coefficient.

Try out various AR1 coefficients for the noise. How do the Noise time series look like and how do they compare to the instrumental data? How does this effect the resulting pseudoproxies?

```{r optional AR1 sensitivity}
rho_values <- c(0, 0.3, 0.6, 0.9)
set.seed(8)

sens_df <- do.call(rbind, lapply(rho_values, function(r) {
  proxy <- if (r == 0) {
    create.white.pseudoproxy(selstat, snr = TARGET_SNR)
  } else {
    create.ar1.acf.pseudoproxy(selstat, rho = r, snr = TARGET_SNR)
  }
  data.frame(
    year = years,
    value = proxy,
    rho = factor(r, levels = rho_values, labels = paste0("rho = ", rho_values))
  )
}))

ggplot(sens_df, aes(x = year, y = value, colour = rho)) +
  geom_line(linewidth = 0.7) +
  labs(
    title = "Effect of noise autocorrelation on pseudoproxies",
    subtitle = paste0(ghcn_name, "  |  SNR = 0.4"),
    x = NULL, y = "Pseudoproxy (°C)", colour = "AR(1) coefficient"
  ) +
  theme_clim()
```

**Answer (optional):** Higher $\rho$ makes noise smoother and can create long pseudo-trends unrelated to climate. Correlation with the true signal stays weak at SNR = 0.4, but reconstructions become harder because noise mimics low-frequency climate variability.

#### Challenge 8.6: Send your pseudoproxy to the other groups. Can you find out, which station the other group's pseudoproxies are based on?

**Our station:** Fichtelberg (GHCN), complete European record, index `noNAstats[8]` in `station_EU`. Other groups can try correlating your sent series with candidate stations or matching variance/SNR fingerprints.

