---
title: "Climate field reconstruction with real data"
output: html_notebook
---

#### Exercise 14: Perform a climate field reconstruction

**What we do here:** We combine a **proxy network** (NTREND tree rings) with **instrumental gridded temperatures** (CRUTEM) to reconstruct past climate fields back to AD 750. The method is **PCR** (Principal Component Regression) — the same approach used in published NTREND reconstructions.

```{r setup}
rm(list = ls())
if (dir.exists("Exercises")) setwd("Exercises")

library(ggplot2)
source("functions_clim_recon.R")
load(file = "../Data/workspace_ex11.RData")

# Consistent plot style (same as Exercises 6 & 8)
theme_clim <- function(base_size = 13) {
  theme_minimal(base_size = base_size, base_family = "sans") +
    theme(
      plot.title = element_text(size = rel(1.2), face = "bold", hjust = 0, margin = margin(b = 6)),
      plot.subtitle = element_text(size = rel(0.92), hjust = 0, colour = "grey30", margin = margin(b = 10)),
      plot.margin = margin(12, 16, 12, 12),
      panel.background = element_rect(fill = "white", colour = NA),
      plot.background = element_rect(fill = "white", colour = NA),
      panel.grid = element_blank(),
      legend.position = "top",
      legend.title = element_blank()
    )
}

col_target <- "#2c6e8a"
col_recon  <- "#c0392b"
col_index  <- "#7f8c8d"
```

##### 1. Define target domain and proxy network

```{r define input datasets}
# Northern Hemisphere extratropics: 40–90°N, full longitude
lon_recon_W <- -180
lon_recon_E <- 180
lat_recon_S <- 40
lat_recon_N <- 90

lonsel_recon <- which(gridded_anom_jjamean$lon >= lon_recon_W & gridded_anom_jjamean$lon <= lon_recon_E)
latsel_recon <- which(gridded_anom_jjamean$lat >= lat_recon_S & gridded_anom_jjamean$lat <= lat_recon_N)

# Crop CRUTEM JJA anomalies to the reconstruction box
gridded_anom_jjamean_NHET <- gridded_anom_jjamean
gridded_anom_jjamean_NHET$data <- gridded_anom_jjamean_NHET$data[lonsel_recon, latsel_recon, ]
gridded_anom_jjamean_NHET$lon <- gridded_anom_jjamean_NHET$lon[lonsel_recon]
gridded_anom_jjamean_NHET$lat <- gridded_anom_jjamean_NHET$lat[latsel_recon]

reconstruction.target <- gridded_anom_jjamean_NHET   # what we want to reconstruct
proxydata <- ntrdata                                # NTREND predictors (unscreened)
```

##### 2. Reconstruction parameters

These settings are passed to `Master_CFR.R`. If unchanged, a **PCR field reconstruction** for **750–2011** is run with calibration **1941–2000**.

```{r parameters}
recon.name  <- "Test_reconstruction"
do.cps      <- FALSE    # PCR (not Composite Plus Scale)
do.field    <- TRUE     # reconstruct every grid cell
do.index    <- TRUE     # also reconstruct the spatial mean
calib.start <- 1941
calib.end   <- 2000
startyear   <- 750
endyear     <- 2011
```

##### 3. Run the course reconstruction script

`Master_CFR.R` reads the settings above and calls the full workflow in `Recon_files/` (proxy nests, PC truncation, variance adjustment, NetCDF output).

**Required packages** (install once if missing):

```{r packages, eval=FALSE}
# install.packages(c("foreach", "doParallel", "mvnfast", "ncdf4"))
```

```{r run reconstruction}
source("Master_CFR.R")
cat("Reconstruction finished.\n")
cat("Output folder:", out.suffix, "\n")
```

The script creates a folder `out.suffix` (e.g. `Test_reconstruction_2026_06_11_162627`) containing:

- `output_<suffix>.nc` — gridded reconstruction (NetCDF)
- `Recon_spatmean_ROSM_<suffix>.txt` — spatial-mean time series
- `Log_<suffix>.txt` — parameter log

##### 4. Load results

```{r load reconstruction results}
library(ncdf4)
loaded <- load_cfr_output(out.suffix, startyear = startyear)
recon          <- loaded$recon
spatmean_recon <- loaded$spatmean_recon

cat("Field dimensions:", paste(dim(recon$data), collapse = " × "), "(lon × lat × time)\n")
```

##### 5. Maps — coldest and warmest year

```{r helpers}
# world_coastline.dat has 2 columns (lon, lat); NA rows mark segment breaks
world <- load_world_coastline("../Data/world_coastline.dat")

field_to_df <- function(field_slice) {
  df <- expand.grid(lon = recon$lon, lat = recon$lat)
  df$value <- as.vector(field_slice)
  df
}

plot_field_map <- function(field_slice, title, subtitle = NULL, limits = c(-3, 3)) {
  ggplot() +
    geom_tile(data = field_to_df(field_slice), aes(x = lon, y = lat, fill = value)) +
    geom_path(data = world, aes(x = lon, y = lat, group = seg), inherit.aes = FALSE,
              colour = "grey35", linewidth = 0.2) +
    scale_fill_distiller(
      palette = "RdBu", name = "Anomaly (°C)",
      limits = limits, oob = scales::squish
    ) +
    coord_fixed(ratio = 1.35, xlim = c(lon_recon_W, lon_recon_E), ylim = c(lat_recon_S, lat_recon_N)) +
    labs(title = title, subtitle = subtitle, x = "Longitude", y = "Latitude") +
    theme_clim()
}
```

```{r coldest warmest maps}
recon_spatial_average <- ts(apply(recon$data, 3, mean, na.rm = TRUE), start = startyear)

coldest_idx  <- which.min(recon_spatial_average)
warmest_idx  <- which.max(recon_spatial_average)
coldest_year <- time(recon_spatial_average)[coldest_idx]
warmest_year <- time(recon_spatial_average)[warmest_idx]

plot_field_map(
  recon$data[, , coldest_idx],
  title = "Coldest year in reconstruction",
  subtitle = paste("Year", coldest_year, "— spatial mean anomaly:", round(recon_spatial_average[coldest_idx], 2), "°C")
)

plot_field_map(
  recon$data[, , warmest_idx],
  title = "Warmest year in reconstruction",
  subtitle = paste("Year", warmest_year, "— spatial mean anomaly:", round(recon_spatial_average[warmest_idx], 2), "°C")
)
```

##### 6. Compare reconstruction to CRUTEM (instrumental target)

Only possible for years covered by CRUTEM (1901–2015).

```{r compare warmest to target}
if (warmest_year %in% reconstruction.target$time) {
  tgt_slice <- reconstruction.target$data[, , which(reconstruction.target$time == warmest_year)]
  plot_field_map(
    tgt_slice,
    title = paste("CRUTEM target —", warmest_year),
    subtitle = "Instrumental JJA anomaly (same year as warmest recon year)"
  )
}
```

##### 7. Spatial mean time series

Three lines: CRUTEM (where available), mean of the **field** reconstruction, and the **direct index** reconstruction from the workflow.

```{r spatial mean ts}
reconstruction.target.spatial.average <- ts(
  apply(reconstruction.target$data, 3, mean, na.rm = TRUE),
  start = reconstruction.target$time[1]
)

ts_long <- rbind(
  data.frame(
    year = as.numeric(time(recon_spatial_average)),
    value = as.numeric(recon_spatial_average),
    series = "Mean of field recon"
  ),
  data.frame(
    year = as.numeric(time(spatmean_recon)),
    value = as.numeric(spatmean_recon),
    series = "Direct index recon"
  ),
  data.frame(
    year = as.numeric(time(reconstruction.target.spatial.average)),
    value = as.numeric(reconstruction.target.spatial.average),
    series = "CRUTEM target"
  )
)

ggplot(subset(ts_long, series != "CRUTEM target" | year >= 1900), aes(x = year, y = value, colour = series)) +
  geom_hline(yintercept = 0, colour = "grey80", linewidth = 0.3) +
  geom_line(linewidth = 0.55) +
  scale_colour_manual(values = c(
    "CRUTEM target" = col_target,
    "Mean of field recon" = col_recon,
    "Direct index recon" = col_index
  )) +
  labs(
    title = "Northern Hemisphere extratropics — JJA temperature anomaly",
    subtitle = "750–2011 reconstruction | calibration 1941–2000 | NTREND proxies",
    x = "Year", y = "Anomaly (°C)", colour = NULL
  ) +
  theme_clim()
```

The field mean and index reconstruction should agree closely — large differences would point to a workflow inconsistency.

##### 8. Save for Exercise 15

```{r save stuff}
save(
  gridded_anom_jjamean_NHET, recon, recon_spatial_average, spatmean_recon,
  reconstruction.target, reconstruction.target.spatial.average, proxydata,
  out.suffix, startyear, endyear, calib.start, calib.end,
  lon_recon_W, lon_recon_E, lat_recon_S, lat_recon_N,
  file = "reconstruction_input_and_output.RData"
)
```
