---
title: "Investigate real proxy data"
output: html_notebook
---
### Exercise 11: Proxy characteristics

Empty memory, set working directory and load functions, libraries and data from previous exercises

```{r}
rm(list=ls())
load(file = "../data/workspace_ex10.RData")   
```

##### Plot a few time series of PAGES proxies and corresponding gridded instr. data (CRU). We use grid data because only very rarely there will be a station nearby at the same elevation, which better represents the local climatic conditions.

```{r}
# load gridded annual or JJA mean data from day 1
# read "cru_abs_jjamean", etc. list with lon/lat info
load(paste0(workdir,"/../data/crutem_abs_jjamean_1901-2015.Rdata"))
load(paste0(workdir,"/../data/crutem_abs_annmean_1901-2015.Rdata"))
load(paste0(workdir,"/../data/crutem_anom_jjamean_1901-2015.Rdata")) 
load(paste0(workdir,"/../data/crutem_anom_annmean_1901-2015.Rdata"))

# Rename cru to gridded for clarity and
gridded_abs_annmean=cru_abs_annmean
gridded_abs_jjamean=cru_abs_jjamean
gridded_anom_annmean=cru_anom_annmean
gridded_anom_jjamean=cru_anom_jjamean

#  getgridboxnum() requires list with $lon and $lat
p2kgridnum <- getgridboxnum(station=p2kdata,grid=gridded_abs_jjamean) 
str(p2kgridnum)
ntrgridnum <- getgridboxnum(station=ntrdata,grid=gridded_abs_jjamean) 
ntrallcor <- vector()

# use code from day 1 to plot a few time series: 
# PAGES2k vs gridded instr. ann. mean temp.
# Hint: try linetype "b" instead of "l". Why is this necessary?
par(mfrow=c(3,3),mar=c(2,2,2,0.5))
for (stat in 1:9){
  plot(p2kdata$time,p2kdata$data[,stat],ty="b",cex=0.1,
       main=p2kdata$type[stat],
       ylim=c(min(c(p2kdata$data[,stat],
              gridded_abs_annmean$data[p2kgridnum$xgridnum[stat],
              p2kgridnum$ygridnum[stat],]),na.rm=T),
              max(c(p2kdata$data[,stat],
              gridded_abs_annmean$data[p2kgridnum$xgridnum[stat],
              p2kgridnum$ygridnum[stat],]),na.rm=T)))
  lines(gridded_abs_annmean$time,gridded_abs_annmean$data[p2kgridnum$xgridnum[stat],
        p2kgridnum$ygridnum[stat],],col="red")
  if (stat==1) {
    legend("topleft",legend=c("station","gridbox"),lty=c(1,1),
           col=c("black","red"))
  }
}
```
Discuss the characteristics of the proy data and their agreement with instrumental observations, i.e. climate signal in proxy data.






###### Calculate correlation coefficients between proxies and instr. data. Plot them as a histogram. Compare to results with instr. data from the first day.

```{r}
p2kallcor_ann <- rep(NA,length(p2kdata$lon))
p2kallcor_ann_pval <- rep(NA,length(p2kdata$lon))
p2kallcor_ann_df <- rep(NA,length(p2kdata$lon))
for (prox in 1:(length(p2kdata$lon))) {
  # correlation 
  # adjust time axis
  postimecru <- which(gridded_abs_annmean$time>1900 & gridded_abs_annmean$time<2013)
  postimeprox <- which(p2kdata$time>1900 & p2kdata$time<2013)
  if (length(which(!is.na(gridded_abs_annmean$data[p2kgridnum$xgridnum[prox],
              p2kgridnum$ygridnum[prox],postimecru])))>5 & 
       length(which(!is.na(p2kdata$data[postimeprox,prox])))>5 ) {
    p2kallcor_ann[prox] <- cor(p2kdata$data[postimeprox,prox],
              gridded_abs_annmean$data[p2kgridnum$xgridnum[prox],
              p2kgridnum$ygridnum[prox],postimecru],
              use="pairwise.complete.obs")
    p2kallcor_ann_pval[prox] <- try(cor.test(p2kdata$data[postimeprox,prox],
              gridded_abs_annmean$data[p2kgridnum$xgridnum[prox],
              p2kgridnum$ygridnum[prox],postimecru],
              use="pairwise.complete.obs")[3])
    p2kallcor_ann_df[prox] <- try(cor.test(p2kdata$data[postimeprox,prox],
              gridded_abs_annmean$data[p2kgridnum$xgridnum[prox],
              p2kgridnum$ygridnum[prox],postimecru],
              use="pairwise.complete.obs")[2])
  }
}
#plot histogram of correlation coefficients
hist(p2kallcor_ann,main="P2k vs CRU ann. mean correlation")
```
Discuss the correlation coefficients between instrumental data and proxies in the PAGES2k data base, which were specifically selected to be temperature sensitive. Compare to results with instrumental station data results from the first day.
 
     
     
         

###### How much better is it if we look at JJA temperature? How much better are NTREND proxies, which are selected to be the "best" temperature recorders

```{r}
p2kallcor_jja <- rep(NA,length(p2kdata$lon))
p2kallcor_jja_pval <- rep(NA,length(p2kdata$lon))
p2kallcor_jja_df <- rep(NA,length(p2kdata$lon))
for (prox in 1:(length(p2kdata$lon))) {
  
# TASK: Repeat correlation calculation of previous exercise for summer (JJA) instead of annual mean
?
    
# TASK: Now repeat the analysis for NTREND in the summer season
?
    
```

TASK: Interpret the differences between annual mean and summer season and the differences between PAGES2k and NTREND proxy collection. And how do they differ from yesterday's station data results?






#####Plot maps with proxy locations and colorscale for correlation coefficient to see if there are spatial patterns

```{r}
par(mar=c(2,2,2,5))
lev <- seq(-0.9,0.9,0.2)
br <- length(lev)-1
colpal <- two.colors(n=br,start="darkblue", end="darkred", alpha=1.0)
datcol <- colpal[as.numeric(cut(p2kallcor_jja,breaks=lev))]
plot(p2kdata$lon, p2kdata$lat, # lon (x) and lat (y)
       cex=1.1,       # cex=point size~precip
       pch=16,        # point type
       col="grey",   #datcol,    # point fill color 
       xlab="Longitude", ylab="Latitude",
       main="PAGES2k/CRUTEM correlation") 
points(p2kdata$lon, p2kdata$lat, # lon (x) and lat (y)
       cex=1.0,       # cex=point size~precip
       pch=16,        # point type
       col=datcol)     #datcol,    # point fill color
image.plot(1:length(datcol),1:1,
           matrix(rep(lev,length.out=length(datcol)),nrow=1),
           col=colpal,legend.only=T)
maps::map("world",interior=F,add=T,ty='l',col='black')

# TASK: add the NTREND map
?
```

TASK: Interpret the spatial patterns in the correlation maps!






##### Separate the proxy archives and compare tree-ring proxies to speleothems, marine and lake sediments, corals, etc. 

```{r}
lev = length(unique(unlist(p2kdata$type)))
datcol <- rainbow(lev)[as.numeric(as.factor(unlist(p2kdata$type)))]
plot(p2kdata$lon, p2kdata$lat, # lon (x) and lat (y)
       cex=1.1,       # cex=point size~precip
       pch=16,        # point type
       col="grey",   #datcol,    # point fill color 
       xlab="Longitude", ylab="Latitude",
       main="PAGES2k archive distribution") 
points(p2kdata$lon, p2kdata$lat, # lon (x) and lat (y)
       cex=1.0,       # cex=point size~precip
       pch=16,        # point type
       col=datcol)     #datcol,    # point fill color
maps::map("world",interior=F,add=T,ty='l',col='black')
```

Does the knowledge of the archive type help you with the explanantion of the correlations found previously? What are potentials and limitations of this distribution of archive types?








##### Try to identify additional problems of real proxy data, e.g. temporal resolution and how does that effect correlation, low vs high frequency variability, thresholds, etc. 
	



	
Save entire workspace to continue with it in the following exercise or if you close R or if R crashes
```{r}
save.image(file = "../data/workspace_ex11.RData")   
```
