---
title: "Proxy network CFR comparison"
subtitle: "DoD2k speleothems vs tree-ring baselines"
author: "Climate reconstruction project"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 120
params:
  quick_run: true       # TRUE: 3 fast experiments; FALSE: full batch
  overwrite_cfr: false    # TRUE: re-run CFR even if cached
execute:
  warning: false
  message: false
---

Work in this file like a notebook: run chunks interactively, then **Render** when you want the HTML report.

```{r}
#| label: setup

knitr::opts_chunk$set(
  echo = TRUE,
  message = FALSE,
  warning = FALSE
)

if (!file.exists("config/experiments.R")) {
  stop("Open this file from proxy-cfr-comparison/ (project root).")
}

source("R/load_project.R")
load_project()
library(ggplot2)

quick_run <- params$quick_run
overwrite_cfr <- params$overwrite_cfr
```

# Introduction

Compare **climate field reconstructions (CFR)** across proxy networks. Focus: **DoD2k v2.0 speleothems** vs tree-ring baselines.

| Dimension | What we vary |
|-----------|----------------|
| **Proxy network** | DoD2k speleothem δ18O, DoD2k trees, combined; N-TREND baseline |
| **Calibration** | Years for PCR in `course/exercises/Master_CFR.R` |
| **Validation** | Independent years vs CRUTEM |

Speleothems in DoD2k: filter **`archiveType = Speleothem`** and **`paleoData_proxy = d18O`** ([DoD2k docs](https://lluecke.github.io/dod2k/)).

# 1. Proxy networks

```{r}
#| label: preview-networks

calib <- CALIBRATION_SCHEMES$course_split$calibration

cat("DoD2k archive × proxy (v2.0):\n")
print(head(summarize_dod2k_archives("v2"), 12))

network_ids <- c("dod2k_speleothems", "dod2k_trees", "dod2k_speleo_trees", "ntrend_trees")
previews <- lapply(network_ids, function(id) {
  net <- build_proxy_network(
    id,
    calib_start = calib[1],
    calib_end = calib[2]
  )
  cbind(proxy_network_map_df(net), label = net$meta$label)
})
proxy_map_df <- do.call(rbind, previews)

knitr::kable(
  unique(proxy_map_df[, c("network", "label")]),
  caption = "Networks after calibration coverage filtering"
)
```

```{r}
#| label: plot-proxy-locations
#| fig-cap: "Proxy locations (course_split calibration filter)"

plot_proxy_map(proxy_map_df, title = "Proxy networks")
```

# 2. Run reconstructions

Each experiment calls `course/exercises/Master_CFR.R` and caches under `output/cfr/`.

```{r}
#| label: experiment-plan

knitr::kable(
  EXPERIMENT_PLAN[EXPERIMENT_PLAN$run, ],
  caption = "Default experiment matrix"
)
```

```{r}
#| label: select-experiments

if (quick_run) {
  selected <- data.frame(
    network_id = c("dod2k_speleothems", "dod2k_speleo_trees", "ntrend_trees"),
    calibration_id = rep("course_split", 3L),
    stringsAsFactors = FALSE
  )
} else {
  selected <- EXPERIMENT_PLAN[EXPERIMENT_PLAN$run, c("network_id", "calibration_id")]
}

selected$experiment_id <- paste(selected$network_id, selected$calibration_id, sep = "__")
knitr::kable(selected, caption = "Experiments to run")
```

```{r}
#| label: run-cfr
#| cache: true

results <- setNames(vector("list", nrow(selected)), selected$experiment_id)

for (i in seq_len(nrow(selected))) {
  id <- selected$experiment_id[i]
  message("=== ", id, " ===")
  results[[id]] <- run_cfr_experiment(
    network_id = selected$network_id[i],
    calibration_id = selected$calibration_id[i],
    overwrite = overwrite_cfr
  )
}

ok <- vapply(results, function(x) is.null(x$error), logical(1))
if (!any(ok)) stop("All CFR experiments failed.")
if (!all(ok)) {
  warning("Some experiments failed: ", paste(names(results)[!ok], collapse = ", "))
}
```

# 3. Validation

Per grid cell (Exercise 6 style): correlation **r**, **RMSE**, **RMSESS** = `1 - RMSE / RMSE_clim`.

```{r}
#| label: validate-all

skills <- lapply(results[ok], function(res) {
  validate_cfr_result(res, calibration_id = res$calibration_id)
})

knitr::kable(
  print_skill_summary(skills),
  caption = "Skill summary (median grid + index validation)"
)
```

```{r}
#| label: plot-maps
#| fig-height: 6
#| fig-cap: "Validation correlation r"

for (s in skills) {
  print(
    plot_skill_map(
      s$grid_validation$r,
      lon = results[[s$experiment_id]]$reconstruction$lon,
      lat = results[[s$experiment_id]]$reconstruction$lat,
      title = paste("Validation r —", s$experiment_id),
      limits = c(-1, 1)
    )
  )
}
```

```{r}
#| label: multi-calibration

if (!quick_run) {
  schemes <- names(CALIBRATION_SCHEMES)
  speleo_results <- lapply(schemes, function(sc) {
    run_cfr_experiment("speleothem_ntrend", sc, overwrite = FALSE)
  })
  names(speleo_results) <- schemes
  cmp <- skill_comparison_table(speleo_results, calibration_id = schemes)
  knitr::kable(round(cmp, 3))
  print(plot_skill_comparison(cmp, "index_val_r"))
} else {
  cat("Set params$quick_run = FALSE to compare all calibration schemes.\n")
}
```

# 4. Discussion

**Proxy networks:** DoD2k v2.0 has 224 speleothem δ18O records globally (~41 after our calibration filter). Tree networks give dense NH coverage; combined networks change the PCA basis.

**Calibration / validation:** Later calibration emphasises recent warming; early validation tests low-frequency variability.

---

```{r}
#| label: session

sessionInfo()
```
