---
title: "DoD2k speleothem data overview"
subtitle: "Load, explore, and visualise the archive"
author:
  - "Madina Vogt"
  - "Florian Christ"
  - "Lukas Batschelet"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 120
execute:
  warning: false
  message: false
---

# Introduction

Explore **DoD2k v2.0** speleothem records from [DoD2k](https://lluecke.github.io/dod2k/).

DoD2k distinguishes **archive** (`archiveType`, e.g. Speleothem) from **measurement** (`paleoData_proxy`: δ18O, δ13C, growth rate, Mg/Ca). This notebook loads the compact CSV bundle, builds a long-format catalog, and plots an overview of speleothem sites.

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(glue)
library(plotly)
library(terra)
```

# Load DoD2k data

Data live in `course/data/dod2k_v2.0/`. The compact format stores one row per dataset; year and proxy values are comma-separated arrays in separate files.

**Flow:** compact CSV → wide table (`load_dod2k_compact`) → long catalog (`load_dod2k_catalog`) → RDS cache for fast reload.

## Paths and cache

```{r}
#| label: paths
#| code-fold: true

#' Find repository root containing `course/data/`.
#'
#' @return Absolute path to the repository root.
find_repo_root <- function() {
  candidates <- c(
    if (dir.exists("course/data")) normalizePath(".", mustWork = TRUE),
    if (dir.exists("../course/data")) normalizePath("..", mustWork = TRUE),
    if (dir.exists("../../course/data")) normalizePath("../..", mustWork = TRUE)
  )
  root <- candidates[!sapply(candidates, is.null)][1]
  if (is.na(root)) {
    stop("Cannot find course/data — open this file from proxy-cfr-comparison/")
  }
  root
}

repo_root <- find_repo_root()
proj_root <- if (file.exists("analysis.qmd")) "." else "proxy-cfr-comparison"
data_dir  <- file.path(repo_root, "course", "data", "dod2k_v2.0")
cache_dir <- file.path(proj_root, "output", "cache")
cache_rds <- file.path(cache_dir, "dod2k_timeseries.rds")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

year_min <- 0L   # reconstruction window (CE)
year_max <- 2000L
```

## Compact CSV loader

Three files are joined: metadata, year arrays, and proxy-value arrays.

```{r}
#| label: dod2k-loader
#| code-fold: true

#' Parse one line from a DoD2k compact array CSV (`id,val1,val2,…`).
#'
#' @param line One CSV line (without header).
#' @return List with `id` and numeric `values`.
parse_array_line <- function(line) {
  parts <- strsplit(line, ",", fixed = TRUE)[[1]]
  list(id = parts[[1]], values = as.numeric(parts[-1]))
}

#' Read a compact array column into a tibble.
#'
#' @param path Path to e.g. `dod2k_v2.0_compact_year.csv`.
#' @return Tibble: `dataset_id`, `values` (list of numeric vectors).
read_array_column <- function(path) {
  lines <- readLines(path, warn = FALSE)
  parsed <- lapply(lines[-1], parse_array_line)
  tibble(
    dataset_id = vapply(parsed, `[[`, character(1), "id"),
    values = lapply(parsed, `[[`, "values")
  )
}

#' Load DoD2k v2.0 compact bundle — one row per dataset, list-columns for time series.
#'
#' @return Wide tibble with metadata, `year`, and `paleoData_values`.
load_dod2k_compact <- function() {
  prefix <- basename(data_dir)
  meta <- read_csv(
    file.path(data_dir, paste0(prefix, "_compact_metadata.csv")),
    show_col_types = FALSE
  ) |>
    rename(dataset_id = any_of(c("datasetId", "...1")))

  years <- read_array_column(file.path(data_dir, paste0(prefix, "_compact_year.csv"))) |>
    rename(year = values)
  values <- read_array_column(
    file.path(data_dir, paste0(prefix, "_compact_paleoData_values.csv"))
  ) |>
    rename(paleoData_values = values)

  meta |>
    left_join(years, by = "dataset_id") |>
    left_join(values, by = "dataset_id")
}
```

## Long-format catalog

Expands list-columns to one row per `(dataset, year)` and caches the result as RDS.

```{r}
#| label: dod2k-catalog
#| code-fold: true

#' Convert compact data to a long catalog (0–2000 CE, finite values only).
#'
#' @param compact Output of [load_dod2k_compact()].
#' @return Tibble: `dataset_id`, `year`, `value`, `lon`, `lat`, `archive_type`, `proxy_type`.
compact_to_catalog <- function(compact) {
  rows <- vector("list", nrow(compact))
  n <- 0L
  for (i in seq_len(nrow(compact))) {
    rec <- compact[i, ]
    yrs  <- rec$year[[1]]
    vals <- rec$paleoData_values[[1]]
    if (!length(yrs) || !length(vals)) next

    ok <- yrs >= year_min & yrs <= year_max & is.finite(vals)
    if (!any(ok)) next

    n <- n + 1L
    rows[[n]] <- tibble(
      dataset_id = rec$dataset_id,
      year = as.integer(yrs[ok]),
      value = as.double(vals[ok]),
      lon = rec$geo_meanLon,
      lat = rec$geo_meanLat,
      archive_type = rec$archiveType,
      proxy_type = rec$paleoData_proxy
    )
  }
  if (!n) return(tibble())
  bind_rows(rows)
}

#' Read cached catalog from `output/cache/dod2k_timeseries.rds`.
#'
#' @return Tibble or `NULL` if cache does not exist.
read_catalog_cache <- function() {
  if (!file.exists(cache_rds)) return(NULL)
  readRDS(cache_rds) |>
    mutate(
      dataset_id = as.character(dataset_id),
      archive_type = as.character(archive_type),
      proxy_type = as.character(proxy_type),
      year = as.integer(year),
      value = as.double(value),
      lon = as.double(lon),
      lat = as.double(lat)
    )
}

#' Load or build the long-format catalog.
#'
#' @param refresh If `TRUE`, rebuild cache from CSV.
#' @return Long-format catalog tibble.
load_dod2k_catalog <- function(refresh = FALSE) {
  if (!refresh) {
    cached <- read_catalog_cache()
    if (!is.null(cached)) return(cached)
  }
  message("Building catalog from compact CSV …")
  catalog <- compact_to_catalog(load_dod2k_compact())
  saveRDS(catalog, cache_rds)
  catalog
}
```

## Archive summary

First run builds the cache; later runs read RDS directly.

```{r}
#| label: dod2k-load
#| cache: true

dod2k_compact <- load_dod2k_compact()
catalog       <- load_dod2k_catalog()

archive_summary <- dod2k_compact |>
  count(archiveType, paleoData_proxy, name = "n", sort = TRUE) |>
  rename(archive_type = archiveType, proxy_type = paleoData_proxy)

glue(
  "Records: {format(nrow(dod2k_compact), big.mark = ',')}  |  ",
  "Speleothems: {sum(dod2k_compact$archiveType == 'Speleothem', na.rm = TRUE)}  |  ",
  "δ18O: {sum(dod2k_compact$archiveType == 'Speleothem' & dod2k_compact$paleoData_proxy == 'd18O', na.rm = TRUE)}"
)

knitr::kable(head(archive_summary, 12), caption = "DoD2k archive × proxy (metadata)")
```

# Speleothem overview

Filter `archiveType == "Speleothem"`. Four proxy types appear in DoD2k v2.0: δ18O, δ13C, growth rate, Mg/Ca.

## Styling

Shared colours, labels, and a minimal ggplot theme for the figures below.

```{r}
#| label: speleothem-config
#| code-fold: true

proxy_order   <- c("d18O", "d13C", "growth rate", "Mg/Ca")
proxy_labels  <- c(d18O = "δ18O", d13C = "δ13C", `growth rate` = "growth rate", `Mg/Ca` = "Mg/Ca")
proxy_colors  <- c(d18O = "#b2182b", d13C = "#d6604d", `growth rate` = "#f4a582", `Mg/Ca` = "#92c5de")
proxy_symbols <- c(d18O = "triangle-up", d13C = "triangle-down", `growth rate` = "square", `Mg/Ca` = "circle")

theme_clean <- function() {
  theme_minimal() + theme(legend.position = "bottom")
}
```

## Site tables

**`n_metadata`** = rows in the archive; **`n_with_data`** = series with values in 0–2000 CE.

```{r}
#| label: speleothem-helpers
#| code-fold: true

#' One row per speleothem series with site metadata.
#'
#' @param catalog Long catalog from [load_dod2k_catalog()].
#' @param compact Wide compact table (for site names).
#' @return Tibble with `dataset_id`, coords, year range, `n_years`, `site`.
speleothem_sites <- function(catalog, compact) {
  sites <- catalog |>
    filter(archive_type == "Speleothem") |>
    group_by(dataset_id, proxy_type, lon, lat) |>
    summarize(
      year_start = min(year),
      year_end = max(year),
      n_years = n(),
      .groups = "drop"
    )

  meta <- compact |>
    filter(archiveType == "Speleothem") |>
    dplyr::select(dataset_id, geo_siteName) |>
    rename(site = geo_siteName)

  sites |>
    left_join(meta, by = "dataset_id") |>
    mutate(
      site = if_else(is.na(site) | site == "" | site == dataset_id, dataset_id, site),
      proxy_type = factor(proxy_type, levels = proxy_order)
    )
}

#' Count metadata rows vs. series with usable time series, by proxy type.
#'
#' @param compact Wide compact table.
#' @param catalog Long catalog.
#' @return Tibble: `proxy_label`, `proxy_type`, `n_metadata`, `n_with_data`.
speleothem_summary <- function(compact, catalog) {
  meta <- compact |>
    filter(archiveType == "Speleothem") |>
    count(paleoData_proxy, name = "n_metadata") |>
    rename(proxy_type = paleoData_proxy)

  with_data <- catalog |>
    filter(archive_type == "Speleothem") |>
    distinct(dataset_id, proxy_type) |>
    count(proxy_type, name = "n_with_data")

  meta |>
    full_join(with_data, by = "proxy_type") |>
    mutate(
      n_metadata = replace_na(n_metadata, 0L),
      n_with_data = replace_na(n_with_data, 0L),
      proxy_label = proxy_labels[proxy_type],
      proxy_type = factor(proxy_type, levels = proxy_order)
    ) |>
    arrange(proxy_type) |>
    transmute(proxy_label, proxy_type, n_metadata, n_with_data)
}
```

```{r}
#| label: speleothem-summary
#| cache: true
#| dependson: dod2k-load

speleo_sites <- speleothem_sites(catalog, dod2k_compact)

speleothem_summary(dod2k_compact, catalog) |>
  knitr::kable(caption = "Speleothem proxy types")
```

## Site map

Interactive world map — colour and symbol by proxy type; hover shows site name and year range.

```{r}
#| label: speleothem-map-fn
#| code-fold: true

#' Interactive plotly map of speleothem sites.
#'
#' @param sites Output of [speleothem_sites()].
#' @return plotly object.
plot_speleothem_map <- function(sites) {
  sites <- sites |>
    mutate(
      proxy_label = proxy_labels[as.character(proxy_type)],
      hover = paste0(
        "<b>", site, "</b><br>", dataset_id, "<br>",
        "Proxy: ", proxy_label, "<br>",
        "Years: ", year_start, "–", year_end, " (n = ", n_years, ")"
      )
    )

  p <- plot_ly(type = "scattergeo", mode = "markers")
  for (px in proxy_order) {
    sub <- filter(sites, as.character(proxy_type) == px)
    if (!nrow(sub)) next
    p <- add_trace(
      p, data = sub, lon = ~lon, lat = ~lat, text = ~hover, hoverinfo = "text",
      name = paste0("Speleothem: ", proxy_labels[[px]], " (n=", nrow(sub), ")"),
      marker = list(
        symbol = proxy_symbols[[px]], size = 9,
        color = proxy_colors[[px]], line = list(width = 0.6, color = "#333")
      )
    )
  }
  p |> layout(
    title = list(
      text = paste0("<b>DoD2k speleothem archive</b><br><sup>", nrow(sites), " series</sup>"),
      x = 0.5, xanchor = "center"
    ),
    geo = list(
      projection = list(type = "natural earth"),
      showland = TRUE, landcolor = "#f0f0f0",
      showocean = TRUE, oceancolor = "#e6f2fa",
      showcountries = TRUE, countrycolor = "#bbbbbb",
      lataxis = list(range = c(-40, 75))
    ),
    legend = list(orientation = "h", y = -0.05, x = 0.5, xanchor = "center")
  )
}
```

```{r}
#| label: speleothem-map
#| dependson: speleothem-summary

plot_speleothem_map(speleo_sites)
```

## Temporal coverage

How many series have at least one dated value per 100-year bin (actual years, not calendar span).

```{r}
#| label: speleothem-density-fn
#| code-fold: true

#' Stacked bar chart — series count per time bin.
#'
#' @param catalog Long-format catalog.
#' @param bin_width Bin width in years (default 100).
#' @return ggplot object.
plot_speleothem_density <- function(catalog, bin_width = 100) {
  density <- catalog |>
    filter(archive_type == "Speleothem") |>
    mutate(bin_start = (as.integer(year %/% bin_width)) * bin_width) |>
    distinct(dataset_id, proxy_type, bin_start) |>
    count(bin_start, proxy_type, name = "n_series") |>
    mutate(
      proxy_type = factor(proxy_type, levels = proxy_order),
      bin_mid = bin_start + bin_width / 2
    )

  ggplot(density, aes(bin_mid, n_series, fill = proxy_type)) +
    geom_col(width = bin_width * 0.9, colour = "white", linewidth = 0.2) +
    scale_fill_manual(values = proxy_colors, labels = proxy_labels[proxy_order], name = "Proxy type") +
    scale_x_continuous(limits = c(0, 2000), breaks = seq(0, 2000, 250), name = "Year (CE)") +
    labs(
      title = "Where observations exist",
      subtitle = "Stacked by proxy type (100-year bins)",
      y = "Series with data"
    ) +
    theme_clean()
}
```

```{r}
#| label: speleothem-density
#| fig-cap: "Series with observations per century"
#| dependson: speleothem-summary

plot_speleothem_density(catalog)
```

## Temporal resolution

Number of dated values per series — more informative than calendar first/last year.

```{r}
#| label: speleothem-resolution-fn
#| code-fold: true

#' Boxplot of observation counts per series.
#'
#' @param sites Output of [speleothem_sites()].
#' @return ggplot object.
plot_speleothem_resolution <- function(sites) {
  ggplot(sites, aes(proxy_type, n_years, fill = proxy_type)) +
    geom_boxplot(outlier.size = 0.8, alpha = 0.85) +
    scale_fill_manual(values = proxy_colors, guide = "none") +
    scale_x_discrete(labels = proxy_labels[proxy_order]) +
    labs(
      title = "Temporal resolution per series",
      subtitle = "Number of dated values — calendar span is often misleading",
      x = NULL, y = "Observations per series"
    ) +
    theme_clean()
}
```

```{r}
#| label: speleothem-resolution
#| fig-cap: "Dated values per series"
#| dependson: speleothem-summary

plot_speleothem_resolution(speleo_sites)
```

# Instrumental correlation screening

Do speleothem proxies covary with local climate? We extract **CRU TS v4.07** temperature and precipitation from the nearest 0.5° grid cell (same idea as Exercise 11/12 with CRUTEM) and correlate each series with:

- annual mean temperature
- JJA mean temperature
- annual precipitation total

Overlap window: **1901–2000 CE**, minimum **10** paired years. This is a first-order screening — speleothem signals depend on region, season, and proxy type (δ18O can track rain amount or temperature).

First run downloads CRU TS (~400 MB compressed) to `output/cache/cru_ts/` and caches extracted site climate.

## CRU grid extraction

```{r}
#| label: cru-setup
#| code-fold: true

cru_dir  <- file.path(cache_dir, "cru_ts")
cru_base <- "https://crudata.uea.ac.uk/cru/data/hrg/cru_ts_4.07/cruts.2304141047.v4.07"
cru_years <- 1901:2022
cru_rds   <- file.path(cache_dir, "cru_site_climate.rds")
dir.create(cru_dir, recursive = TRUE, showWarnings = FALSE)

#' Download and decompress one CRU TS variable if needed.
#'
#' @param var `"tmp"` or `"pre"`.
#' @return Path to the uncompressed NetCDF file.
ensure_cru_nc <- function(var) {
  fname <- paste0("cru_ts4.07.1901.2022.", var, ".dat.nc")
  nc_path <- file.path(cru_dir, fname)
  gz_path <- paste0(nc_path, ".gz")
  if (!file.exists(nc_path)) {
    if (!file.exists(gz_path)) {
      message("Downloading CRU TS ", var, " …")
      url <- paste(cru_base, var, paste0(fname, ".gz"), sep = "/")
      if (nzchar(Sys.which("curl"))) {
        status <- system2("curl", c("-L", "-C", "-", "--retry", "3", "--max-time", "900", "-o", gz_path, url))
        if (status != 0) stop("CRU download failed for ", var)
      } else {
        utils::download.file(url, gz_path, mode = "wb", timeout = 900)
      }
    }
    message("Decompressing ", basename(gz_path), " …")
    status <- system2("gunzip", c("-k", "-f", gz_path))
    if (status != 0) stop("gunzip failed — decompress ", gz_path, " manually")
  }
  nc_path
}

#' Aggregate monthly CRU matrices to annual temperature means and precipitation sums.
cru_monthly_to_annual <- function(mat_tmp, mat_pre, years = cru_years) {
  n_pt <- nrow(mat_tmp)
  n_yr <- length(years)
  tmp_ann <- tmp_jja <- pre_ann <- matrix(NA_real_, n_pt, n_yr)
  for (j in seq_len(n_yr)) {
    m <- (j - 1L) * 12L + 1:12
    tmp_ann[, j] <- rowMeans(mat_tmp[, m, drop = FALSE], na.rm = TRUE)
    tmp_jja[, j] <- rowMeans(mat_tmp[, m[6:8], drop = FALSE], na.rm = TRUE)
    pre_ann[, j] <- rowSums(mat_pre[, m, drop = FALSE], na.rm = TRUE)
  }
  list(tmp_ann = tmp_ann, tmp_jja = tmp_jja, pre_ann = pre_ann)
}

#' Extract CRU TS climate at speleothem locations (cached as RDS).
#'
#' @param loc Tibble with `loc_id`, `lon`, `lat`.
#' @param refresh Rebuild even when RDS exists.
#' @return Long tibble: `loc_id`, `lon`, `lat`, `year`, `tmp_ann`, `tmp_jja`, `pre_ann`.
load_cru_site_climate <- function(loc, refresh = FALSE) {
  if (!refresh && file.exists(cru_rds)) return(readRDS(cru_rds))

  pts <- vect(loc, geom = c("lon", "lat"), crs = "EPSG:4326")
  tmp_nc <- ensure_cru_nc("tmp")
  pre_nc <- ensure_cru_nc("pre")

  mat_tmp <- as.matrix(
    terra::extract(rast(tmp_nc, subds = "tmp"), pts, method = "simple")[, -1]
  )
  mat_pre <- as.matrix(
    terra::extract(rast(pre_nc, subds = "pre"), pts, method = "simple")[, -1]
  )
  ann <- cru_monthly_to_annual(mat_tmp, mat_pre)

  clim <- map_dfr(seq_len(nrow(loc)), \(i) {
    tibble(
      loc_id = loc$loc_id[i], lon = loc$lon[i], lat = loc$lat[i],
      year = cru_years,
      tmp_ann = ann$tmp_ann[i, ], tmp_jja = ann$tmp_jja[i, ], pre_ann = ann$pre_ann[i, ]
    )
  })
  saveRDS(clim, cru_rds)
  clim
}
```

## Correlation functions

Pearson correlation with two-tailed p-value; series with fewer than 10 overlap years are skipped.

```{r}
#| label: corr-setup
#| code-fold: true

instr_y0 <- 1901L
instr_y1 <- 2000L
min_overlap <- 10L

climate_vars <- c(
  r_tmp_ann = "Temperature (annual)",
  r_tmp_jja = "Temperature (JJA)",
  r_pre_ann = "Precipitation (annual)"
)

#' Correlate one proxy series with one climate variable.
corr_series <- function(proxy_df, clim_df, clim_col,
                        y0 = instr_y0, y1 = instr_y1, min_n = min_overlap) {
  d <- proxy_df |>
    filter(year >= y0, year <= y1) |>
    inner_join(clim_df, by = "year")
  ok <- is.finite(d$value) & is.finite(d[[clim_col]])
  d <- d[ok, , drop = FALSE]
  if (nrow(d) < min_n) {
    return(tibble(r = NA_real_, p = NA_real_, n = nrow(d)))
  }
  ct <- cor.test(d$value, d[[clim_col]], method = "pearson")
  tibble(r = unname(ct$estimate), p = ct$p.value, n = nrow(d))
}

#' Build annual speleothem series (mean if multiple values per year).
speleothem_annual <- function(catalog) {
  catalog |>
    filter(archive_type == "Speleothem") |>
    group_by(dataset_id, proxy_type, lon, lat, year) |>
    summarize(value = mean(value), .groups = "drop")
}

#' Correlate all speleothem series with local CRU grid-box climate.
compute_speleo_correlations <- function(catalog, clim, sites,
                                        y0 = instr_y0, y1 = instr_y1) {
  loc <- sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
  proxy_ann <- speleothem_annual(catalog) |> inner_join(loc, by = c("lon", "lat"))

  res <- map_dfr(group_split(proxy_ann, dataset_id), function(df) {
    cdf <- clim |> filter(loc_id == df$loc_id[1])
    t1 <- corr_series(df, cdf, "tmp_ann", y0, y1)
    t2 <- corr_series(df, cdf, "tmp_jja", y0, y1)
    t3 <- corr_series(df, cdf, "pre_ann", y0, y1)
    tibble(
      dataset_id = df$dataset_id[1],
      proxy_type = df$proxy_type[1],
      lon = df$lon[1], lat = df$lat[1],
      n_overlap = t1$n,
      r_tmp_ann = t1$r, p_tmp_ann = t1$p,
      r_tmp_jja = t2$r, p_tmp_jja = t2$p,
      r_pre_ann = t3$r, p_pre_ann = t3$p
    )
  })

  res |>
    rowwise() |>
    mutate(
      best_var = {
        rs <- c(tmp_ann = r_tmp_ann, tmp_jja = r_tmp_jja, pre_ann = r_pre_ann)
        if (all(is.na(rs))) NA_character_ else names(rs)[which.max(abs(rs))]
      },
      best_r = {
        rs <- c(r_tmp_ann, r_tmp_jja, r_pre_ann)
        if (all(is.na(rs))) NA_real_ else rs[which.max(abs(rs))]
      }
    ) |>
    ungroup()
}

#' Add display labels to correlation results.
enrich_correlations <- function(corr) {
  corr |>
    mutate(
      proxy_type = factor(proxy_type, levels = proxy_order),
      proxy_label = proxy_labels[as.character(proxy_type)],
      best_var_label = recode(
        best_var,
        tmp_ann = "Temperature (annual)",
        tmp_jja = "Temperature (JJA)",
        pre_ann = "Precipitation (annual)",
        .default = NA_character_
      )
    )
}
```

## All stations

```{r}
#| label: corr-run
#| cache: true
#| dependson: speleothem-summary

corr_rds <- file.path(cache_dir, "speleo_instrument_correlations.rds")
cru_locs <- speleo_sites |> distinct(lon, lat) |> mutate(loc_id = row_number())

if (file.exists(corr_rds) && file.exists(cru_rds)) {
  speleo_corr <- enrich_correlations(readRDS(corr_rds))
} else {
  cru_clim <- load_cru_site_climate(cru_locs)
  speleo_corr <- enrich_correlations(compute_speleo_correlations(catalog, cru_clim, speleo_sites))
  saveRDS(speleo_corr, corr_rds)
}

n_tested <- sum(is.finite(speleo_corr$best_r))
n_sig <- speleo_corr |>
  transmute(sig = p_tmp_ann < 0.05 | p_tmp_jja < 0.05 | p_pre_ann < 0.05) |>
  pull(sig) |>
  sum(na.rm = TRUE)
glue(
  "Series tested (≥{min_overlap} overlap years, {instr_y0}–{instr_y1}): {n_tested}  |  ",
  "Significant (p < 0.05, any variable): {n_sig}"
)
```

```{r}
#| label: corr-summary-table
#| dependson: corr-run

speleo_corr |>
  filter(is.finite(r_tmp_ann) | is.finite(r_pre_ann)) |>
  group_by(proxy_type) |>
  summarize(
    n = n(),
    median_r_tmp = median(r_tmp_ann, na.rm = TRUE),
    median_r_pre = median(r_pre_ann, na.rm = TRUE),
    sig_tmp = sum(p_tmp_ann < 0.05, na.rm = TRUE),
    sig_pre = sum(p_pre_ann < 0.05, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(proxy_label = proxy_labels[as.character(proxy_type)]) |>
  transmute(
    proxy_label, n, median_r_tmp, median_r_pre, sig_tmp, sig_pre
  ) |>
  knitr::kable(
    digits = 3,
    caption = paste0("Local CRU correlations by proxy type (", instr_y0, "–", instr_y1, ")")
  )
```

Distribution of correlation coefficients across all series — weak median correlations are typical for heterogeneous archives (cf. Exercise 12).

```{r}
#| label: corr-histogram-fn
#| code-fold: true

#' Histogram of correlation coefficients, faceted by climate variable.
plot_correlation_histogram <- function(corr) {
  long <- corr |>
    filter(is.finite(r_tmp_ann) | is.finite(r_tmp_jja) | is.finite(r_pre_ann)) |>
    pivot_longer(
      c(r_tmp_ann, r_tmp_jja, r_pre_ann),
      names_to = "climate", values_to = "r"
    ) |>
    mutate(
      climate = factor(climate, levels = names(climate_vars), labels = climate_vars),
      proxy_type = factor(proxy_type, levels = proxy_order)
    )

  ggplot(long, aes(r, fill = proxy_type)) +
    geom_histogram(bins = 30, alpha = 0.75, position = "identity") +
    facet_wrap(~climate, ncol = 1, scales = "free_y") +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    scale_fill_manual(values = proxy_colors, labels = proxy_labels[proxy_order], name = "Proxy type") +
    labs(
      title = "Proxy–climate correlations (all speleothem series)",
      subtitle = paste0("Nearest CRU TS grid cell, ", instr_y0, "–", instr_y1),
      x = "Pearson r", y = "Count"
    ) +
    theme_clean()
}
```

```{r}
#| label: corr-histogram
#| fig-cap: "Distribution of correlation coefficients"
#| dependson: [corr-run, corr-histogram-fn]

plot_correlation_histogram(speleo_corr)
```

## By proxy type

Compare temperature vs. precipitation fit and count significant correlations (p < 0.05).

```{r}
#| label: corr-by-type-fn
#| code-fold: true

#' Boxplot of correlations by proxy type and climate variable.
plot_correlation_by_type <- function(corr) {
  long <- corr |>
    pivot_longer(
      c(r_tmp_ann, r_tmp_jja, r_pre_ann),
      names_to = "climate", values_to = "r"
    ) |>
    mutate(
      climate = factor(climate, levels = names(climate_vars), labels = climate_vars),
      proxy_type = factor(proxy_type, levels = proxy_order)
    )

  ggplot(long, aes(proxy_type, r, fill = proxy_type)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_boxplot(outlier.size = 0.7, alpha = 0.85) +
    facet_wrap(~climate, ncol = 1) +
    scale_fill_manual(values = proxy_colors, guide = "none") +
    scale_x_discrete(labels = proxy_labels[proxy_order]) +
    labs(
      title = "Correlation by proxy type",
      x = NULL, y = "Pearson r"
    ) +
    theme_clean()
}

#' Count significant correlations (p < 0.05) by proxy type and climate variable.
plot_significance_counts <- function(corr) {
  sig <- corr |>
    transmute(
      proxy_type,
      tmp_ann = p_tmp_ann < 0.05,
      tmp_jja = p_tmp_jja < 0.05,
      pre_ann = p_pre_ann < 0.05
    ) |>
    pivot_longer(-proxy_type, names_to = "climate", values_to = "sig") |>
    filter(!is.na(sig)) |>
    group_by(proxy_type, climate) |>
    summarize(n_sig = sum(sig), .groups = "drop") |>
    mutate(
      climate = factor(climate, levels = names(climate_vars), labels = climate_vars),
      proxy_type = factor(proxy_type, levels = proxy_order)
    )

  ggplot(sig, aes(climate, n_sig, fill = proxy_type)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    scale_fill_manual(values = proxy_colors, labels = proxy_labels[proxy_order], name = "Proxy type") +
    labs(
      title = "Significant correlations (p < 0.05)",
      x = NULL, y = "Number of series"
    ) +
    theme_clean()
}
```

```{r}
#| label: corr-by-type
#| fig-cap: "Correlations and significance by proxy type"
#| dependson: [corr-run, corr-by-type-fn]

plot_correlation_by_type(speleo_corr)
```

```{r}
#| label: corr-significance
#| dependson: [corr-run, corr-by-type-fn]

plot_significance_counts(speleo_corr)
```

## Best-fit climate variable

For each series, which of the three climate targets has the highest |r|?

```{r}
#| label: corr-best-var-fn
#| code-fold: true

plot_best_variable_counts <- function(corr) {
  best <- corr |>
    filter(!is.na(best_var)) |>
    count(proxy_type, best_var_label, name = "n") |>
    mutate(proxy_type = factor(proxy_type, levels = proxy_order))

  ggplot(best, aes(best_var_label, n, fill = proxy_type)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    scale_fill_manual(values = proxy_colors, labels = proxy_labels[proxy_order], name = "Proxy type") +
    labs(
      title = "Strongest |r| per series",
      subtitle = "Which climate variable fits best?",
      x = NULL, y = "Number of series"
    ) +
    theme_clean() +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r}
#| label: corr-best-var
#| dependson: [corr-run, corr-best-var-fn]

plot_best_variable_counts(speleo_corr)
```

## Correlation map

Points coloured by the strongest local correlation (signed).

```{r}
#| label: corr-map-fn
#| code-fold: true

#' Map speleothem sites coloured by best correlation with CRU climate.
plot_correlation_map <- function(corr) {
  map_df <- corr |>
    filter(is.finite(best_r)) |>
    left_join(
      speleo_sites |> select(dataset_id, site),
      by = "dataset_id"
    ) |>
    mutate(
      hover = paste0(
        "<b>", site, "</b><br>", dataset_id, "<br>",
        proxy_labels[as.character(proxy_type)], "<br>",
        "Best: ", best_var_label, "<br>",
        "r = ", round(best_r, 3), " (n = ", n_overlap, ")"
      )
    )

  map_df |>
    plot_ly(
      type = "scattergeo", mode = "markers",
      lon = ~lon, lat = ~lat, text = ~hover, hoverinfo = "text",
      color = ~best_r, colors = c("#2166ac", "#f7f7f7", "#b2182b"),
      marker = list(size = 7, line = list(width = 0.4, color = "#333"))
    ) |>
    layout(
      title = list(
        text = paste0("<b>Best proxy–climate correlation</b><br><sup>CRU TS, ", instr_y0, "–", instr_y1, "</sup>"),
        x = 0.5, xanchor = "center"
      ),
      geo = list(
        projection = list(type = "natural earth"),
        showland = TRUE, landcolor = "#f0f0f0",
        showocean = TRUE, oceancolor = "#e6f2fa",
        showcountries = TRUE, countrycolor = "#bbbbbb"
      )
    )
}
```

```{r}
#| label: corr-map
#| dependson: [corr-run, corr-map-fn, speleothem-summary]

plot_correlation_map(speleo_corr)
```

## Example fits

One well-correlated series per proxy type — proxy vs. climate scatter for the variable with highest |r|.

```{r}
#| label: corr-examples-fn
#| code-fold: true

#' Scatter plots for the best-correlated series per proxy type.
plot_correlation_examples <- function(corr, catalog, clim) {
  loc <- speleo_sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
  proxy_ann <- speleothem_annual(catalog) |> inner_join(loc, by = c("lon", "lat"))

  picks <- corr |>
    filter(is.finite(best_r)) |>
    group_by(proxy_type) |>
    slice_max(abs(best_r), n = 1, with_ties = FALSE) |>
    ungroup()

  clim_col <- c(tmp_ann = "tmp_ann", tmp_jja = "tmp_jja", pre_ann = "pre_ann")

  examples <- map_dfr(seq_len(nrow(picks)), function(i) {
    row <- picks[i, ]
    px <- proxy_ann |> filter(dataset_id == row$dataset_id)
    cd <- clim |>
      filter(loc_id == px$loc_id[1]) |>
      select(year, climate = all_of(clim_col[[row$best_var]]))
    px |>
      filter(year >= instr_y0, year <= instr_y1) |>
      inner_join(cd, by = "year") |>
      mutate(
        dataset_id = row$dataset_id,
        proxy_label = proxy_labels[as.character(row$proxy_type)],
        climate_label = row$best_var_label,
        r = row$best_r
      )
  })

  ggplot(examples, aes(climate, value)) +
    geom_point(alpha = 0.55, size = 1.5, colour = "#444444") +
    geom_smooth(method = "lm", se = TRUE, colour = "#b2182b", linewidth = 0.8) +
    facet_wrap(~paste0(proxy_label, " — ", dataset_id, "\nr = ", round(r, 2)),
               scales = "free", ncol = 2) +
    labs(
      title = "Best local fits per proxy type",
      subtitle = paste0("CRU TS grid box, ", instr_y0, "–", instr_y1),
      x = "Climate (CRU TS)", y = "Proxy value"
    ) +
    theme_clean() +
    theme(strip.text = element_text(size = 8))
}
```

```{r}
#| label: corr-examples
#| fig-cap: "Example proxy–climate scatter plots"
#| dependson: corr-run

if (!exists("cru_clim")) cru_clim <- readRDS(cru_rds)
plot_correlation_examples(speleo_corr, catalog, cru_clim)
```
