---
title: "DoD2k speleothem data overview"
subtitle: "Load, explore, and visualise the archive"
author:
  - "Madina Vogt"
  - "Florian Christ"
  - "Lukas Batschelet"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 120
execute:
  warning: false
  message: false
---

All R code lives in this notebook. Functions are defined in the **setup** chunk below and documented in **Roxygen style** (`#'` with `@param`, `@return`, `@details`).

```{r}
#| label: setup
#| include: false

# =============================================================================
# Setup — libraries, paths, DoD2k I/O, speleothem plots
# =============================================================================
#
# Data flow:
#   dod2k_v2.0 compact CSV  →  load_dod2k_compact()  →  one row per dataset
#   compact                  →  compact_to_catalog()  →  long tibble (year × value)
#   catalog                  →  saveRDS cache          →  fast reload on next run
#
# DoD2k field names (see https://lluecke.github.io/dod2k/):
#   archiveType      — archive material (Speleothem, Wood, Coral, …)
#   paleoData_proxy  — measurement type (d18O, d13C, growth rate, Mg/Ca, …)
# =============================================================================

library(tidyverse)
library(glue)
library(plotly)


# --- Paths -------------------------------------------------------------------

#' Find repository root containing `course/data/`.
#'
#' Walks upward from the working directory until `course/data` exists.
#' Works when knitting from `proxy-cfr-comparison/` or the repo root.
#'
#' @return Character scalar — absolute path to the repository root.
#' @throws Error if `course/data` cannot be found.
find_repo_root <- function() {
  candidates <- c(
    if (dir.exists("course/data")) normalizePath(".", mustWork = TRUE),
    if (dir.exists("../course/data")) normalizePath("..", mustWork = TRUE),
    if (dir.exists("../../course/data")) normalizePath("../..", mustWork = TRUE)
  )
  root <- candidates[!sapply(candidates, is.null)][1]
  if (is.na(root)) {
    stop("Cannot find course/data — open this file from proxy-cfr-comparison/")
  }
  root
}

repo_root <- find_repo_root()
proj_root <- if (file.exists("analysis.qmd")) "." else "proxy-cfr-comparison"
data_dir  <- file.path(repo_root, "course", "data", "dod2k_v2.0")
cache_dir <- file.path(proj_root, "output", "cache")
cache_rds <- file.path(cache_dir, "dod2k_timeseries.rds")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

# Reconstruction period on the DoD2k file axis (calendar years CE)
year_min <- 0L
year_max <- 2000L


# --- DoD2k compact CSV parser ------------------------------------------------

#' Parse one row from a DoD2k *compact array* CSV file.
#'
#' Compact files store comma-separated numeric arrays per dataset, e.g.
#' `sisal_001,1000.0,1001.0,1002.0,…` (years or proxy values).
#'
#' @param line Character scalar — one line from the CSV (without header).
#' @return Named list with `id` (dataset id) and `values` (numeric vector).
parse_array_line <- function(line) {
  parts <- strsplit(line, ",", fixed = TRUE)[[1]]
  list(id = parts[[1]], values = as.numeric(parts[-1]))
}

#' Read a DoD2k compact array column into a tibble.
#'
#' @param path Path to e.g. `dod2k_v2.0_compact_year.csv`.
#' @return Tibble with columns `dataset_id` and `values` (list of numeric vectors).
read_array_column <- function(path) {
  lines <- readLines(path, warn = FALSE)
  parsed <- lapply(lines[-1], parse_array_line)  # skip header row
  tibble(
    dataset_id = vapply(parsed, `[[`, character(1), "id"),
    values = lapply(parsed, `[[`, "values")
  )
}

#' Load the DoD2k v2.0 compact CSV bundle (metadata + years + values).
#'
#' Joins three files from `data_dir`:
#' `_compact_metadata.csv`, `_compact_year.csv`, `_compact_paleoData_values.csv`.
#'
#' @return Tibble — one row per dataset; `year` and `paleoData_values` are list-columns
#'   of numeric vectors (full time series per record).
#' @details This is the *wide* representation. Use [compact_to_catalog()] for analysis.
load_dod2k_compact <- function() {
  prefix <- basename(data_dir)
  meta <- read_csv(
    file.path(data_dir, paste0(prefix, "_compact_metadata.csv")),
    show_col_types = FALSE
  ) |>
    rename(dataset_id = any_of(c("datasetId", "...1")))

  years <- read_array_column(file.path(data_dir, paste0(prefix, "_compact_year.csv"))) |>
    rename(year = values)
  values <- read_array_column(
    file.path(data_dir, paste0(prefix, "_compact_paleoData_values.csv"))
  ) |>
    rename(paleoData_values = values)

  meta |>
    left_join(years, by = "dataset_id") |>
    left_join(values, by = "dataset_id")
}

#' Convert compact DoD2k data to a long-format catalog.
#'
#' Expands list-columns into one row per `(dataset_id, year)` with a finite proxy value
#' inside `[year_min, year_max]`.
#'
#' @param compact Output of [load_dod2k_compact()].
#' @return Tibble with columns:
#'   `dataset_id`, `year`, `value`, `lon`, `lat`, `archive_type`, `proxy_type`.
#' @details Rows with missing or out-of-range years are dropped silently.
compact_to_catalog <- function(compact) {
  rows <- vector("list", nrow(compact))
  n <- 0L
  for (i in seq_len(nrow(compact))) {
    rec <- compact[i, ]
    yrs  <- rec$year[[1]]
    vals <- rec$paleoData_values[[1]]
    if (!length(yrs) || !length(vals)) next

    ok <- yrs >= year_min & yrs <= year_max & is.finite(vals)
    if (!any(ok)) next

    n <- n + 1L
    rows[[n]] <- tibble(
      dataset_id = rec$dataset_id,
      year = as.integer(yrs[ok]),
      value = as.double(vals[ok]),
      lon = rec$geo_meanLon,
      lat = rec$geo_meanLat,
      archive_type = rec$archiveType,
      proxy_type = rec$paleoData_proxy
    )
  }
  if (!n) return(tibble())
  bind_rows(rows)
}

#' Read the cached long-format catalog from disk.
#'
#' @return Tibble or `NULL` if the RDS cache file does not exist yet.
read_catalog_cache <- function() {
  if (!file.exists(cache_rds)) return(NULL)
  readRDS(cache_rds) |>
    mutate(
      dataset_id = as.character(dataset_id),
      archive_type = as.character(archive_type),
      proxy_type = as.character(proxy_type),
      year = as.integer(year),
      value = as.double(value),
      lon = as.double(lon),
      lat = as.double(lat)
    )
}

#' Load (or build) the long-format DoD2k catalog.
#'
#' On first call, parses compact CSV and writes `output/cache/dod2k_timeseries.rds`.
#' Subsequent calls read the cache unless `refresh = TRUE`.
#'
#' @param refresh Logical — if `TRUE`, rebuild cache from CSV even when RDS exists.
#' @return Long-format catalog tibble (see [compact_to_catalog()]).
load_dod2k_catalog <- function(refresh = FALSE) {
  if (!refresh) {
    cached <- read_catalog_cache()
    if (!is.null(cached)) return(cached)
  }
  message("Building catalog from compact CSV …")
  catalog <- compact_to_catalog(load_dod2k_compact())
  saveRDS(catalog, cache_rds)
  catalog
}


# --- Speleothem constants (DoD2k legend) -------------------------------------

#' Display order, labels, colours, and map symbols for speleothem proxy types.
proxy_order   <- c("d18O", "d13C", "growth rate", "Mg/Ca")
proxy_labels  <- c(d18O = "δ18O", d13C = "δ13C", `growth rate` = "growth rate", `Mg/Ca` = "Mg/Ca")
proxy_colors  <- c(d18O = "#b2182b", d13C = "#d6604d", `growth rate` = "#f4a582", `Mg/Ca` = "#92c5de")
proxy_symbols <- c(d18O = "triangle-up", d13C = "triangle-down", `growth rate` = "square", `Mg/Ca` = "circle")

#' Minimal ggplot theme for overview figures.
#'
#' @return A [ggplot2::theme] object.
theme_clean <- function() {
  theme_minimal() + theme(legend.position = "bottom")
}


# --- Speleothem tables -------------------------------------------------------

#' Build one row per speleothem time series with site metadata.
#'
#' @param catalog Long-format catalog from [load_dod2k_catalog()].
#' @param compact Wide compact table from [load_dod2k_compact()] (for site names).
#' @return Tibble with columns:
#'   `dataset_id`, `proxy_type`, `lon`, `lat`, `year_start`, `year_end`,
#'   `n_years`, `span_years`, `site`.
#' @details
#'   `n_years` counts dated observations; `span_years` is calendar range
#'   (last − first + 1) and can be much larger than `n_years`.
speleothem_sites <- function(catalog, compact) {
  sites <- catalog |>
    filter(archive_type == "Speleothem") |>
    group_by(dataset_id, proxy_type, lon, lat) |>
    summarize(
      year_start = min(year),
      year_end = max(year),
      n_years = n(),
      .groups = "drop"
    ) |>
    mutate(span_years = year_end - year_start + 1L)

  meta <- compact |>
    filter(archiveType == "Speleothem") |>
    dplyr::select(dataset_id, geo_siteName) |>
    rename(site = geo_siteName)

  sites |>
    left_join(meta, by = "dataset_id") |>
    mutate(
      site = if_else(is.na(site) | site == "" | site == dataset_id, dataset_id, site),
      proxy_type = factor(proxy_type, levels = proxy_order)
    )
}

#' Count speleothem records in metadata vs. records with usable time series.
#'
#' @param compact Wide compact table (metadata rows).
#' @param catalog Long catalog (after year/value filtering).
#' @return Tibble: `proxy_label`, `proxy_type`, `n_metadata`, `n_with_data`.
speleothem_summary <- function(compact, catalog) {
  meta <- compact |>
    filter(archiveType == "Speleothem") |>
    count(paleoData_proxy, name = "n_metadata") |>
    rename(proxy_type = paleoData_proxy)

  with_data <- catalog |>
    filter(archive_type == "Speleothem") |>
    distinct(dataset_id, proxy_type) |>
    count(proxy_type, name = "n_with_data")

  meta |>
    full_join(with_data, by = "proxy_type") |>
    mutate(
      n_metadata = replace_na(n_metadata, 0L),
      n_with_data = replace_na(n_with_data, 0L),
      proxy_label = proxy_labels[proxy_type],
      proxy_type = factor(proxy_type, levels = proxy_order)
    ) |>
    arrange(proxy_type) |>
    transmute(proxy_label, proxy_type, n_metadata, n_with_data)
}


# --- Speleothem plots --------------------------------------------------------

#' Interactive world map of speleothem sites (plotly).
#'
#' Points are coloured and symbolled by `paleoData_proxy` (DoD2k legend).
#' Hover shows site name, dataset id, proxy type, year range, and observation count.
#'
#' @param sites Output of [speleothem_sites()].
#' @return A plotly object (renders interactively in HTML).
plot_speleothem_map <- function(sites) {
  sites <- sites |>
    mutate(
      proxy_label = proxy_labels[as.character(proxy_type)],
      hover = paste0(
        "<b>", site, "</b><br>", dataset_id, "<br>",
        "Proxy: ", proxy_label, "<br>",
        "Years: ", year_start, "–", year_end, " (n = ", n_years, ")"
      )
    )

  p <- plot_ly(type = "scattergeo", mode = "markers")
  for (px in proxy_order) {
    sub <- filter(sites, as.character(proxy_type) == px)
    if (!nrow(sub)) next
    p <- add_trace(
      p, data = sub, lon = ~lon, lat = ~lat, text = ~hover, hoverinfo = "text",
      name = paste0("Speleothem: ", proxy_labels[[px]], " (n=", nrow(sub), ")"),
      marker = list(
        symbol = proxy_symbols[[px]], size = 9,
        color = proxy_colors[[px]], line = list(width = 0.6, color = "#333")
      )
    )
  }
  p |> layout(
    title = list(
      text = paste0("<b>DoD2k speleothem archive</b><br><sup>", nrow(sites), " series</sup>"),
      x = 0.5, xanchor = "center"
    ),
    geo = list(
      projection = list(type = "natural earth"),
      showland = TRUE, landcolor = "#f0f0f0",
      showocean = TRUE, oceancolor = "#e6f2fa",
      showcountries = TRUE, countrycolor = "#bbbbbb",
      lataxis = list(range = c(-40, 75))
    ),
    legend = list(orientation = "h", y = -0.05, x = 0.5, xanchor = "center")
  )
}

#' Stacked bar chart — how many series have data in each time bin.
#'
#' Counts series with ≥1 observation per `(bin, proxy_type)`.
#' Uses *actual dated years*, not calendar first/last year.
#'
#' @param catalog Long-format catalog.
#' @param bin_width Width of time bins in years (default 100 = centuries).
#' @return A ggplot object.
plot_speleothem_density <- function(catalog, bin_width = 100) {
  density <- catalog |>
    filter(archive_type == "Speleothem") |>
    mutate(bin_start = (as.integer(year %/% bin_width)) * bin_width) |>
    distinct(dataset_id, proxy_type, bin_start) |>
    count(bin_start, proxy_type, name = "n_series") |>
    mutate(
      proxy_type = factor(proxy_type, levels = proxy_order),
      bin_mid = bin_start + bin_width / 2
    )

  ggplot(density, aes(bin_mid, n_series, fill = proxy_type)) +
    geom_col(width = bin_width * 0.9, colour = "white", linewidth = 0.2) +
    scale_fill_manual(values = proxy_colors, labels = proxy_labels[proxy_order], name = "Proxy type") +
    scale_x_continuous(limits = c(0, 2000), breaks = seq(0, 2000, 250), name = "Year (CE)") +
    labs(
      title = "Where observations exist",
      subtitle = "Stacked by proxy type (100-year bins)",
      y = "Series with data"
    ) +
    theme_clean()
}

#' Boxplot of observation counts per speleothem series.
#'
#' Shows `n_years` (number of dated values) by proxy type.
#' Prefer this over calendar span — many series cover 0–2000 CE nominally
#' but contain far fewer measurements.
#'
#' @param sites Output of [speleothem_sites()].
#' @return A ggplot object.
plot_speleothem_resolution <- function(sites) {
  ggplot(sites, aes(proxy_type, n_years, fill = proxy_type)) +
    geom_boxplot(outlier.size = 0.8, alpha = 0.85) +
    scale_fill_manual(values = proxy_colors, guide = "none") +
    scale_x_discrete(labels = proxy_labels[proxy_order]) +
    labs(
      title = "Temporal resolution per series",
      subtitle = "Number of dated values — calendar span is often misleading",
      x = NULL, y = "Observations per series"
    ) +
    theme_clean()
}
```

# Introduction

Explore **DoD2k v2.0** speleothem records from [DoD2k](https://lluecke.github.io/dod2k/).

DoD2k distinguishes **archive** (`archiveType`, e.g. Speleothem) from **measurement** (`paleoData_proxy`: δ18O, δ13C, growth rate, Mg/Ca).

# 1. Load DoD2k

Data: `course/data/dod2k_v2.0/`. First run builds `output/cache/dod2k_timeseries.rds`; later runs are fast.

```{r}
#| label: dod2k-load
#| cache: true

# Wide metadata + list-columns (one row per dataset)
dod2k_compact <- load_dod2k_compact()
# Long catalog: one row per (dataset, year) with a finite value in 0–2000 CE
catalog       <- load_dod2k_catalog()

# Cross-tab of all archive types in the metadata (not only speleothems)
archive_summary <- dod2k_compact |>
  count(archiveType, paleoData_proxy, name = "n", sort = TRUE) |>
  rename(archive_type = archiveType, proxy_type = paleoData_proxy)

glue(
  "Records: {format(nrow(dod2k_compact), big.mark = ',')}  |  ",
  "Speleothems: {sum(dod2k_compact$archiveType == 'Speleothem', na.rm = TRUE)}  |  ",
  "δ18O: {sum(dod2k_compact$archiveType == 'Speleothem' & dod2k_compact$paleoData_proxy == 'd18O', na.rm = TRUE)}"
)

knitr::kable(head(archive_summary, 12), caption = "DoD2k archive × proxy (metadata)")
```

# 2. Speleothem overview

**`n_metadata`** = rows in the archive; **`n_with_data`** = series with values in 0–2000 CE.

```{r}
#| label: speleothem-summary
#| cache: true
#| dependson: dod2k-load

# One row per speleothem series (site coords, year range, observation count)
speleo_sites <- speleothem_sites(catalog, dod2k_compact)

speleothem_summary(dod2k_compact, catalog) |>
  knitr::kable(caption = "Speleothem proxy types")
```

```{r}
#| label: speleothem-map
#| dependson: speleothem-summary

# Interactive plotly map — colour/symbol by proxy type
plot_speleothem_map(speleo_sites)
```

```{r}
#| label: speleothem-density
#| fig-cap: "Series with observations per century"
#| dependson: speleothem-summary

# Temporal coverage: how many series have ≥1 dated value per 100-year bin
plot_speleothem_density(catalog)
```

```{r}
#| label: speleothem-resolution
#| fig-cap: "Dated values per series"
#| dependson: speleothem-summary

# Resolution: number of dated values per series (not calendar span)
plot_speleothem_resolution(speleo_sites)
```
