---
title: "Proxy network CFR comparison"
subtitle: "DoD2k speleothems vs tree-ring baselines"
author: "Climate reconstruction project"
date: last-modified
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: false
    code-tools: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 120
params:
  quick_run: true
  overwrite_cfr: false
execute:
  warning: false
  message: false
jupyter: dod2k-cfr
---

One Quarto notebook: **`{python}`** chunks for DoD2k, **`{r}`** chunks for CFR and tidyverse analysis.  
Languages meet via a Parquet cache (`output/cache/dod2k_timeseries.parquet`) — no R code calling Python at runtime.

```{r}
#| label: setup-r

source(file.path(
  if (file.exists("R/load_project.R")) "." else "proxy-cfr-comparison",
  "R", "load_project.R"
))
load_project()

library(tidyverse)
library(glue)

if (!requireNamespace("arrow", quietly = TRUE)) {
  stop(
    "R package 'arrow' is required for the Parquet bridge. Run:\n",
    "  Rscript -e \"install.packages('arrow', repos='https://cloud.r-project.org')\"\n",
    "If permission errors occur, remove any 00LOCK-arrow in your R library and retry."
  )
}

quick_run <- if (exists("params", inherits = TRUE)) params$quick_run else TRUE
overwrite_cfr <- if (exists("params", inherits = TRUE)) params$overwrite_cfr else FALSE

# RStudio: use Quarto "Run Cell" on {python} chunks (kernel dod2k-cfr), not reticulate::repl_python()
if (requireNamespace("reticulate", quietly = TRUE)) {
  venv <- file.path(
    if (file.exists(".venv")) "." else "proxy-cfr-comparison",
    ".venv"
  )
  if (dir.exists(venv)) reticulate::use_virtualenv(venv, required = FALSE)
}
```

# Introduction

Compare **climate field reconstructions (CFR)** across proxy networks (DoD2k speleothems vs tree rings).

| Step | Language | Task |
|------|----------|------|
| DoD2k load & filter | Python | [dod2k toolkit](https://lluecke.github.io/dod2k/) |
| CFR, validation, plots | R | tidyverse + `Master_CFR.R` |

DoD2k distinguishes **archive** (`archiveType`, e.g. Speleothem) from **measurement** (`paleoData_proxy`, e.g. δ18O, δ13C, growth rate, Mg/Ca). CFR below uses speleothem **δ18O** only; the map shows all four speleothem proxy types.

**Setup (once):** `bash scripts/setup_python.sh` — then render or **Run All** on this file. Python chunks use the Jupyter kernel `dod2k-cfr` (not `reticulate::repl_python()`). R needs `arrow` for the Parquet bridge.

# 1. DoD2k (Python)

```{python}
#| label: dod2k-load

import sys
from pathlib import Path

for _cand in (Path.cwd(), Path.cwd() / "proxy-cfr-comparison"):
    _py = _cand / "python"
    if _py.is_dir() and str(_py) not in sys.path:
        sys.path.insert(0, str(_py))

from project_env import bootstrap_python_path, data_paths
from dod2k_io import load_dod2k, summarize_archives, filter_dod2k, write_dod2k_cache

root = bootstrap_python_path()
paths = data_paths(root)

dod2k_df = load_dod2k(paths["dod2k_csv"])
archive_summary = summarize_archives(dod2k_df)
speleo = filter_dod2k(dod2k_df, archive_types=["Speleothem"])
speleo_d18o = filter_dod2k(dod2k_df, archive_types=["Speleothem"], proxy_types=["d18O"])

cache_path = write_dod2k_cache(dod2k_df, paths["cache"])
print(f"Records: {len(dod2k_df):,}  |  Speleothems (all types): {len(speleo):,}  |  δ18O: {len(speleo_d18o):,}")
print(f"Cache: {cache_path}")
archive_summary.head(12)
```

## Speleothem proxy types

Per [DoD2k documentation](https://lluecke.github.io/dod2k/), speleothems in v2.0 carry four `paleoData_proxy` values (symbols match the official plots). **`n_metadata`** = archive rows; **`n_with_data`** = series with finite values in 0–2000 CE (used on the map).

```{python}
#| label: speleothem-summary
#| dependson: dod2k-load

from speleothem_viz import speleothem_records, speleothem_summary_table

speleo_records = speleothem_records(dod2k_df)
speleothem_summary_table(dod2k_df)
```

```{python}
#| label: speleothem-map
#| dependson: speleothem-summary

from speleothem_viz import build_speleothem_map

map_fig = build_speleothem_map(speleo_records)
map_fig.show()
```

```{python}
#| label: speleothem-timeline
#| dependson: speleothem-summary

from speleothem_viz import build_speleothem_timeline

timeline_fig = build_speleothem_timeline(speleo_records, max_rows=70)
timeline_fig.show()
```

# 2. Proxy networks (R)

```{r}
#| label: preview-networks
#| dependson: dod2k-load

calib <- CALIBRATION_SCHEMES$course_split$calibration
catalog <- read_dod2k_catalog()

archive_summary_r <- summarize_dod2k_catalog(catalog)
print(archive_summary_r, n = 12)

network_ids <- c("dod2k_speleothems", "dod2k_trees", "dod2k_speleo_trees", "ntrend_trees")

proxy_map_df <- network_ids |>
  map(\(id) {
    build_proxy_network(
      id,
      calib_start = calib[1],
      calib_end = calib[2],
      catalog = catalog
    ) |>
      proxy_network_map_df() |>
      mutate(label = PROXY_NETWORKS[[id]]$label)
  }) |>
  list_rbind()

proxy_map_df |>
  distinct(network, label) |>
  knitr::kable(caption = "Networks after calibration coverage filter")
```

```{r}
#| label: plot-proxy-locations
#| fig-cap: "Proxy locations"
#| dependson: preview-networks

proxy_map_df |>
  ggplot(aes(lon, lat, colour = network)) +
  geom_point(size = 2.2, alpha = 0.85) +
  borders("world", fill = NA, colour = "grey70") +
  coord_fixed(1.3) +
  labs(title = "Proxy networks", x = "Longitude", y = "Latitude") +
  theme_minimal()
```

# 3. Run reconstructions (R)

```{r}
#| label: select-experiments

selected <- if (quick_run) {
  tribble(
    ~network_id,           ~calibration_id,
    "dod2k_speleothems",   "course_split",
    "dod2k_speleo_trees",  "course_split",
    "ntrend_trees",        "course_split"
  )
} else {
  EXPERIMENT_PLAN |>
    filter(run) |>
    select(network_id, calibration_id)
}

selected <- selected |>
  mutate(experiment_id = str_c(network_id, calibration_id, sep = "__"))

knitr::kable(selected, caption = "Experiments to run")
```

```{r}
#| label: run-cfr
#| cache: true
#| dependson: select-experiments

results <- selected |>
  pmap(\(network_id, calibration_id, experiment_id) {
    message(glue("=== {experiment_id} ==="))
    run_cfr_experiment(
      network_id = network_id,
      calibration_id = calibration_id,
      overwrite = overwrite_cfr
    )
  }) |>
  set_names(selected$experiment_id)

ok <- map_lgl(results, \(x) is.null(x$error))
if (!any(ok)) abort("All CFR experiments failed.")
if (!all(ok)) {
  warn(glue("Failed: {paste(names(results)[!ok], collapse = ', ')}"))
}
```

# 4. Validation (R)

```{r}
#| label: validate-all
#| dependson: run-cfr

skill_summary <- results[ok] |>
  map(\(res) validate_cfr_result(res, calibration_id = res$calibration_id)) |>
  print_skill_summary()

knitr::kable(skill_summary, caption = "Skill summary")
```

```{r}
#| label: plot-maps
#| fig-height: 6
#| fig-cap: "Validation correlation r"
#| dependson: validate-all

walk(
  results[ok],
  \(res) {
    skill <- validate_cfr_result(res, calibration_id = res$calibration_id)
    print(
      plot_skill_map(
        skill$grid_validation$r,
        lon = res$reconstruction$lon,
        lat = res$reconstruction$lat,
        title = glue("Validation r — {res$experiment_id}"),
        limits = c(-1, 1)
      )
    )
  }
)
```

# 5. Discussion

DoD2k v2.0 contains hundreds of speleothem δ18O records; tree-ring networks give dense NH coverage. Combined networks change the PCA basis of the reconstruction.

---

```{r}
#| label: session

sessionInfo()
```
