---
title: "Speleothem proxy screening for climate reconstruction"
subtitle: "DoD2k screening, regional composites, and ENSO — course project report"
author:
  - "Madina Vogt"
  - "Florian Christ"
  - "Lukas Batschelet"
date: last-modified
bibliography: references.bib
csl: https://www.zotero.org/styles/american-geophysical-union
format:
  html:
    toc: true
    toc-depth: 3
    number-sections: true
    code-fold: true
    code-tools: true
    fig-width: 8
    fig-height: 5
    fig-dpi: 120
execute:
  warning: false
  message: false
---

# Introduction

**Question:** Can speleothem records from DoD2k v2.0 [@evans2026] be used to reconstruct climate — and is there an ENSO signal in regions where literature expects one [@tadros2016]?

**Data:** speleothem time series (δ18O, δ13C, growth rate, Mg/Ca) from **DoD2k v2.0** / SISALv3 [@kaushal2024]; instrumental **CRU TS v4.07** [@harris2020] at the nearest grid box; **Niño3.4** SST anomalies (NOAA PSL) for ENSO.

**Why this matters:** Rain and temperature at the surface do not reach a stalagmite unchanged — water mixes in the karst aquifer [@fairchild2006]. We therefore test proxies **site by site** and **region by region**, not as one global average.

## Three analysis steps

| Step | What we do | Why |
|------|------------|-----|
| **1 — Screen** | Correlate each series with local CRU temperature and precipitation (1901–2000) | See whether any proxy tracks instruments at all |
| **2 — Composite** | Detrend, calibrate 1901–1980, validate 1981–2000; regional mean of screened series | Test whether a *group* of caves in one region carries a reconstructable signal |
| **3 — ENSO** | Focus on Australasia / SW Pacific; compare proxies to Niño3.4 and local rainfall | Literature links SE Australian hydroclimate to ENSO [@tadros2016] |

Instrumental window: **1901–2000 CE**. Calibration: **1901–1980**; validation: **1981–2000**.

## Proxies at a glance

| Proxy | Often reflects | Caveat |
|-------|----------------|--------|
| **δ18O** | Rainfall amount, moisture source (region-dependent) | Strongly filtered by karst mixing |
| **Mg/Ca** | Cave temperature or moisture deficit (PCP) | In SE Australia linked to palaeorainfall under ENSO [@tadros2016] |
| **δ13C**, **growth rate** | Soil activity, wet vs. dry phases | Indirect climate indicators |

```{r}
#| label: setup
#| include: false

library(tidyverse)
library(glue)
library(plotly)
library(terra)
select <- dplyr::select
```

# Step 1 — Explore the archive

## Load DoD2k data

Data live in `course/data/dod2k_v2.0/`. The compact format stores one row per dataset; year and proxy values are comma-separated arrays in separate files.

**Flow:** compact CSV → wide table (`load_dod2k_compact`) → long catalog (`load_dod2k_catalog`) → RDS cache for fast reload.

## Paths and cache

```{r}
#| label: paths
#| code-fold: true

#' Find repository root containing `course/data/`.
#'
#' @return Absolute path to the repository root.
find_repo_root <- function() {
  candidates <- c(
    if (dir.exists("course/data")) normalizePath(".", mustWork = TRUE),
    if (dir.exists("../course/data")) normalizePath("..", mustWork = TRUE),
    if (dir.exists("../../course/data")) normalizePath("../..", mustWork = TRUE)
  )
  root <- candidates[!sapply(candidates, is.null)][1]
  if (is.na(root)) {
    stop("Cannot find course/data — open this file from proxy-cfr-comparison/")
  }
  root
}

repo_root <- find_repo_root()
proj_root <- if (file.exists("analysis.qmd")) "." else "proxy-cfr-comparison"
data_dir  <- file.path(repo_root, "course", "data", "dod2k_v2.0")
cache_dir <- file.path(proj_root, "output", "cache")
cache_rds <- file.path(cache_dir, "dod2k_timeseries.rds")
dir.create(cache_dir, recursive = TRUE, showWarnings = FALSE)

year_min <- 0L   # reconstruction window (CE)
year_max <- 2000L
```

## Compact CSV loader

Three files are joined: metadata, year arrays, and proxy-value arrays.

```{r}
#| label: dod2k-loader
#| code-fold: true

#' Parse one line from a DoD2k compact array CSV (`id,val1,val2,…`).
#'
#' @param line One CSV line (without header).
#' @return List with `id` and numeric `values`.
parse_array_line <- function(line) {
  parts <- strsplit(line, ",", fixed = TRUE)[[1]]
  list(id = parts[[1]], values = as.numeric(parts[-1]))
}

#' Read a compact array column into a tibble.
#'
#' @param path Path to e.g. `dod2k_v2.0_compact_year.csv`.
#' @return Tibble: `dataset_id`, `values` (list of numeric vectors).
read_array_column <- function(path) {
  lines <- readLines(path, warn = FALSE)
  parsed <- lapply(lines[-1], parse_array_line)
  tibble(
    dataset_id = vapply(parsed, `[[`, character(1), "id"),
    values = lapply(parsed, `[[`, "values")
  )
}

#' Load DoD2k v2.0 compact bundle — one row per dataset, list-columns for time series.
#'
#' @return Wide tibble with metadata, `year`, and `paleoData_values`.
load_dod2k_compact <- function() {
  prefix <- basename(data_dir)
  meta <- read_csv(
    file.path(data_dir, paste0(prefix, "_compact_metadata.csv")),
    show_col_types = FALSE
  ) |>
    rename(dataset_id = any_of(c("datasetId", "...1")))

  years <- read_array_column(file.path(data_dir, paste0(prefix, "_compact_year.csv"))) |>
    rename(year = values)
  values <- read_array_column(
    file.path(data_dir, paste0(prefix, "_compact_paleoData_values.csv"))
  ) |>
    rename(paleoData_values = values)

  meta |>
    left_join(years, by = "dataset_id") |>
    left_join(values, by = "dataset_id")
}
```

## Long-format catalog

Expands list-columns to one row per `(dataset, year)` and caches the result as RDS.

```{r}
#| label: dod2k-catalog
#| code-fold: true

#' Convert compact data to a long catalog (0–2000 CE, finite values only).
#'
#' @param compact Output of [load_dod2k_compact()].
#' @return Tibble: `dataset_id`, `year`, `value`, `lon`, `lat`, `archive_type`, `proxy_type`.
compact_to_catalog <- function(compact) {
  rows <- vector("list", nrow(compact))
  n <- 0L
  for (i in seq_len(nrow(compact))) {
    rec <- compact[i, ]
    yrs  <- rec$year[[1]]
    vals <- rec$paleoData_values[[1]]
    if (!length(yrs) || !length(vals)) next

    ok <- yrs >= year_min & yrs <= year_max & is.finite(vals)
    if (!any(ok)) next

    n <- n + 1L
    rows[[n]] <- tibble(
      dataset_id = rec$dataset_id,
      year = as.integer(yrs[ok]),
      value = as.double(vals[ok]),
      lon = rec$geo_meanLon,
      lat = rec$geo_meanLat,
      archive_type = rec$archiveType,
      proxy_type = rec$paleoData_proxy
    )
  }
  if (!n) return(tibble())
  bind_rows(rows)
}

#' Read cached catalog from `output/cache/dod2k_timeseries.rds`.
#'
#' @return Tibble or `NULL` if cache does not exist.
read_catalog_cache <- function() {
  if (!file.exists(cache_rds)) return(NULL)
  readRDS(cache_rds) |>
    mutate(
      dataset_id = as.character(dataset_id),
      archive_type = as.character(archive_type),
      proxy_type = as.character(proxy_type),
      year = as.integer(year),
      value = as.double(value),
      lon = as.double(lon),
      lat = as.double(lat)
    )
}

#' Load or build the long-format catalog.
#'
#' @param refresh If `TRUE`, rebuild cache from CSV.
#' @return Long-format catalog tibble.
load_dod2k_catalog <- function(refresh = FALSE) {
  if (!refresh) {
    cached <- read_catalog_cache()
    if (!is.null(cached)) return(cached)
  }
  message("Building catalog from compact CSV …")
  catalog <- compact_to_catalog(load_dod2k_compact())
  saveRDS(catalog, cache_rds)
  catalog
}
```

## Archive summary

First run builds the cache; later runs read RDS directly.

```{r}
#| label: dod2k-load
#| cache: true

dod2k_compact <- load_dod2k_compact()
catalog       <- load_dod2k_catalog()

archive_summary <- dod2k_compact |>
  count(archiveType, paleoData_proxy, name = "n", sort = TRUE) |>
  rename(archive_type = archiveType, proxy_type = paleoData_proxy)

glue(
  "Records: {format(nrow(dod2k_compact), big.mark = ',')}  |  ",
  "Speleothems: {sum(dod2k_compact$archiveType == 'Speleothem', na.rm = TRUE)}  |  ",
  "δ18O: {sum(dod2k_compact$archiveType == 'Speleothem' & dod2k_compact$paleoData_proxy == 'd18O', na.rm = TRUE)}"
)

knitr::kable(head(archive_summary, 12), caption = "DoD2k archive × proxy (metadata)")
```

# Proxy types and spatial coverage

Where are speleothems, which of the four proxy types are available, and how well are series resolved in time?

## Styling

Shared colours, labels, and a consistent ggplot/plotly theme for all figures.

```{r}
#| label: speleothem-config
#| code-fold: true

proxy_order   <- c("d18O", "d13C", "growth rate", "Mg/Ca")
proxy_labels  <- c(d18O = "δ18O", d13C = "δ13C", `growth rate` = "growth rate", `Mg/Ca` = "Mg/Ca")
proxy_colors  <- c(d18O = "#b2182b", d13C = "#d6604d", `growth rate` = "#f4a582", `Mg/Ca` = "#92c5de")
proxy_symbols <- c(d18O = "triangle-up", d13C = "triangle-down", `growth rate` = "square", `Mg/Ca` = "circle")

# Line weights — thicker on time-series panels
line_lwd      <- 1.05
line_lwd_thin <- 0.35
point_sz      <- 2.0
point_sz_sm   <- 1.6

#' Left-aligned bold plotly title with optional subtitle.
plotly_title <- function(title, subtitle = NULL) {
  sub_txt <- if (!is.null(subtitle) && nzchar(subtitle)) {
    paste0("<br><span style='font-size:12px;color:#555555'>", subtitle, "</span>")
  } else ""
  list(
    text = paste0("<b style='font-size:17px'>", title, "</b>", sub_txt),
    x = 0.02, xanchor = "left",
    y = 0.98, yanchor = "top"
  )
}

#' Minimal ggplot theme: bold left title, subtitle guidance, grid on demand.
#'
#' @param grid `"none"`, `"x"`, `"y"`, or `"xy"` — major grid lines only where they aid reading.
theme_clean <- function(grid = "none") {
  grid <- match.arg(grid, c("none", "x", "y", "xy"))
  base <- theme_minimal(base_size = 11, base_family = "sans") +
    theme(
      plot.title = element_text(
        face = "bold", size = 15, hjust = 0,
        margin = margin(b = 4), colour = "grey10"
      ),
      plot.subtitle = element_text(
        size = 10.5, hjust = 0, colour = "grey35",
        margin = margin(b = 10), lineheight = 1.15
      ),
      plot.caption = element_text(
        size = 9, hjust = 0, colour = "grey45",
        margin = margin(t = 10), lineheight = 1.2
      ),
      plot.margin = margin(12, 16, 10, 10),
      legend.position = "bottom",
      legend.title = element_text(face = "bold", size = 10),
      legend.text = element_text(size = 9.5),
      legend.key.height = unit(0.45, "cm"),
      axis.title = element_text(size = 10, colour = "grey20"),
      axis.title.x = element_text(margin = margin(t = 8)),
      axis.title.y = element_text(margin = margin(r = 8)),
      axis.text = element_text(size = 9, colour = "grey30"),
      strip.text = element_text(face = "bold", size = 9.5, colour = "grey15"),
      strip.background = element_rect(fill = "grey96", colour = NA),
      panel.spacing = unit(1.1, "lines")
    )
  grid_theme <- switch(grid,
    none = theme(panel.grid = element_blank()),
    x = theme(
      panel.grid.major.x = element_line(colour = "grey91", linewidth = 0.35),
      panel.grid.major.y = element_blank(),
      panel.grid.minor = element_blank()
    ),
    y = theme(
      panel.grid.major.y = element_line(colour = "grey91", linewidth = 0.35),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank()
    ),
    xy = theme(
      panel.grid.major = element_line(colour = "grey91", linewidth = 0.35),
      panel.grid.minor = element_blank()
    )
  )
  base + grid_theme
}
```

## Site tables

**`n_metadata`** = rows in the archive; **`n_with_data`** = series with values in 0–2000 CE.

```{r}
#| label: speleothem-helpers
#| code-fold: true

#' One row per speleothem series with site metadata.
#'
#' @param catalog Long catalog from [load_dod2k_catalog()].
#' @param compact Wide compact table (for site names).
#' @return Tibble with `dataset_id`, coords, year range, `n_years`, `site`.
speleothem_sites <- function(catalog, compact) {
  sites <- catalog |>
    filter(archive_type == "Speleothem") |>
    group_by(dataset_id, proxy_type, lon, lat) |>
    summarize(
      year_start = min(year),
      year_end = max(year),
      n_years = n(),
      .groups = "drop"
    )

  meta <- compact |>
    filter(archiveType == "Speleothem") |>
    dplyr::select(dataset_id, geo_siteName) |>
    rename(site = geo_siteName)

  sites |>
    left_join(meta, by = "dataset_id") |>
    mutate(
      site = if_else(is.na(site) | site == "" | site == dataset_id, dataset_id, site),
      proxy_type = factor(proxy_type, levels = proxy_order)
    )
}

#' Count metadata rows vs. series with usable time series, by proxy type.
#'
#' @param compact Wide compact table.
#' @param catalog Long catalog.
#' @return Tibble: `proxy_label`, `proxy_type`, `n_metadata`, `n_with_data`.
speleothem_summary <- function(compact, catalog) {
  meta <- compact |>
    filter(archiveType == "Speleothem") |>
    count(paleoData_proxy, name = "n_metadata") |>
    rename(proxy_type = paleoData_proxy)

  with_data <- catalog |>
    filter(archive_type == "Speleothem") |>
    distinct(dataset_id, proxy_type) |>
    count(proxy_type, name = "n_with_data")

  meta |>
    full_join(with_data, by = "proxy_type") |>
    mutate(
      n_metadata = replace_na(n_metadata, 0L),
      n_with_data = replace_na(n_with_data, 0L),
      proxy_label = proxy_labels[proxy_type],
      proxy_type = factor(proxy_type, levels = proxy_order)
    ) |>
    arrange(proxy_type) |>
    transmute(proxy_label, proxy_type, n_metadata, n_with_data)
}
```

```{r}
#| label: speleothem-summary
#| cache: true
#| dependson: dod2k-load

speleo_sites <- speleothem_sites(catalog, dod2k_compact)

speleothem_summary(dod2k_compact, catalog) |>
  knitr::kable(caption = "Speleothem proxy types")
```

## Site map

Interactive world map — colour and symbol by proxy type; hover shows site name and year range.

```{r}
#| label: speleothem-map-fn
#| code-fold: true

#' Interactive plotly map of speleothem sites.
#'
#' @param sites Output of [speleothem_sites()].
#' @return plotly object.
plot_speleothem_map <- function(sites) {
  sites <- sites |>
    mutate(
      proxy_label = proxy_labels[as.character(proxy_type)],
      hover = paste0(
        "<b>", site, "</b><br>", dataset_id, "<br>",
        "Proxy: ", proxy_label, "<br>",
        "Years: ", year_start, "–", year_end, " (n = ", n_years, ")"
      )
    )

  p <- plot_ly(type = "scattergeo", mode = "markers")
  for (px in proxy_order) {
    sub <- filter(sites, as.character(proxy_type) == px)
    if (!nrow(sub)) next
    p <- add_trace(
      p, data = sub, lon = ~lon, lat = ~lat, text = ~hover, hoverinfo = "text",
      name = paste0("Speleothem: ", proxy_labels[[px]], " (n=", nrow(sub), ")"),
      marker = list(
        symbol = proxy_symbols[[px]], size = 9,
        color = proxy_colors[[px]], line = list(width = 0.6, color = "#333")
      )
    )
  }
  p |> layout(
    title = plotly_title(
      "DoD2k speleothem archive",
      paste0(nrow(sites), " series — symbol = proxy type; hover for site name and year range")
    ),
    margin = list(t = 72, b = 0, l = 0, r = 0),
    geo = list(
      projection = list(type = "natural earth"),
      showland = TRUE, landcolor = "#f0f0f0",
      showocean = TRUE, oceancolor = "#e6f2fa",
      showcountries = TRUE, countrycolor = "#bbbbbb",
      lataxis = list(range = c(-40, 75))
    ),
    legend = list(orientation = "h", y = -0.05, x = 0.02, xanchor = "left")
  )
}
```

```{r}
#| label: speleothem-map
#| fig-cap: "Global distribution of DoD2k speleothem series. Compare proxy-type clusters with regions used in Steps 2–3."
#| dependson: speleothem-summary

plot_speleothem_map(speleo_sites)
```

## Temporal coverage

How many series have at least one dated value per 100-year bin (actual years, not calendar span).

```{r}
#| label: speleothem-density-fn
#| code-fold: true

#' Stacked bar chart — series count per time bin.
#'
#' @param catalog Long-format catalog.
#' @param bin_width Bin width in years (default 100).
#' @return ggplot object.
plot_speleothem_density <- function(catalog, bin_width = 100) {
  density <- catalog |>
    filter(archive_type == "Speleothem") |>
    mutate(bin_start = (as.integer(year %/% bin_width)) * bin_width) |>
    distinct(dataset_id, proxy_type, bin_start) |>
    count(bin_start, proxy_type, name = "n_series") |>
    mutate(
      proxy_type = factor(proxy_type, levels = proxy_order),
      bin_mid = bin_start + bin_width / 2
    )

  ggplot(density, aes(bin_mid, n_series, fill = proxy_type)) +
    geom_col(width = bin_width * 0.9, colour = "white", linewidth = 0.2) +
    scale_fill_manual(values = proxy_colors, labels = proxy_labels[proxy_order], name = "Proxy type") +
    scale_x_continuous(limits = c(0, 2000), breaks = seq(0, 2000, 250), name = "Year (CE)") +
    labs(
      title = "Where observations exist",
      subtitle = "Series with at least one dated value per 100-year bin, stacked by proxy type",
      caption = "Watch for: pre-1600 gaps and whether δ18O dominates recent centuries.",
      y = "Series with data"
    ) +
    theme_clean(grid = "y")
}
```

```{r}
#| label: speleothem-density
#| fig-cap: "Temporal coverage by century. Tall stacks in a bin mean many caves contributed dated values in that period."
#| dependson: speleothem-summary

plot_speleothem_density(catalog)
```

## Temporal resolution

Number of dated values per series — more informative than calendar first/last year.

```{r}
#| label: speleothem-resolution-fn
#| code-fold: true

#' Boxplot of observation counts per series.
#'
#' @param sites Output of [speleothem_sites()].
#' @return ggplot object.
plot_speleothem_resolution <- function(sites) {
  ggplot(sites, aes(proxy_type, n_years, fill = proxy_type)) +
    geom_boxplot(outlier.size = 0.8, alpha = 0.85) +
    scale_fill_manual(values = proxy_colors, guide = "none") +
    scale_x_discrete(labels = proxy_labels[proxy_order]) +
    labs(
      title = "Temporal resolution per series",
      subtitle = "Number of dated values — more informative than calendar first/last year",
      caption = "Watch for: long spans with few points (low effective resolution).",
      x = NULL, y = "Observations per series"
    ) +
    theme_clean(grid = "y")
}
```

```{r}
#| label: speleothem-resolution
#| fig-cap: "Resolution per proxy type. Compare median (thick line) across δ18O, δ13C, growth rate, and Mg/Ca."
#| dependson: speleothem-summary

plot_speleothem_resolution(speleo_sites)
```

# Screen against instrumental climate

Each speleothem series is compared to the **nearest CRU grid box** (0.5°): annual mean temperature, JJA temperature, and annual precipitation [@harris2020]. We use **Pearson correlation** on detrended annual values (1901–2000). A series needs at least 10 overlapping years.

## CRU grid extraction

```{r}
#| label: cru-setup
#| code-fold: true

cru_dir  <- file.path(cache_dir, "cru_ts")
cru_base <- "https://crudata.uea.ac.uk/cru/data/hrg/cru_ts_4.07/cruts.2304141047.v4.07"
cru_years <- 1901:2022
cru_rds   <- file.path(cache_dir, "cru_site_climate.rds")
dir.create(cru_dir, recursive = TRUE, showWarnings = FALSE)

#' Download and decompress one CRU TS variable if needed.
#'
#' @param var `"tmp"` or `"pre"`.
#' @return Path to the uncompressed NetCDF file.
ensure_cru_nc <- function(var) {
  fname <- paste0("cru_ts4.07.1901.2022.", var, ".dat.nc")
  nc_path <- file.path(cru_dir, fname)
  gz_path <- paste0(nc_path, ".gz")
  if (!file.exists(nc_path)) {
    if (!file.exists(gz_path)) {
      message("Downloading CRU TS ", var, " …")
      url <- paste(cru_base, var, paste0(fname, ".gz"), sep = "/")
      if (nzchar(Sys.which("curl"))) {
        status <- system2("curl", c("-L", "-C", "-", "--retry", "3", "--max-time", "900", "-o", gz_path, url))
        if (status != 0) stop("CRU download failed for ", var)
      } else {
        utils::download.file(url, gz_path, mode = "wb", timeout = 900)
      }
    }
    message("Decompressing ", basename(gz_path), " …")
    status <- system2("gunzip", c("-k", "-f", gz_path))
    if (status != 0) stop("gunzip failed — decompress ", gz_path, " manually")
  }
  nc_path
}

#' Aggregate monthly CRU matrices to annual temperature means and precipitation sums.
cru_monthly_to_annual <- function(mat_tmp, mat_pre, years = cru_years) {
  n_pt <- nrow(mat_tmp)
  n_yr <- length(years)
  tmp_ann <- tmp_jja <- pre_ann <- matrix(NA_real_, n_pt, n_yr)
  for (j in seq_len(n_yr)) {
    m <- (j - 1L) * 12L + 1:12
    tmp_ann[, j] <- rowMeans(mat_tmp[, m, drop = FALSE], na.rm = TRUE)
    tmp_jja[, j] <- rowMeans(mat_tmp[, m[6:8], drop = FALSE], na.rm = TRUE)
    pre_ann[, j] <- rowSums(mat_pre[, m, drop = FALSE], na.rm = TRUE)
  }
  list(tmp_ann = tmp_ann, tmp_jja = tmp_jja, pre_ann = pre_ann)
}

#' Extract CRU TS climate at speleothem locations (cached as RDS).
#'
#' @param loc Tibble with `loc_id`, `lon`, `lat`.
#' @param refresh Rebuild even when RDS exists.
#' @return Long tibble: `loc_id`, `lon`, `lat`, `year`, `tmp_ann`, `tmp_jja`, `pre_ann`.
load_cru_site_climate <- function(loc, refresh = FALSE) {
  if (!refresh && file.exists(cru_rds)) return(readRDS(cru_rds))

  pts <- vect(loc, geom = c("lon", "lat"), crs = "EPSG:4326")
  tmp_nc <- ensure_cru_nc("tmp")
  pre_nc <- ensure_cru_nc("pre")

  mat_tmp <- as.matrix(
    terra::extract(rast(tmp_nc, subds = "tmp"), pts, method = "simple")[, -1]
  )
  mat_pre <- as.matrix(
    terra::extract(rast(pre_nc, subds = "pre"), pts, method = "simple")[, -1]
  )
  ann <- cru_monthly_to_annual(mat_tmp, mat_pre)

  clim <- map_dfr(seq_len(nrow(loc)), \(i) {
    tibble(
      loc_id = loc$loc_id[i], lon = loc$lon[i], lat = loc$lat[i],
      year = cru_years,
      tmp_ann = ann$tmp_ann[i, ], tmp_jja = ann$tmp_jja[i, ], pre_ann = ann$pre_ann[i, ]
    )
  })
  saveRDS(clim, cru_rds)
  clim
}
```

## Correlation functions

Pearson correlation with two-tailed p-value; series with fewer than 10 overlap years are skipped.

```{r}
#| label: corr-setup
#| code-fold: true
#| cache: true

instr_y0 <- 1901L
instr_y1 <- 2000L
min_overlap <- 10L

climate_vars <- c(
  r_tmp_ann = "Temperature (annual)",
  r_tmp_jja = "Temperature (JJA)",
  r_pre_ann = "Precipitation (annual)"
)

#' Correlate one proxy series with one climate variable.
corr_series <- function(proxy_df, clim_df, clim_col,
                        y0 = instr_y0, y1 = instr_y1, min_n = min_overlap) {
  d <- proxy_df |>
    filter(year >= y0, year <= y1) |>
    inner_join(clim_df, by = "year")
  ok <- is.finite(d$value) & is.finite(d[[clim_col]])
  d <- d[ok, , drop = FALSE]
  if (nrow(d) < min_n) {
    return(tibble(r = NA_real_, p = NA_real_, n = nrow(d)))
  }
  ct <- cor.test(d$value, d[[clim_col]], method = "pearson")
  tibble(r = unname(ct$estimate), p = ct$p.value, n = nrow(d))
}

#' Build annual speleothem series (mean if multiple values per year).
speleothem_annual <- function(catalog) {
  catalog |>
    filter(archive_type == "Speleothem") |>
    group_by(dataset_id, proxy_type, lon, lat, year) |>
    summarize(value = mean(value), .groups = "drop")
}

#' Correlate all speleothem series with local CRU grid-box climate.
compute_speleo_correlations <- function(catalog, clim, sites,
                                        y0 = instr_y0, y1 = instr_y1) {
  loc <- sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
  proxy_ann <- speleothem_annual(catalog) |> inner_join(loc, by = c("lon", "lat"))

  res <- map_dfr(group_split(proxy_ann, dataset_id), function(df) {
    cdf <- clim |> filter(loc_id == df$loc_id[1])
    t1 <- corr_series(df, cdf, "tmp_ann", y0, y1)
    t2 <- corr_series(df, cdf, "tmp_jja", y0, y1)
    t3 <- corr_series(df, cdf, "pre_ann", y0, y1)
    tibble(
      dataset_id = df$dataset_id[1],
      proxy_type = df$proxy_type[1],
      lon = df$lon[1], lat = df$lat[1],
      n_overlap = t1$n,
      r_tmp_ann = t1$r, p_tmp_ann = t1$p,
      r_tmp_jja = t2$r, p_tmp_jja = t2$p,
      r_pre_ann = t3$r, p_pre_ann = t3$p
    )
  })

  res |>
    rowwise() |>
    mutate(
      best_var = {
        rs <- c(tmp_ann = r_tmp_ann, tmp_jja = r_tmp_jja, pre_ann = r_pre_ann)
        if (all(is.na(rs))) NA_character_ else names(rs)[which.max(abs(rs))]
      },
      best_r = {
        rs <- c(r_tmp_ann, r_tmp_jja, r_pre_ann)
        if (all(is.na(rs))) NA_real_ else rs[which.max(abs(rs))]
      }
    ) |>
    ungroup()
}

#' Add display labels to correlation results.
enrich_correlations <- function(corr) {
  corr |>
    mutate(
      proxy_type = factor(proxy_type, levels = proxy_order),
      proxy_label = proxy_labels[as.character(proxy_type)],
      best_var_label = recode(
        best_var,
        tmp_ann = "Temperature (annual)",
        tmp_jja = "Temperature (JJA)",
        pre_ann = "Precipitation (annual)",
        .default = NA_character_
      )
    )
}
```

## All stations

```{r}
#| label: corr-run
#| cache: true
#| dependson: speleothem-summary

corr_rds <- file.path(cache_dir, "speleo_instrument_correlations.rds")
cru_locs <- speleo_sites |> distinct(lon, lat) |> mutate(loc_id = row_number())

if (file.exists(corr_rds) && file.exists(cru_rds)) {
  speleo_corr <- enrich_correlations(readRDS(corr_rds))
} else {
  cru_clim <- load_cru_site_climate(cru_locs)
  speleo_corr <- enrich_correlations(compute_speleo_correlations(catalog, cru_clim, speleo_sites))
  saveRDS(speleo_corr, corr_rds)
}

n_tested <- sum(is.finite(speleo_corr$best_r))
n_sig <- speleo_corr |>
  transmute(sig = p_tmp_ann < 0.05 | p_tmp_jja < 0.05 | p_pre_ann < 0.05) |>
  pull(sig) |>
  sum(na.rm = TRUE)
glue(
  "Series tested (≥{min_overlap} overlap years, {instr_y0}–{instr_y1}): {n_tested}  |  ",
  "Significant (p < 0.05, any variable): {n_sig}"
)
```

```{r}
#| label: corr-summary-table
#| dependson: corr-run

speleo_corr |>
  filter(is.finite(r_tmp_ann) | is.finite(r_pre_ann)) |>
  group_by(proxy_type) |>
  summarize(
    n = n(),
    median_r_tmp = median(r_tmp_ann, na.rm = TRUE),
    median_r_pre = median(r_pre_ann, na.rm = TRUE),
    sig_tmp = sum(p_tmp_ann < 0.05, na.rm = TRUE),
    sig_pre = sum(p_pre_ann < 0.05, na.rm = TRUE),
    .groups = "drop"
  ) |>
  mutate(proxy_label = proxy_labels[as.character(proxy_type)]) |>
  transmute(
    proxy_label, n, median_r_tmp, median_r_pre, sig_tmp, sig_pre
  ) |>
  knitr::kable(
    digits = 3,
    caption = paste0("Local CRU correlations by proxy type (", instr_y0, "–", instr_y1, ")")
  )
```

Distribution of correlation coefficients across all series — weak median correlations are typical for heterogeneous archives (cf. Exercise 12).

```{r}
#| label: corr-histogram-fn
#| code-fold: true
#| cache: true

#' Histogram of correlation coefficients, faceted by climate variable.
plot_correlation_histogram <- function(corr) {
  long <- corr |>
    filter(is.finite(r_tmp_ann) | is.finite(r_tmp_jja) | is.finite(r_pre_ann)) |>
    pivot_longer(
      c(r_tmp_ann, r_tmp_jja, r_pre_ann),
      names_to = "climate", values_to = "r"
    ) |>
    mutate(
      climate = factor(climate, levels = names(climate_vars), labels = climate_vars),
      proxy_type = factor(proxy_type, levels = proxy_order)
    )

  ggplot(long, aes(r, fill = proxy_type)) +
    geom_histogram(bins = 30, alpha = 0.75, position = "identity") +
    facet_wrap(~climate, ncol = 1, scales = "free_y") +
    geom_vline(xintercept = 0, linetype = "dashed", colour = "grey50") +
    scale_fill_manual(values = proxy_colors, labels = proxy_labels[proxy_order], name = "Proxy type") +
    labs(
      title = "Proxy–climate correlations (all speleothem series)",
      subtitle = paste0("Local CRU grid box, ", instr_y0, "–", instr_y1),
      x = "Pearson r", y = "Count"
    ) +
    theme_clean(grid = "y")
}
```

```{r}
#| label: corr-histogram
#| fig-cap: "Distribution of Pearson r for each climate target. Peaks near zero suggest weak local screening signal overall."
#| dependson: corr-run

plot_correlation_histogram(speleo_corr)
```

## By proxy type

Compare temperature vs. precipitation fit and count significant correlations (p < 0.05).

```{r}
#| label: corr-by-type-fn
#| code-fold: true
#| cache: true

#' Boxplot of correlations by proxy type and climate variable.
plot_correlation_by_type <- function(corr) {
  long <- corr |>
    pivot_longer(
      c(r_tmp_ann, r_tmp_jja, r_pre_ann),
      names_to = "climate", values_to = "r"
    ) |>
    mutate(
      climate = factor(climate, levels = names(climate_vars), labels = climate_vars),
      proxy_type = factor(proxy_type, levels = proxy_order)
    )

  ggplot(long, aes(proxy_type, r, fill = proxy_type)) +
    geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
    geom_boxplot(outlier.size = 0.7, alpha = 0.85) +
    facet_wrap(~climate, ncol = 1) +
    scale_fill_manual(values = proxy_colors, guide = "none") +
    scale_x_discrete(labels = proxy_labels[proxy_order]) +
    labs(
      title = "Correlation by proxy type",
      subtitle = "Spread and median r per proxy — compare temperature vs. precipitation panels",
      caption = "Boxes near zero across types → no single proxy consistently tracks local CRU.",
      x = NULL, y = "Pearson r"
    ) +
    theme_clean(grid = "y")
}

#' Count significant correlations (p < 0.05) by proxy type and climate variable.
plot_significance_counts <- function(corr) {
  sig <- corr |>
    transmute(
      proxy_type,
      tmp_ann = p_tmp_ann < 0.05,
      tmp_jja = p_tmp_jja < 0.05,
      pre_ann = p_pre_ann < 0.05
    ) |>
    pivot_longer(-proxy_type, names_to = "climate", values_to = "sig") |>
    filter(!is.na(sig)) |>
    group_by(proxy_type, climate) |>
    summarize(n_sig = sum(sig), .groups = "drop") |>
    mutate(
      climate = factor(climate, levels = names(climate_vars), labels = climate_vars),
      proxy_type = factor(proxy_type, levels = proxy_order)
    )

  ggplot(sig, aes(climate, n_sig, fill = proxy_type)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    scale_fill_manual(values = proxy_colors, labels = proxy_labels[proxy_order], name = "Proxy type") +
    labs(
      title = "Significant correlations (p < 0.05)",
      subtitle = "Count of series passing local significance test per proxy type",
      caption = "Low counts are expected — screening is strict and speleothems are heterogeneous.",
      x = NULL, y = "Number of series"
    ) +
    theme_clean(grid = "y")
}
```

```{r}
#| label: corr-by-type
#| fig-cap: "Correlations and significance by proxy type — compare spread across the three climate targets."
#| dependson: corr-run

plot_correlation_by_type(speleo_corr)
```

```{r}
#| label: corr-significance
#| fig-cap: "Number of series with p < 0.05 per proxy type — a strict local screening threshold."
#| dependson: corr-run

plot_significance_counts(speleo_corr)
```

## Best-fit climate variable

For each series, which of the three climate targets has the highest |r|?

```{r}
#| label: corr-best-var-fn
#| code-fold: true
#| cache: true

plot_best_variable_counts <- function(corr) {
  best <- corr |>
    filter(!is.na(best_var)) |>
    count(proxy_type, best_var_label, name = "n") |>
    mutate(proxy_type = factor(proxy_type, levels = proxy_order))

  ggplot(best, aes(best_var_label, n, fill = proxy_type)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.7) +
    scale_fill_manual(values = proxy_colors, labels = proxy_labels[proxy_order], name = "Proxy type") +
    labs(
      title = "Strongest |r| per series",
      subtitle = "Which climate variable (T annual, T JJA, precipitation) fits best locally?",
      caption = "Precipitation bars dominating → rainfall-sensitive proxies for Step 2/3.",
      x = NULL, y = "Number of series"
    ) +
    theme_clean(grid = "y") +
    theme(axis.text.x = element_text(angle = 20, hjust = 1))
}
```

```{r}
#| label: corr-best-var
#| fig-cap: "Which climate variable wins the |r| contest per series — guides target choice in Step 2."
#| dependson: corr-run

plot_best_variable_counts(speleo_corr)
```

## Correlation map

Points coloured by the strongest local correlation (signed).

```{r}
#| label: corr-map-fn
#| code-fold: true
#| cache: true

#' Map speleothem sites coloured by best correlation with CRU climate.
plot_correlation_map <- function(corr) {
  map_df <- corr |>
    filter(is.finite(best_r)) |>
    left_join(
      speleo_sites |> dplyr::select(dataset_id, site),
      by = "dataset_id"
    ) |>
    mutate(
      hover = paste0(
        "<b>", site, "</b><br>", dataset_id, "<br>",
        proxy_labels[as.character(proxy_type)], "<br>",
        "Best: ", best_var_label, "<br>",
        "r = ", round(best_r, 3), " (n = ", n_overlap, ")"
      )
    )

  map_df |>
    plot_ly(
      type = "scattergeo", mode = "markers",
      lon = ~lon, lat = ~lat, text = ~hover, hoverinfo = "text",
      color = ~best_r, colors = c("#2166ac", "#f7f7f7", "#b2182b"),
      marker = list(size = 7, line = list(width = 0.4, color = "#333"))
    ) |>
    layout(
      title = plotly_title(
        "Best proxy–climate correlation",
        paste0("Signed r at nearest CRU grid cell, ", instr_y0, "–", instr_y1, " — red = positive, blue = negative")
      ),
      margin = list(t = 72, b = 0, l = 0, r = 0),
      geo = list(
        projection = list(type = "natural earth"),
        showland = TRUE, landcolor = "#f0f0f0",
        showocean = TRUE, oceancolor = "#e6f2fa",
        showcountries = TRUE, countrycolor = "#bbbbbb"
      )
    )
}
```

```{r}
#| label: corr-map
#| fig-cap: "Geographic pattern of best local fit. Strong regional clusters may reflect real hydroclimate — or archive bias."
#| dependson: [corr-run, speleothem-summary]

plot_correlation_map(speleo_corr)
```

## Example fits

One well-correlated series per proxy type — proxy vs. climate scatter for the variable with highest |r|.

```{r}
#| label: corr-examples-fn
#| code-fold: true

#' Scatter plots for the best-correlated series per proxy type.
plot_correlation_examples <- function(corr, catalog, clim) {
  loc <- speleo_sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
  proxy_ann <- speleothem_annual(catalog) |> inner_join(loc, by = c("lon", "lat"))

  picks <- corr |>
    filter(is.finite(best_r)) |>
    group_by(proxy_type) |>
    slice_max(abs(best_r), n = 1, with_ties = FALSE) |>
    ungroup()

  clim_col <- c(tmp_ann = "tmp_ann", tmp_jja = "tmp_jja", pre_ann = "pre_ann")

  examples <- map_dfr(seq_len(nrow(picks)), function(i) {
    row <- picks[i, ]
    px <- proxy_ann |> filter(dataset_id == row$dataset_id)
    cd <- clim |>
      filter(loc_id == px$loc_id[1]) |>
      dplyr::select(year, climate = all_of(clim_col[[row$best_var]]))
    px |>
      filter(year >= instr_y0, year <= instr_y1) |>
      inner_join(cd, by = "year") |>
      mutate(
        dataset_id = row$dataset_id,
        proxy_label = proxy_labels[as.character(row$proxy_type)],
        climate_label = row$best_var_label,
        r = row$best_r
      )
  })

  ggplot(examples, aes(climate, value)) +
    geom_point(alpha = 0.6, size = point_sz_sm, colour = "#444444") +
    geom_smooth(method = "lm", se = TRUE, colour = "#b2182b", linewidth = line_lwd) +
    facet_wrap(~paste0(proxy_label, " — ", dataset_id, "\nr = ", round(r, 2)),
               scales = "free", ncol = 2) +
    labs(
      title = "Best local fits per proxy type",
      subtitle = paste0("Strongest |r| series per proxy vs. CRU TS (", instr_y0, "–", instr_y1, ")"),
      caption = "Slope direction is site-specific — do not generalise across the archive.",
      x = "Climate (CRU TS)", y = "Proxy value"
    ) +
    theme_clean(grid = "y") +
    theme(strip.text = element_text(size = 8.5, face = "bold"))
}
```

```{r}
#| label: corr-examples
#| fig-cap: "Illustrative scatter plots — one showcase series per proxy type. Grey band = linear fit ± SE."
#| dependson: corr-run

if (!exists("cru_clim")) cru_clim <- readRDS(cru_rds)
plot_correlation_examples(speleo_corr, catalog, cru_clim)
```

## Step 1 takeaways

```{r}
#| label: step1-takeaways-prep
#| include: false
#| dependson: corr-run

median_r_tmp <- median(speleo_corr$r_tmp_ann, na.rm = TRUE)
median_r_pre <- median(speleo_corr$r_pre_ann, na.rm = TRUE)

best_target <- speleo_corr |>
  filter(!is.na(best_var)) |>
  count(proxy_type, best_var, name = "n") |>
  group_by(proxy_type) |>
  slice_max(n, n = 1, with_ties = FALSE) |>
  ungroup() |>
  mutate(
    proxy_label = proxy_labels[as.character(proxy_type)],
    target = recode(best_var,
      tmp_ann = "annual temperature",
      tmp_jja = "JJA temperature",
      pre_ann = "annual precipitation"
    )
  )

step1_proxy_bullets <- best_target |>
  mutate(line = glue("- **{proxy_label}**: strongest local fits most often with *{target}* ({n} series)")) |>
  pull(line) |>
  paste(collapse = "\n\n")
```

Across `r n_tested` series (`r instr_y0`–`r instr_y1`), median local correlations stay close to zero (temperature *r* ≈ `r sprintf("%.2f", median_r_tmp)`; precipitation *r* ≈ `r sprintf("%.2f", median_r_pre)`). Only `r n_sig` series reach p < 0.05.

Most common best-fit target per proxy type:

`r step1_proxy_bullets`

**Takeaway for Step 2:** screen locally, combine regionally — do not average all speleothems globally.

# Step 2 — Regional composite reconstruction

We group caves into coarse regions, keep only series whose local correlation passes p < 0.05, **detrend** each series, **calibrate** it to CRU on 1901–1980, and check whether a regional average still matches instruments on **1981–2000** (validation — data the model did not see during calibration).

## Regional bins and screening

```{r}
#| label: recon-setup
#| code-fold: true
#| cache: true

cal_y0 <- 1901L
cal_y1 <- 1980L
val_y0 <- 1981L
val_y1 <- 2000L
min_screen_n <- 3L   # minimum screened series per region × target
screen_p <- 0.05

climate_col <- c(tmp_ann = "tmp_ann", tmp_jja = "tmp_jja", pre_ann = "pre_ann")
p_col <- c(tmp_ann = "p_tmp_ann", tmp_jja = "p_tmp_jja", pre_ann = "p_pre_ann")
r_col <- c(tmp_ann = "r_tmp_ann", tmp_jja = "r_tmp_jja", pre_ann = "r_pre_ann")

#' Assign coarse continent/region from coordinates (speleothem-focused bins).
assign_region <- function(lon, lat) {
  case_when(
    lat < -10 & lon > 100 ~ "Australasia",
    lat <= 15 & lon >= -85 & lon < -30 ~ "South America",
    lat > 15 & lon < -30 ~ "Americas (north)",
    lat >= 35 & lon >= -15 & lon < 45 ~ "Europe",
    lat >= -35 & lat < 40 & lon >= -20 & lon < 55 ~ "Africa / Med",
    lon >= 50 ~ "Asia",
    TRUE ~ "Other"
  )
}

#' Remove linear trend (preserve mean).
detrend_linear <- function(years, values) {
  ok <- is.finite(values)
  if (sum(ok) < 5L) return(rep(NA_real_, length(values)))
  fit <- lm(values[ok] ~ years[ok])
  resid <- values - predict(fit, newdata = data.frame(years = years))
  resid + mean(values[ok])
}

#' Z-score using reference years only.
zscore_ref <- function(x, years, ref_y0, ref_y1) {
  ref <- x[years >= ref_y0 & years <= ref_y1 & is.finite(x)]
  if (length(ref) < 5L) return(rep(NA_real_, length(x)))
  s <- sd(ref)
  if (!is.finite(s) || s < 1e-8) return(rep(0, length(x)))
  (x - mean(ref)) / s
}

#' Screen series: significant p for assigned best climate variable.
screen_correlations <- function(corr, p_cut = screen_p) {
  corr |>
    mutate(
      region = assign_region(lon, lat),
      p_best = case_when(
        best_var == "tmp_ann" ~ p_tmp_ann,
        best_var == "tmp_jja" ~ p_tmp_jja,
        best_var == "pre_ann" ~ p_pre_ann,
        TRUE ~ NA_real_
      ),
      screened = is.finite(best_r) & !is.na(p_best) & p_best < p_cut & n_overlap >= min_overlap
    )
}

#' Calibrate one site; return annual recon + target in z-space.
calibrate_one_site <- function(proxy_df, clim_df, climate_var,
                               cal_y0, cal_y1, val_y0, val_y1) {
  col <- as.character(climate_col[[as.character(climate_var)]])
  d <- proxy_df |>
    inner_join(
      clim_df |> dplyr::transmute(year, climate = .data[[col]]),
      by = "year"
    ) |>
    filter(year >= cal_y0, year <= val_y1)

  if (nrow(d) < min_overlap) return(NULL)

  d <- d |>
    mutate(
      proxy_dt = detrend_linear(year, value),
      proxy_z = zscore_ref(proxy_dt, year, cal_y0, cal_y1),
      climate_z = zscore_ref(climate, year, cal_y0, cal_y1)
    )

  cal <- d |> filter(year >= cal_y0, year <= cal_y1)
  ok <- complete.cases(cal[, c("proxy_z", "climate_z")])
  if (sum(ok) < min_overlap) return(NULL)

  fit <- lm(climate_z ~ proxy_z, data = cal[ok, ])
  d <- d |>
    mutate(recon_z = predict(fit, newdata = d))

  list(
    data = d |> dplyr::select(year, proxy_z, climate_z, recon_z),
    coef = coef(fit)
  )
}

#' Safe Pearson r (returns NA if fewer than 3 pairs).
safe_cor <- function(x, y) {
  ok <- is.finite(x) & is.finite(y)
  if (sum(ok) < 3L) return(NA_real_)
  cor(x[ok], y[ok])
}

#' Build regional composites for each (region, climate_var) with enough screened series.
build_regional_reconstructions <- function(corr, catalog, clim, sites) {
  loc <- sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
  proxy_ann <- speleothem_annual(catalog) |> inner_join(loc, by = c("lon", "lat"))
  screened <- corr |> filter(screened)

  combos <- screened |>
    count(region, best_var, name = "n") |>
    filter(n >= min_screen_n, !is.na(best_var))

  if (!nrow(combos)) return(tibble())

  bind_rows(compact(map(seq_len(nrow(combos)), function(i) {
    reg <- combos$region[i]
    cv  <- combos$best_var[i]
    sub <- screened |> filter(region == reg, best_var == cv)

    site_recons <- compact(map(sub$dataset_id, function(id) {
      row <- sub |> filter(dataset_id == id)
      px <- proxy_ann |> filter(dataset_id == id)
      cdf <- clim |> filter(loc_id == px$loc_id[1])
      out <- calibrate_one_site(px, cdf, cv, cal_y0, cal_y1, val_y0, val_y1)
      if (is.null(out)) return(NULL)
      out$data |> mutate(
        dataset_id = id,
        sign_r = sign(row[[r_col[[cv]]]][1])
      )
    }))
    if (!length(site_recons)) return(NULL)

    all_years <- bind_rows(site_recons) |>
      mutate(recon_z_signed = if_else(sign_r < 0, -recon_z, recon_z))

    # Regional CRU target: mean across sites, then one z-score (not mean of site z-scores).
    col <- as.character(climate_col[[as.character(cv)]])
    regional_clim <- map_dfr(sub$dataset_id, function(id) {
      px <- proxy_ann |> filter(dataset_id == id) |> slice(1)
      clim |>
        filter(loc_id == px$loc_id) |>
        dplyr::transmute(year, climate = .data[[col]])
    }) |>
      group_by(year) |>
      summarize(climate = mean(climate, na.rm = TRUE), .groups = "drop") |>
      mutate(climate_z = zscore_ref(climate, year, cal_y0, cal_y1))

    composite <- all_years |>
      group_by(year) |>
      summarize(recon_z = mean(recon_z_signed, na.rm = TRUE), .groups = "drop") |>
      left_join(regional_clim |> dplyr::select(year, climate_z), by = "year")

    cal_idx <- composite$year >= cal_y0 & composite$year <= cal_y1
    val_idx <- composite$year >= val_y0 & composite$year <= val_y1

    tibble(
      region = reg,
      climate_var = cv,
      climate_label = recode(cv,
        tmp_ann = "Temperature (annual)",
        tmp_jja = "Temperature (JJA)",
        pre_ann = "Precipitation (annual)"
      ),
      n_series = nrow(sub),
      r_cal = safe_cor(composite$climate_z[cal_idx], composite$recon_z[cal_idx]),
      r_val = safe_cor(composite$climate_z[val_idx], composite$recon_z[val_idx]),
      rmse_cal = sqrt(mean((composite$climate_z[cal_idx] - composite$recon_z[cal_idx])^2, na.rm = TRUE)),
      rmse_val = sqrt(mean((composite$climate_z[val_idx] - composite$recon_z[val_idx])^2, na.rm = TRUE)),
      composite = list(composite),
      site_ids = list(sub$dataset_id)
    )
  })))
}

#' Bar chart of calibration vs validation correlation by region.
plot_recon_skill <- function(recon_summary) {
  long <- recon_summary |>
    dplyr::select(region, climate_label, n_series, r_cal, r_val) |>
    pivot_longer(c(r_cal, r_val), names_to = "period", values_to = "r") |>
    mutate(
      period = factor(
        period,
        levels = c("r_cal", "r_val"),
        labels = c("Calibration (1901–1980)", "Validation (1981–2000)")
      ),
      label = paste0(region, " — ", climate_label, " (n=", n_series, ")")
    ) |>
    mutate(label = fct_reorder(label, r, .fun = max, .desc = TRUE))

  ggplot(long, aes(r, label, fill = period)) +
    geom_col(position = position_dodge(width = 0.75), width = 0.65) +
    geom_vline(xintercept = 0, linewidth = line_lwd_thin, colour = "grey40") +
    scale_fill_manual(values = c("#4393c3", "#d6604d"), name = NULL) +
    labs(
      title = "Regional composite skill",
      subtitle = "Blue = calibration (1901–1980); red = validation (1981–2000)",
      caption = "Similar bar lengths → composite generalises; red near zero → poor out-of-sample skill.",
      x = "Pearson r", y = NULL
    ) +
    theme_clean(grid = "x") +
    theme(legend.position = "bottom")
}

#' Time series of regional composite vs regional CRU mean (z-scores).
plot_recon_timeseries <- function(recon_summary, focus = c("validation", "full")) {
  focus <- match.arg(focus)
  ts <- recon_summary |>
    dplyr::select(region, climate_label, n_series, composite) |>
    unnest(composite) |>
    mutate(
      panel = paste0(region, " — ", climate_label, " (n = ", n_series, ")"),
      panel = fct_inorder(panel)
    ) |>
    pivot_longer(c(climate_z, recon_z), names_to = "series", values_to = "z") |>
    mutate(series = recode(series,
      climate_z = "CRU regional mean",
      recon_z = "Speleothem composite"
    ))

  if (focus == "validation") {
    ts <- ts |> filter(year >= val_y0, year <= val_y1)
    title <- "Validation: composite vs CRU"
    subtitle <- paste0(
      val_y0, "–", val_y1, " — blue = CRU regional mean, red = speleothem composite (independent period)"
    )
  } else {
    title <- "Regional composites vs CRU"
    subtitle <- paste0(
      "Grey band = calibration (", cal_y0, "–", cal_y1,
      "); compare co-variation of blue and red lines"
    )
  }

  p <- ggplot(ts, aes(year, z, colour = series, linetype = series))

  if (focus == "full") {
    p <- p + geom_rect(
      data = data.frame(xmin = cal_y0, xmax = cal_y1 + 0.5),
      aes(xmin = xmin, xmax = xmax, ymin = -Inf, ymax = Inf),
      inherit.aes = FALSE, fill = "grey85", alpha = 0.35, colour = NA
    )
  }

  p +
    geom_hline(yintercept = 0, linewidth = line_lwd_thin, colour = "grey75") +
    geom_line(linewidth = line_lwd, alpha = 0.95) +
    facet_wrap(~panel, scales = "free_y", ncol = 2) +
    scale_colour_manual(
      values = c("CRU regional mean" = "#2166ac", "Speleothem composite" = "#b2182b"),
      name = NULL
    ) +
    scale_linetype_manual(
      values = c("CRU regional mean" = "solid", "Speleothem composite" = "22"),
      name = NULL
    ) +
    labs(
      title = title, subtitle = subtitle,
      caption = "Flat red line while blue varies → proxies capture little year-to-year variance.",
      x = "Year (CE)", y = "Z-score"
    ) +
    theme_clean(grid = "y") +
    theme(
      legend.position = "bottom",
      strip.text = element_text(size = 9, face = "bold"),
      panel.spacing.y = unit(0.9, "lines")
    )
}

#' Map sites used in regional reconstructions.
plot_recon_site_map <- function(corr, recon_summary) {
  used_ids <- unique(unlist(recon_summary$site_ids))
  map_df <- corr |>
    mutate(
      region = assign_region(lon, lat),
      in_recon = dataset_id %in% used_ids,
      hover = paste0(dataset_id, "<br>", region, "<br>",
                     if_else(in_recon, "Used in Step 2", "Not screened"))
    )

  plot_ly(map_df, type = "scattergeo", mode = "markers",
          lon = ~lon, lat = ~lat, text = ~hover, hoverinfo = "text",
          color = ~in_recon, colors = c("#bbbbbb", "#b2182b"),
          sizes = c(6, 10)) |>
    layout(
      title = plotly_title(
        "Sites used in regional composites",
        "Red = passed screening and entered a Step 2 composite; grey = not used"
      ),
      margin = list(t = 72, b = 0, l = 0, r = 0),
      geo = list(
        projection = list(type = "natural earth"),
        showland = TRUE, landcolor = "#f0f0f0",
        showocean = TRUE, oceancolor = "#e6f2fa",
        showcountries = TRUE, countrycolor = "#bbbbbb"
      )
    )
}
```

```{r}
#| label: recon-screen
#| cache: true
#| dependson: [corr-run, recon-setup]

if (!exists("cru_clim")) cru_clim <- readRDS(cru_rds)

speleo_corr <- speleo_corr |> screen_correlations()

screen_tbl <- speleo_corr |>
  filter(screened) |>
  count(region, best_var_label, name = "n") |>
  arrange(desc(n))

knitr::kable(screen_tbl, caption = "Screened series (p < 0.05 for best-fit climate variable)")

glue(
  "Screened: {sum(speleo_corr$screened, na.rm = TRUE)} of {nrow(speleo_corr)} series"
)
```

## Calibrate and validate

```{r}
#| label: recon-run
#| cache: true
#| dependson: [corr-run, recon-setup, recon-screen]

recon_summary <- build_regional_reconstructions(
  speleo_corr, catalog, cru_clim, speleo_sites
)

if (nrow(recon_summary)) {
  recon_summary |>
    dplyr::select(region, climate_label, n_series, r_cal, r_val, rmse_cal, rmse_val) |>
    mutate(across(where(is.numeric), \(x) round(x, 3))) |>
    knitr::kable(caption = "Regional composite skill (z-scores)")
} else {
  cat("No region × climate target with ≥", min_screen_n, "screened series.")
}
```

## Skill and time series

```{r}
#| label: recon-skill-plot
#| fig-cap: "Calibration vs validation Pearson r. Compare blue and red bar lengths — divergence flags overfitting."
#| fig-width: 9
#| fig-height: 6
#| dependson: recon-run

if (nrow(recon_summary)) plot_recon_skill(recon_summary)
```

```{r}
#| label: recon-ts-plot
#| fig-cap: "Independent validation period (1981–2000). Track whether red (composite) follows blue (CRU) year to year."
#| fig-width: 9
#| fig-height: 10
#| dependson: recon-run

if (nrow(recon_summary)) plot_recon_timeseries(recon_summary, focus = "validation")
```

```{r}
#| label: recon-map
#| fig-cap: "Geographic footprint of Step 2 composites. Sparse red points limit regional representativeness."
#| dependson: recon-run

if (nrow(recon_summary)) plot_recon_site_map(speleo_corr, recon_summary)
```

## Step 2 takeaways

```{r}
#| label: step2-takeaways-prep
#| include: false
#| dependson: recon-run

n_screened <- sum(speleo_corr$screened, na.rm = TRUE)
n_series_total <- nrow(speleo_corr)
step2_ok <- nrow(recon_summary) > 0L

if (step2_ok) {
  best_val <- recon_summary |> slice_max(r_val, n = 1, with_ties = FALSE)
  worst_val <- recon_summary |> slice_min(r_val, n = 1, with_ties = FALSE)
}
```

`r if (step2_ok) glue("Of {n_series_total} speleothem series, {n_screened} passed screening (p < {screen_p} for the best-fit climate variable). From these, **{nrow(recon_summary)} regional composite(s)** were built (≥ {min_screen_n} series per region × target).") else glue("Step 2 could not run — fewer than {min_screen_n} screened series per region × climate target. Screened: {n_screened} of {n_series_total} series; relax screening or extend the overlap window.")`

`r if (step2_ok) glue("**Best validation:** {best_val$region} — {best_val$climate_label} (*r*~val~ = {sprintf('%.3f', best_val$r_val)}, *n* = {best_val$n_series}). **Weakest:** {worst_val$region} — {worst_val$climate_label} (*r*~val~ = {sprintf('%.3f', worst_val$r_val)}).")`

Validation *r* below calibration *r* is common when caves in one region respond differently [@fairchild2006].

# Step 3 — ENSO-focused screening

**ENSO** (El Niño–Southern Oscillation) is a shift in tropical Pacific sea-surface temperature. We use the **Niño3.4 index** — warm anomalies in the central Pacific often change rainfall in eastern Australia [@tadros2016].

Our speleothem series are **one value per calendar year**. ENSO peaks in **boreal winter (DJF)**, so we also test a **year-mean Niño index** and **lags of ±1 year**, because rain at the cave can respond with a delay. As a second target we use **local summer rainfall (JJA)** from CRU — often a more direct link than the Pacific index itself.

**Regions:** SE Australia [@tadros2016], eastern Australasia, SW Pacific. **Proxies:** Mg/Ca and δ18O only.

```{r}
#| label: enso-setup
#| code-fold: true
#| cache: true
#| dependson: [corr-setup, recon-setup]

enso_y0 <- 1901L
enso_y1 <- 2000L
enso_min_overlap <- min_overlap   # same as Step 1 (10 yr)
enso_p_cut <- 0.05
enso_p_explore <- 0.20            # relaxed for exploratory composites
nino34_url <- "https://www.psl.noaa.gov/data/timeseries/month/data/nino34.long.anom.data"
nino34_dat <- file.path(cache_dir, "nino34.long.anom.data")
nino34_rds <- file.path(cache_dir, "nino34_index.rds")

enso_proxy_order <- c("Mg/Ca", "d18O")
enso_proxy_labels <- c(`Mg/Ca` = "Mg/Ca", d18O = "δ18O")

#' Parse NOAA PSL Niño3.4 monthly anomalies (HadISST, 1981–2010 base).
load_nino34_monthly <- function(refresh = FALSE) {
  if (!refresh && file.exists(nino34_rds)) return(readRDS(nino34_rds))

  if (!file.exists(nino34_dat)) {
    message("Downloading Niño3.4 index …")
    if (nzchar(Sys.which("curl"))) {
      status <- system2("curl", c("-L", "--retry", "3", "-o", nino34_dat, nino34_url))
      if (status != 0) stop("Niño3.4 download failed")
    } else {
      utils::download.file(nino34_url, nino34_dat, mode = "wb")
    }
  }

  lines <- readLines(nino34_dat, warn = FALSE)
  hl <- grep("^[[:space:]]*[0-9]{4}", lines)
  raw <- do.call(rbind, lapply(lines[hl], \(l) as.numeric(scan(text = l, quiet = TRUE))))
  colnames(raw) <- c("year", month.abb)
  monthly <- as_tibble(raw) |>
    pivot_longer(-year, names_to = "month", values_to = "nino34") |>
    mutate(month_num = match(month, month.abb)) |>
    filter(nino34 > -90)

  saveRDS(monthly, nino34_rds)
  monthly
}

#' Annual and DJF (Dec–Feb) Niño3.4 for climate year starting in Dec of prior year.
build_nino34_annual <- function(monthly) {
  djf <- monthly |>
    filter(month_num %in% c(12L, 1L, 2L)) |>
    mutate(
      climate_year = if_else(month_num == 12L, year + 1L, year)
    ) |>
    group_by(climate_year) |>
    summarize(nino34_djf = mean(nino34, na.rm = TRUE), .groups = "drop") |>
    rename(year = climate_year)

  ann <- monthly |>
    group_by(year) |>
    summarize(nino34_ann = mean(nino34, na.rm = TRUE), .groups = "drop")

  full_join(ann, djf, by = "year")
}

#' ENSO teleconnection regions [@tadros2016 and SW Pacific literature].
assign_enso_region <- function(lon, lat) {
  case_when(
    lon >= 138 & lon <= 155 & lat >= -40 & lat <= -28 ~ "SE Australia (Tadros)",
    lon >= 130 & lon <= 180 & lat >= -45 & lat <= -10 ~ "Eastern Australasia",
    lon >= 110 & lon <= 180 & lat >= -30 & lat <= 5 ~ "SW Pacific / IASM",
    TRUE ~ NA_character_
  )
}

#' Correlate detrended speleothem annual series with Niño3.4 and local rainfall.
compute_enso_correlations <- function(catalog, clim, sites, nino_ann,
                                      y0 = enso_y0, y1 = enso_y1,
                                      proxies = enso_proxy_order) {
  loc <- sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
  proxy_ann <- speleothem_annual(catalog) |>
    inner_join(loc, by = c("lon", "lat")) |>
    filter(proxy_type %in% proxies)

  map_dfr(group_split(proxy_ann, dataset_id), function(df) {
    reg <- assign_enso_region(df$lon[1], df$lat[1])
    if (is.na(reg)) return(NULL)

    cdf <- clim |> filter(loc_id == df$loc_id[1])
    d <- df |>
      filter(year >= y0, year <= y1) |>
      inner_join(cdf |> dplyr::select(year, pre_ann), by = "year") |>
      inner_join(nino_ann |> dplyr::select(year, nino34_djf, nino34_ann), by = "year")

    d <- d |>
      mutate(value_dt = detrend_linear(year, value))

    ok <- complete.cases(d$value_dt, d$nino34_djf, d$pre_ann)
    d <- d[ok, , drop = FALSE]
    if (nrow(d) < enso_min_overlap) return(NULL)

    ct_djf <- cor.test(d$value_dt, d$nino34_djf, method = "pearson")
    ct_ann <- cor.test(d$value_dt, d$nino34_ann, method = "pearson")
    ct_pre <- cor.test(d$value_dt, d$pre_ann, method = "pearson")

    tibble(
      dataset_id = df$dataset_id[1],
      proxy_type = df$proxy_type[1],
      lon = df$lon[1], lat = df$lat[1],
      enso_region = reg,
      n_overlap = nrow(d),
      r_nino_djf = unname(ct_djf$estimate),
      p_nino_djf = ct_djf$p.value,
      r_nino_ann = unname(ct_ann$estimate),
      p_nino_ann = ct_ann$p.value,
      r_pre_ann = unname(ct_pre$estimate),
      p_pre_ann = ct_pre$p.value
    )
  })
}

lag_join_nino <- function(df, nino_df, col, lag = 0L) {
  nj <- nino_df |>
    dplyr::select(year, nino_v = all_of(col)) |>
    mutate(year = year + lag)
  inner_join(df, nj, by = "year")
}

load_enso_seasonal_clim <- function(loc) {
  pre_nc <- file.path(cru_dir, "cru_ts4.07.1901.2022.pre.dat.nc")
  if (!file.exists(pre_nc)) return(NULL)
  pts <- vect(loc, geom = c("lon", "lat"), crs = "EPSG:4326")
  mat <- as.matrix(
    terra::extract(rast(pre_nc, subds = "pre"), pts, method = "simple")[, -1]
  )
  map_dfr(seq_len(nrow(loc)), function(i) {
    m <- mat[i, ]
    tibble(
      loc_id = loc$loc_id[i],
      year = cru_years,
      pre_djf = vapply(cru_years, function(yr) {
        dec <- (yr - 1901L) * 12L + 12L
        if (yr == 1901L) return(NA_real_)
        sum(m[c(dec, (yr - 1901L) * 12L + 1:2)], na.rm = TRUE)
      }, numeric(1)),
      pre_jja = vapply(cru_years, function(yr) {
        sum(m[(yr - 1901L) * 12L + 6:8], na.rm = TRUE)
      }, numeric(1))
    )
  })
}

enso_corr_test <- function(x, y) {
  ok <- complete.cases(x, y)
  if (sum(ok) < enso_min_overlap) return(list(r = NA_real_, p = NA_real_, n = sum(ok)))
  ct <- cor.test(x[ok], y[ok], method = "pearson")
  list(r = unname(ct$estimate), p = ct$p.value, n = sum(ok))
}

compute_enso_sweep <- function(catalog, clim, sites, nino_ann, clim_season = NULL,
                               lags = -1:1) {
  loc <- sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
  proxy_ann <- speleothem_annual(catalog) |>
    inner_join(loc, by = c("lon", "lat")) |>
    filter(proxy_type %in% enso_proxy_order)

  nino_cols <- c("nino34_djf", "nino34_ann")
  clim_cols <- c("pre_ann", "pre_djf", "pre_jja")

  map_dfr(group_split(proxy_ann, dataset_id), function(df) {
    reg <- assign_enso_region(df$lon[1], df$lat[1])
    if (is.na(reg)) return(NULL)

    cdf <- clim |> filter(loc_id == df$loc_id[1])
    if (!is.null(clim_season)) {
      cdf <- cdf |>
        left_join(clim_season |> dplyr::select(loc_id, year, pre_djf, pre_jja),
                  by = c("loc_id", "year"))
    }

    base <- df |>
      filter(year >= enso_y0, year <= enso_y1) |>
      inner_join(cdf, by = c("loc_id", "year")) |>
      mutate(value_dt = detrend_linear(year, value))

    rows <- list()
    for (lag in lags) {
      for (ncol in nino_cols) {
        d <- lag_join_nino(base, nino_ann, ncol, lag)
        ok <- complete.cases(d$value_dt, d$nino_v)
        if (sum(ok) < enso_min_overlap) next
        ct <- enso_corr_test(d$value_dt[ok], d$nino_v[ok])
        rows[[length(rows) + 1L]] <- tibble(
          dataset_id = df$dataset_id[1], proxy_type = df$proxy_type[1],
          enso_region = reg, n_overlap = sum(ok),
          target = ncol, lag = lag, r = ct$r, p = ct$p
        )
      }
      for (ccol in intersect(clim_cols, names(base))) {
        ok <- complete.cases(base$value_dt, base[[ccol]])
        if (sum(ok) < enso_min_overlap) next
        ct <- enso_corr_test(base$value_dt[ok], base[[ccol]][ok])
        rows[[length(rows) + 1L]] <- tibble(
          dataset_id = df$dataset_id[1], proxy_type = df$proxy_type[1],
          enso_region = reg, n_overlap = sum(ok),
          target = ccol, lag = 0L, r = ct$r, p = ct$p
        )
      }
    }
    if (!length(rows)) return(NULL)
    bind_rows(rows)
  })
}

build_enso_composite <- function(sweep, catalog, clim, sites, nino_ann,
                                 region, proxy_type = "d18O",
                                 method = c("baseline", "tuned"),
                                 min_n = 3L) {
  method <- match.arg(method)
  pool <- sweep |>
    filter(enso_region == region, proxy_type == !!proxy_type)

  if (method == "baseline") {
    pick <- pool |>
      filter(target == "nino34_djf", lag == 0L) |>
      arrange(desc(abs(r))) |>
      slice_head(n = max(min_n, 3L))
    nino_col <- "nino34_djf"
    lag <- 0L
  } else {
    pick <- pool |>
      filter(grepl("^nino", target)) |>
      group_by(dataset_id) |>
      slice_max(abs(r), n = 1, with_ties = FALSE) |>
      ungroup() |>
      arrange(desc(abs(r))) |>
      slice_head(n = max(min_n, 5L))
    if (!nrow(pick)) return(NULL)
    nino_col <- pick$target[1]
    lag <- pick$lag[1]
  }

  if (nrow(pick) < min_n) return(NULL)

  loc <- sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
  proxy_ann <- speleothem_annual(catalog) |> inner_join(loc, by = c("lon", "lat"))

  site_ts <- map(pick$dataset_id, function(id) {
    row <- pick |> filter(dataset_id == id)
    px <- proxy_ann |> filter(dataset_id == id)
    cdf <- clim |> filter(loc_id == px$loc_id[1])
    d <- px |>
      filter(year >= enso_y0, year <= enso_y1) |>
      mutate(proxy_dt = detrend_linear(year, value)) |>
      inner_join(cdf |> dplyr::select(year, pre_ann), by = "year")
    d <- lag_join_nino(d, nino_ann, row$target[1], row$lag[1])
    if (nrow(d) < enso_min_overlap) return(NULL)
    d |>
      mutate(
        proxy_z = zscore_ref(proxy_dt, year, enso_y0, 1980L),
        sign_r = sign(row$r[1]),
        proxy_z_signed = if_else(sign_r < 0, -proxy_z, proxy_z)
      ) |>
      dplyr::select(year, proxy_z_signed, nino_v, pre_ann)
  })
  site_ts <- compact(site_ts)
  if (length(site_ts) < min_n) return(NULL)

  composite <- bind_rows(site_ts, .id = "site_idx") |>
    group_by(year) |>
    summarize(
      proxy_z = mean(proxy_z_signed, na.rm = TRUE),
      nino_v = mean(nino_v, na.rm = TRUE),
      pre_ann = mean(pre_ann, na.rm = TRUE),
      n_series = n(),
      .groups = "drop"
    ) |>
    mutate(
      nino_z = zscore_ref(nino_v, year, enso_y0, 1980L),
      proxy_z = zscore_ref(proxy_z, year, enso_y0, 1980L),
      pre_z = zscore_ref(pre_ann, year, enso_y0, 1980L)
    )

  val_idx <- composite$year >= 1981L & composite$year <= enso_y1
  cal_idx <- composite$year >= enso_y0 & composite$year <= 1980L

  nino_label <- if (nino_col == "nino34_djf") "Niño3.4 DJF" else "Niño3.4 annual"

  list(
    region = region,
    proxy_type = proxy_type,
    method = method,
    nino_col = nino_col,
    nino_lag = lag,
    nino_label = nino_label,
    n_series = length(site_ts),
    site_ids = pick$dataset_id[seq_len(length(site_ts))],
    r_cal = safe_cor(composite$nino_z[cal_idx], composite$proxy_z[cal_idx]),
    r_val = safe_cor(composite$nino_z[val_idx], composite$proxy_z[val_idx]),
    r_pre_val = safe_cor(composite$pre_z[val_idx], composite$proxy_z[val_idx]),
    data = composite
  )
}

plot_enso_screen <- function(enso_corr) {
  long <- enso_corr |>
    filter(is.finite(r_nino_djf)) |>
    mutate(
      proxy_label = enso_proxy_labels[proxy_type],
      label = paste0(enso_region, " — ", proxy_label)
    ) |>
    group_by(label) |>
    summarize(
      best_r = r_nino_djf[which.max(abs(r_nino_djf))],
      .groups = "drop"
    ) |>
    mutate(label = fct_reorder(label, best_r))

  ggplot(long, aes(best_r, label, fill = best_r > 0)) +
    geom_col(width = 0.65) +
    geom_vline(xintercept = 0, linewidth = line_lwd_thin, colour = "grey40") +
    scale_fill_manual(values = c("TRUE" = "#4393c3", "FALSE" = "#d6604d"), guide = "none") +
    labs(
      title = "Best site-wise correlation with Niño3.4 (DJF)",
      subtitle = paste0("Detrended proxies, ", enso_y0, "–", enso_y1),
      x = "Pearson r (proxy vs Niño3.4 DJF)", y = NULL
    ) +
    theme_clean(grid = "x")
}

plot_enso_best_scatter <- function(enso_corr, catalog, sites, nino_ann) {
  best <- enso_corr |>
    filter(is.finite(r_nino_djf)) |>
    group_by(enso_region, proxy_type) |>
    slice_max(abs(r_nino_djf), n = 1, with_ties = FALSE) |>
    ungroup()
  if (!nrow(best)) return(invisible(NULL))

  loc <- sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
  proxy_ann <- speleothem_annual(catalog) |> inner_join(loc, by = c("lon", "lat"))

  pts <- map_dfr(seq_len(nrow(best)), function(i) {
    row <- best[i, ]
    proxy_ann |>
      filter(dataset_id == row$dataset_id) |>
      filter(year >= enso_y0, year <= enso_y1) |>
      mutate(value_dt = detrend_linear(year, value)) |>
      inner_join(nino_ann |> dplyr::select(year, nino34_djf), by = "year") |>
      mutate(
        facet = paste0(
          row$enso_region, " — ", enso_proxy_labels[row$proxy_type],
          "\n", row$dataset_id, " (r = ", sprintf("%.2f", row$r_nino_djf), ")"
        )
      )
  })

  ggplot(pts, aes(nino34_djf, value_dt)) +
    geom_point(alpha = 0.65, size = point_sz, colour = "#444444") +
    geom_smooth(method = "lm", se = TRUE, colour = "#b2182b", linewidth = line_lwd) +
    facet_wrap(~facet, scales = "free_y", ncol = 2) +
    labs(
      title = "Best ENSO teleconnection per region × proxy",
      subtitle = "Detrended proxy vs Niño3.4 DJF SST anomaly",
      x = "Niño3.4 DJF SST anomaly (°C)", y = "Detrended proxy"
    ) +
    theme_clean(grid = "y") +
    theme(strip.text = element_text(size = 8.5, face = "bold"))
}

plot_enso_method_comparison <- function(comp_summary) {
  comp_summary <- comp_summary |>
    mutate(
      panel = paste0(region, " — ", enso_proxy_labels[proxy_type]),
      method = factor(
        method,
        levels = c("Niño DJF (baseline)", "Niño tuned", "Local JJA rain (mean |r|)")
      )
    )

  ggplot(comp_summary, aes(r, method, fill = method)) +
    geom_col(width = 0.6) +
    geom_vline(xintercept = 0, linewidth = line_lwd_thin, colour = "grey40") +
    facet_wrap(~panel, ncol = 1, scales = "free_y") +
    scale_fill_manual(
      values = c(
        "Niño DJF (baseline)" = "#92c5de",
        "Niño tuned" = "#4393c3",
        "Local JJA rain (mean |r|)" = "#d6604d"
      ),
      guide = "none"
    ) +
    labs(
      title = "ENSO region: which target fits best?",
      subtitle = "Validation r (1981–2000) for composites; mean |r| for local summer rainfall",
      x = "Pearson r", y = NULL
    ) +
    theme_clean(grid = "x")
}

plot_enso_composite_ts <- function(comp) {
  if (is.null(comp)) return(invisible(NULL))
  d <- comp$data |> filter(year >= 1981L, year <= enso_y1)

  long <- d |>
    pivot_longer(c(nino_z, proxy_z), names_to = "series", values_to = "z") |>
    mutate(series = recode(series,
      nino_z = comp$nino_label,
      proxy_z = "Speleothem composite"
    ))

  ggplot(long, aes(year, z, colour = series)) +
    geom_hline(yintercept = 0, linewidth = line_lwd_thin, colour = "grey75") +
    geom_line(linewidth = line_lwd, alpha = 0.95) +
    scale_colour_manual(
      values = setNames(
        c("#2166ac", "#b2182b"),
        c(comp$nino_label, "Speleothem composite")
      ),
      name = NULL
    ) +
    labs(
      title = paste0(comp$region, " — ", enso_proxy_labels[comp$proxy_type], " composite"),
      subtitle = paste0(
        comp$method, " method; validation ", 1981, "–", enso_y1,
        "; ", comp$n_series, " series; r_val = ", sprintf("%.2f", comp$r_val)
      ),
      x = "Year (CE)", y = "Z-score (1901–1980 base)"
    ) +
    theme_clean(grid = "y") +
    theme(legend.position = "bottom")
}
```

## Data and screening

```{r}
#| label: enso-load
#| cache: true
#| dependson: [enso-setup, corr-run]

nino_monthly <- load_nino34_monthly()
nino_ann <- build_nino34_annual(nino_monthly)

if (!exists("cru_clim")) cru_clim <- readRDS(cru_rds)
cru_locs <- speleo_sites |> distinct(lon, lat) |> mutate(loc_id = row_number())
clim_season <- load_enso_seasonal_clim(cru_locs)

enso_corr <- compute_enso_correlations(catalog, cru_clim, speleo_sites, nino_ann)
enso_sweep <- compute_enso_sweep(catalog, cru_clim, speleo_sites, nino_ann, clim_season)

n_enso_sites <- nrow(enso_corr)
n_sig_djf <- sum(enso_corr$p_nino_djf < enso_p_cut, na.rm = TRUE)
n_sig_sweep <- enso_sweep |>
  filter(grepl("^nino", target)) |>
  group_by(dataset_id) |>
  slice_max(abs(r), n = 1, with_ties = FALSE) |>
  ungroup() |>
  filter(p < enso_p_cut) |>
  nrow()

glue(
  "{n_enso_sites} series in ENSO regions ({enso_y0}–{enso_y1}). ",
  "Baseline (Niño DJF, lag 0): {n_sig_djf} significant (p < {enso_p_cut}). ",
  "With lags and annual Niño: {n_sig_sweep} significant."
)
```

```{r}
#| label: enso-summary-table
#| dependson: enso-load

enso_corr |>
  mutate(
    proxy_label = enso_proxy_labels[proxy_type],
    sig_djf = p_nino_djf < enso_p_cut
  ) |>
  group_by(enso_region, proxy_label) |>
  summarize(
    n = n(),
    n_sig = sum(sig_djf, na.rm = TRUE),
    best_r_djf = r_nino_djf[which.max(abs(r_nino_djf))],
    best_p_djf = p_nino_djf[which.max(abs(r_nino_djf))],
    best_id = dataset_id[which.max(abs(r_nino_djf))],
    .groups = "drop"
  ) |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "ENSO screening by region and proxy (best |r| vs Niño3.4 DJF)")
```

## Screen vs Niño3.4

```{r}
#| label: enso-screen-plot
#| fig-cap: "Strongest single-series Niño3.4 DJF teleconnection per region and proxy type."
#| fig-width: 9
#| fig-height: 4
#| dependson: enso-load

if (nrow(enso_corr)) plot_enso_screen(enso_corr)
```

```{r}
#| label: enso-scatter-plot
#| fig-cap: "Showcase scatter plots for the best |r| series in each region × proxy panel."
#| fig-width: 9
#| fig-height: 8
#| dependson: enso-load

if (nrow(enso_corr)) {
  plot_enso_best_scatter(enso_corr, catalog, speleo_sites, nino_ann)
}
```

## Composites and comparison

For each region we build δ18O composites two ways: **baseline** (Niño DJF, lag 0, top 3 series by |r|) and **tuned** (best Niño index + lag per series, up to 5 series). We compare validation *r* (1981–2000) against the mean |r| to **local JJA rainfall**.

```{r}
#| label: enso-composite-run
#| cache: true
#| dependson: enso-load

enso_regions <- c("SE Australia (Tadros)", "SW Pacific / IASM")

enso_comp_base <- map(enso_regions, \(reg) {
  build_enso_composite(enso_sweep, catalog, cru_clim, speleo_sites, nino_ann,
                       region = reg, proxy_type = "d18O", method = "baseline")
}) |> set_names(enso_regions)

enso_comp_tuned <- map(enso_regions, \(reg) {
  build_enso_composite(enso_sweep, catalog, cru_clim, speleo_sites, nino_ann,
                       region = reg, proxy_type = "d18O", method = "tuned")
}) |> set_names(enso_regions)

enso_best_comp <- enso_comp_tuned[["SE Australia (Tadros)"]]
if (is.null(enso_best_comp)) enso_best_comp <- enso_comp_tuned[["SW Pacific / IASM"]]

mean_rain_r <- function(region) {
  enso_sweep |>
    filter(enso_region == region, target == "pre_jja") |>
    summarize(r = mean(abs(r), na.rm = TRUE), .groups = "drop") |>
    pull(r)
}

enso_method_summary <- bind_rows(
  map_dfr(enso_regions, function(reg) {
    b <- enso_comp_base[[reg]]
    t <- enso_comp_tuned[[reg]]
    bind_rows(
      if (!is.null(b)) tibble(region = reg, proxy_type = "d18O",
        method = "Niño DJF (baseline)", r = b$r_val, n_series = b$n_series),
      if (!is.null(t)) tibble(region = reg, proxy_type = "d18O",
        method = "Niño tuned", r = t$r_val, n_series = t$n_series),
      tibble(region = reg, proxy_type = "d18O",
        method = "Local JJA rain (mean |r|)", r = mean_rain_r(reg), n_series = NA_integer_)
    )
  })
)

enso_method_summary |>
  mutate(across(where(is.numeric), \(x) round(x, 3))) |>
  knitr::kable(caption = "Step 3 — validation-period skill by method (δ18O composites)")
```

```{r}
#| label: enso-method-plot
#| fig-cap: "Comparison of ENSO screening approaches in the two main regions. Tuned Niño settings improve validation r; local summer rain remains a strong alternative target."
#| fig-width: 8
#| fig-height: 6
#| dependson: enso-composite-run

if (nrow(enso_method_summary)) plot_enso_method_comparison(enso_method_summary)
```

```{r}
#| label: enso-composite-plot
#| fig-cap: "Tuned δ18O composites vs Niño index — validation period 1981–2000 (z-scores)."
#| fig-width: 9
#| fig-height: 7
#| dependson: enso-composite-run

if (!is.null(enso_comp_tuned[["SE Australia (Tadros)"]])) {
  plot_enso_composite_ts(enso_comp_tuned[["SE Australia (Tadros)"]])
}
if (!is.null(enso_comp_tuned[["SW Pacific / IASM"]])) {
  plot_enso_composite_ts(enso_comp_tuned[["SW Pacific / IASM"]])
}
```

## Step 3 takeaways

```{r}
#| label: step3-takeaways-prep
#| include: false
#| dependson: [enso-load, enso-composite-run]

step3_ok <- nrow(enso_corr) > 0L
if (step3_ok) {
  best_djf <- enso_corr |> slice_max(abs(r_nino_djf), n = 1, with_ties = FALSE)
  best_sweep <- enso_sweep |>
    filter(grepl("^nino", target)) |>
    slice_max(abs(r), n = 1, with_ties = FALSE)
  se_mg <- sum(enso_corr$enso_region == "SE Australia (Tadros)" &
                 enso_corr$proxy_type == "Mg/Ca")
}
step3_tuned_ok <- !is.null(enso_best_comp)
```

`r if (step3_ok) glue("In ENSO regions we tested **{n_enso_sites}** series. A simple Niño-DJF screen gives **{n_sig_djf}** significant results; allowing **lags and an annual Niño index** raises that to **{n_sig_sweep}**.") else "Step 3 could not run."`

`r if (step3_ok) glue("Best baseline teleconnection: **{best_djf$dataset_id}** (*r* = {sprintf('%.2f', best_djf$r_nino_djf)} vs Niño DJF). Best after retuning: **{best_sweep$dataset_id}** (*r* = {sprintf('%.2f', best_sweep$r)} for {best_sweep$target}, lag {best_sweep$lag}). Only **{se_mg}** Mg/Ca series exist in SE Australia.")`

`r if (step3_tuned_ok) glue("Best tuned δ18O composite (**{enso_best_comp$region}**, {enso_best_comp$n_series} series): validation *r* = **{sprintf('%.2f', enso_best_comp$r_val)}** vs {enso_best_comp$nino_label} (lag {enso_best_comp$nino_lag}).")`

DoD2k is useful for **hypothesis screening**, not for a publishable ENSO reconstruction: few Mg/Ca records, annual resolution, and karst mixing damp the signal [@fairchild2006]. A convincing palaeo-ENSO study would need **high-resolution, site-calibrated stalagmites** and often **multi-proxy networks** [@tadros2016].

# Conclusion

| Step | Main result |
|------|-------------|
| **1 — Screen** | Most speleothems correlate weakly with local CRU climate; the archive is heterogeneous. |
| **2 — Composite** | Some regional composites show moderate validation skill; others fail — site mixing matters. |
| **3 — ENSO** | Retuned Niño indices and lags improve correlations modestly; **local rainfall** is often the cleaner target. |

# References

::: {#refs}
:::
